### 2021-06-16 paper/main.tex
whoops recreating dropped files

\documentclass{article}
\usepackage[utf8]{inputenc}

% math macros
\usepackage{amsmath, amsfonts}
\DeclareMathOperator*{\argmax}{argmax}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\uvec}{\vec{\hat{e}}}
\newcommand{\tensor}[1]{\mathbf{#1}}
\newcommand{\air}{\text{air}}
\newcommand{\water}{\text{water}}
\newcommand{\boat}{\text{boat}}
\newcommand{\destination}{\text{dest}}
\newcommand{\sail}{\text{s}}
\newcommand{\keel}{\text{k}}
\newcommand{\vair}{\vec{v}_\air}
\newcommand{\vwater}{\vec{v}_\water}
\newcommand{\vboat}{\vec{v}_\boat}

% text macros
\usepackage{biblatex}
\addbibresource{sailing.bib}
\newcommand{\documentname}{\textsl{Article}}

% typesetting adjustments
\addtolength{\textheight}{2.20in}
\addtolength{\topmargin}{-1.00in}
\linespread{1.08}
\frenchspacing\sloppy\sloppypar\raggedbottom

\title{\bfseries%
An anisotropic ram-pressure model for sailing}
\author{DWH, MK, others?}
\date{June 2021}
\begin{document}
\maketitle

\begin{abstract}\noindent
    We present a simplified model for the forces on a sailboat.
    Our model is that there is a ram-pressure force on the sail, with a magnitude and direction that depend on the alignment of the relative air--boat velocity with the sail orientation according to an anisotropic drag tensor.
    Similarly, there is a ram-pressure force on the keel, depending on the relative water--boat velocity.
    The resulting expression for the total force on the boat is coordinate-free and consistent with Galilean relativity.
    The solutions corresponding to sailing in steady wind correspond to boat velocities that deliver vanishing total force, given the angles of the sail and keel.
    We use this model to derive simple rules for sailing that are not far off from standard sailing practice, for sailing in different directions (including upwind).
    One consequence of this model is that a sailboat does not move (relative to the water) precisely in the direction that its keel points (the direction of the prow); it moves slightly downwind of that.
    Another consequence is that it is possible for an extremely anisotropic (extremely well designed) boat to sail downwind (relative to the water) \emph{faster than the wind}, on a tacking course.
    Interestingly, nothing about the mechanical model or discussion depend on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing.
\end{abstract}

\section{Introduction}

KLEBAN: Write things about ddwfttw, maybe?

HOGG: Stuff about Bernoulli and flying and aerofoils that turned out to be wrong; a motivation to reconsider sailing too.

\section{Ram-pressure sailboat}

We are going to work in 2D, where velocities $\vec{v}$ are column vectors (or $\vec{v}\in\mathbb{R}^{2\times 1}$).
To stay coordinate-free, we are going to talk about angles as orientations of unit vectors like $\uvec_{\perp\sail}$ and $\uvec_{\parallel\sail}$ (which will be defined below).
If you prefer to think in terms of angles, you can choose a coordinate system and then
\begin{align}
    \uvec_{\perp,\sail} &= \begin{bmatrix}\cos\theta_\sail \\ \sin\theta_\sail\end{bmatrix} ~ ; ~ \uvec_{\parallel,\sail} = \begin{bmatrix}-\sin\theta_\sail \\ \cos\theta_\sail\end{bmatrix} ~,
\end{align}
for some angle $\theta_\sail$ in that coordinate system.
We won't particularly use angles much in what follows, but if you want to come back to angles, you can take the arctangent of the components of $\uvec_{\perp\sail}$.

Our sailboat is going to be an anisotropic object such that wind that is perpendicular to the sail will produce more ram pressure than wind at any other angle (and far more than wind that ``goes along'' the sail direction).
In detail, the ram-pressure force from the air on the sailboat will be
\begin{align}
    \vec{F}_\air &= \rho_\air\,|\vair - \vboat|\,\tensor{A}_\air\,(\vair - \vboat) \label{eq:Fair}
    \\
    \tensor{A}_\air &= A_{\perp\sail}\,\uvec_{\perp\sail}\,\uvec_{\perp\sail}^\top + A_{\parallel\sail},\uvec_{\parallel\sail}\,\uvec_{\parallel\sail}^\top \label{eq:Aair}
    \\
    A_{\perp\sail} &\gg A_{\parallel\sail} \label{eq:sail}
    ~,
\end{align}
where $\rho_\air$ is the mass density of air, $\vair,\vboat$ are the velocities of the air and boat, $\tensor{A}_\air$ is a Hermitian tensor with units of area, $A_{\perp\sail}, A_{\parallel\sail}$ are two constants with units of area defining the geometry of the sail, $\uvec_{\perp\sail}$ is a unit vector pointing normal to the surface of the (assumed planar) sail, and $\uvec_{\parallel\sail}$ is a unit vector perpendicular to that.
If you are uncomfortable with the expression for the tensor $\tensor{A}_\air$ in \eqref{eq:Aair}, look at the discussion of tensors in, for example, \cite{kusse}.
The inequality \eqref{eq:sail} says (loosely) that the sail (plus hull) is much bigger face-on than edge-on.
The force from the water is similar
\begin{align}
    \vec{F}_\water &= \rho_\water\,|\vwater - \vboat|\,\tensor{A}_\water\,(\vwater - \vboat) \label{eq:Fwater}
    \\
    \tensor{A}_\water &= A_{\perp\keel}\,\uvec_{\perp\keel}\,\uvec_{\perp\keel}^\top + A_{\parallel\keel},\uvec_{\parallel\keel}\,\uvec_{\parallel\keel}^\top
    \\
    A_{\perp\keel} &\gg A_{\parallel\keel} ~,
\end{align}
where we have gone from air to water and sail ``s'' to keel ``k''.
The ram-pressure forces given in \eqref{eq:Fair} and \eqref{eq:Fwater} can be justified with a simple molecular model in which air and water molecules are elastically bouncing off the sail and keel surfaces; this is the same set of assumptions that leads to, for example, the ideal gas law (HOGG CITE).

Mathematically, steady sailing in steady wind and steady current (constant $\vair, \vwater$) is given by
\begin{align}\label{eq:sailing}
    \vec{F} &= \vec{F}_\air + \vec{F}_\water = \vec{0} ~.
\end{align}
If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel. KLEBAN CHECK THIS: This is a quartic root-finding problem so it has no simple closed form; we solve it numerically in the code associated with this \documentname.

HOGG: PUT FIGURES HERE SHOWING THE BEHAVIOR OF THE BOAT IN THE WIND.

\section{Good sailing}

In general a type-A sailor is trying to get from a current position to a destination as quickly as possible.
Our view is that this sailor should start by specifying the unit vector $\uvec_\destination$ that points from the current position of the boat towards the destination of the boat.
The sailor should then maximize the scalar product $\uvec_\destination^\top\,\vboat$
\begin{align}
    \uvec_{\perp,\sail}^\ast,\uvec_{\perp\keel}^\ast &\leftarrow \argmax_{\uvec_{\perp,\sail},\uvec_{\perp\keel}} (\uvec_\destination^\top\,\vboat) ~,
\end{align}
where $\uvec_{\perp,\sail}^\ast,\uvec_{\perp\keel}^\ast$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and implicitly $\vboat$ is being thought of as a function of these orientations.

While this is simple to state, the numerical optimization involves a quantity $\vboat$ that is itself the root of a quartic expression.
Numerically this optimization can be implemented using a Lagrange multiplier and it looks like this...HOGG

HOGG PUT FIGURES HERE SHOWING THE OPTIMAL SETTINGS OF SAIL AND KEEL ORIENTATIONS FOR DIFFERENT DESTINATION DIRECTIONS.

Now if you want to plan a complete path or journey for the sailboat in (unrealistically) steady wind, you... HOGG

HOGG PUT FIGURES HERE SHOWING TRAJECTORIES FOR SIMPLE JOURNEYS.

\section{Discussion}

HOGG: What did we find? What's important about that?

HOGG: Galilean symmetry, coordinate freedom?

HOGG: What were our assumptions? What are the myriad ways in which those assumptions must be wrong? For example, boat has an asymmetry in the wind too, so really the boat must also matter. Viscosity; why can we ignore it? But turbulence; we can't ignore that!

HOGG: Take some time to bash the Bernoulli bullshit.

HOGG: All code used to make the figures in this \documentname{} are available HOGG WHERE?

\paragraph{Acknowledgements:}
It is a pleasure to thank Hans-Walter Rix (MPIA) for valuable discussions.

\raggedright
\printbibliography
\end{document}

### 2021-06-17 paper/main.tex
Updates from Overleaf

@@
+% This paper is copyright 2021 the authors.
+
+% people who might comment on it:
+% - Rix
+% - KSF
+% - Kilian Walsh
+
 \documentclass{article}
@@
 \DeclareMathOperator*{\argmax}{argmax}
+\newcommand{\dd}{\mathrm{d}}
 \renewcommand{\vec}[1]{\boldsymbol{#1}}
@@
 \newcommand{\air}{\text{air}}
-\newcommand{\water}{\text{water}}
+\newcommand{\water}{\text{wat}}
 \newcommand{\boat}{\text{boat}}
@@
 \newcommand{\documentname}{\textsl{Article}}
+\newcommand{\secref}[1]{Section~\ref{#1}}
 
 % typesetting adjustments
+\makeatletter
+\renewcommand\section{\@startsection {section}{1}{\z@}%
+  {-3.25ex \@plus -1ex \@minus -.2ex}%
+  {1.5ex \@plus .2ex}%
+  {\raggedright\normalfont\large\bfseries}}
+\makeatother
 \addtolength{\textheight}{2.20in}
@@
 \title{\bfseries%
-An anisotropic ram-pressure model for sailing}
+A simple, anisotropic ram-pressure model for sailing}
 \author{DWH, MK, others?}
@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
@@
 
-\section{Ram-pressure sailboat}
+\section{Ram-pressure sailboat}\label{sec:boat}
 
@@
 \begin{align}
-    \vec{F}_\air &= \rho_\air\,|\vair - \vboat|\,\tensor{A}_\air\,(\vair - \vboat) \label{eq:Fair}
+    \Delta\vair &\equiv \vair-\vboat
+    \\
+    \vec{F}_\air &= \rho_\air\,|\Delta\vair|\,\tensor{A}_\air\,\Delta\vair \label{eq:Fair}
     \\
@@
 \begin{align}
-    \vec{F}_\water &= \rho_\water\,|\vwater - \vboat|\,\tensor{A}_\water\,(\vwater - \vboat) \label{eq:Fwater}
+    \Delta\vwater &\equiv \vwater-\vboat
+    \\
+    \vec{F}_\water &= \rho_\water\,|\Delta\vwater|\,\tensor{A}_\water\,\Delta\vwater \label{eq:Fwater}
     \\
@@
 If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel.
-KLEBAN CHECK THIS: This is a quartic root-finding problem so it has no simple closed form; we solve it numerically in the code associated with this \documentname.
+KLEBAN CHECK THIS: This is a quartic root-finding problem so it has no simple closed form; we solve it numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method.
+For Newton's method we make use of the derivatives
+\begin{align}
+    \frac{\dd\vec{F}_\air}{\dd\vboat} &= -\frac{\rho_\air}{|\Delta\vair|}\,A_\air\,\Delta\vair\,\Delta\vair^\top-\rho_\air\,|\Delta\vair|\,A_\air
+    \\
+    \frac{\dd\vec{F}_\water}{\dd\vboat} &= -\frac{\rho_\water}{|\Delta\vwater|}\,A_\water\,\Delta\vwater\,\Delta\vwater^\top-\rho_\water\,|\Delta\vwater|\,A_\water
+    ~,
+\end{align}
+which are order-2 tensors.
 
@@
 
-\section{Good sailing}
+HOGG: Comments: Don't sail DIRECTLY in the direction the keel is pointed. Many of these settings sail backwards: We don't distinguish forwards and backwards! Our boat is quadrupolar!
 
-In general a type-A sailor is trying to get from a current position to a destination as quickly as possible.
+\section{Good sailing}\label{sec:sailing}
+
+In \secref{sec:boat} we showed how the boat moves, given any arbitrary setting of the orientation of the keel and the sail.
+However, one doesn't sail by randomly setting the keel and sail!
+One sails by pointing the boat some direction and setting the sail (or sails) appropriately.
+In our simple, ram-pressure boat, how should we set the planar sail?
+
+HOGG:...
+
+\section{Competitive sailing}\label{sec:racing}
+
+A type-A sailor---or maybe a competetive sailor---is trying to get from a current position to a destination as quickly as possible.\footnote{We don't particularly endorse this attitude towards sailing.
+The journey \emph{is} the destination.}
 Our view is that this sailor should start by specifying the unit vector $\uvec_\destination$ that points from the current position of the boat towards the destination of the boat.
@@
 where $\uvec_{\perp,\sail}^\ast,\uvec_{\perp\keel}^\ast$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and implicitly $\vboat$ is being thought of as a function of these orientations.
+Note that the sailor explicitly \emph{does not want} to travel directly towards the destination:
+The sailor wants to make as much progress as possible in that direction, but is willing to tack back-and-forth to get there, if it gets the boat there faster.
+So the metric really is the projection of the boat velocity onto the displacement vector pointing from the current position of the boat to the destination.
 
-While this is simple to state, the numerical optimization involves a quantity $\vboat$ that is itself the root of a quartic expression.
-Numerically this optimization can be implemented using a Lagrange multiplier and it looks like this...HOGG
+While this is simple to state, the numerical optimization hurts...HOGG
 
@@
 
-\section{Discussion}
+\section{Numerical implementation notes}\label{sec:implementation}
+
+Newton's method, invert, solve, etc.
+
+Look-up tables for proper boat trim.
+
+Path planning.
+
+\section{Discussion}\label{sec:discussion}
 

### 2021-06-18 paper/main.tex
notes

@@
 % people who might comment on it:
-% - Rix
+% -------------------------------
 % - KSF
 % - Kilian Walsh
+
+% to-do items:
+% ------------
+% — Put in the dumbell figure
+% - Reconsider the post-equation commas? (HWR SAY).
+% — Find references for, and destroy, Bernoulli.
+% — Make some comments on the situations in which the ram-pressure force approximation is more appropriate---gale force or lull?
 
@@
 Our sailboat is going to be an anisotropic object such that wind that is perpendicular to the sail will produce more ram pressure than wind at any other angle (and far more than wind that ``goes along'' the sail direction).
-In detail, the ram-pressure force from the air on the sailboat will be
+In detail, the ram-pressure force $\vec{F}_\air$ from the air on the sailboat will be
 \begin{align}
@@
 The inequality \eqref{eq:sail} says (loosely) that the sail (plus hull) is much bigger face-on than edge-on.
-The force from the water is similar
+The force $\vec{F}_\water$ from the water is similar
 \begin{align}
@@
 If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel.
-KLEBAN CHECK THIS: This is a quartic root-finding problem so it has no simple closed form; we solve it numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method.
-For Newton's method we make use of the derivatives
-\begin{align}
-    \frac{\dd\vec{F}_\air}{\dd\vboat} &= -\frac{\rho_\air}{|\Delta\vair|}\,A_\air\,\Delta\vair\,\Delta\vair^\top-\rho_\air\,|\Delta\vair|\,A_\air
-    \\
-    \frac{\dd\vec{F}_\water}{\dd\vboat} &= -\frac{\rho_\water}{|\Delta\vwater|}\,A_\water\,\Delta\vwater\,\Delta\vwater^\top-\rho_\water\,|\Delta\vwater|\,A_\water
-    ~,
-\end{align}
-which are order-2 tensors.
+KLEBAN CHECK THIS: This is a quartic root-finding problem so it has no simple closed form; we solve it numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
+
+HOGG: Give the settings of all our parameters for the figures we are about to show.
+
+HOGG: Define direction cosines with the wind using vectors.
 
@@
 
-HOGG: Comments: Don't sail DIRECTLY in the direction the keel is pointed. Many of these settings sail backwards: We don't distinguish forwards and backwards! Our boat is quadrupolar!
+HOGG: Comments: You can see from the figures that you don't sail DIRECTLY in the direction the keel is pointed.
+
+HOGG: Right now, the concept of forwards and backwards is a bit soft; our boat is quadrupolar! But we will get more serious about deliberately sailing in the correct direction in the next Section.
 
@@
 However, one doesn't sail by randomly setting the keel and sail!
-One sails by pointing the boat some direction and setting the sail (or sails) appropriately.
+One sails by pointing the boat in some direction and then setting the sail (or sails) appropriately.
 In our simple, ram-pressure boat, how should we set the planar sail?
@@
 where $\uvec_{\perp,\sail}^\ast,\uvec_{\perp\keel}^\ast$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and implicitly $\vboat$ is being thought of as a function of these orientations.
-Note that the sailor explicitly \emph{does not want} to travel directly towards the destination:
+Note that the sailor \emph{does not necessarily want} to travel directly towards the destination:
 The sailor wants to make as much progress as possible in that direction, but is willing to tack back-and-forth to get there, if it gets the boat there faster.
@@
 
-Newton's method, invert, solve, etc.
+HOGG: Newton's method (CITE), invert, solve, etc.
+Note that in practice, Newton's method works amazingly well.
+For Newton's method we make use of the derivatives
+\begin{align}
+    \frac{\dd\vec{F}_\air}{\dd\vboat} &= -\frac{\rho_\air}{|\Delta\vair|}\,A_\air\,\Delta\vair\,\Delta\vair^\top-\rho_\air\,|\Delta\vair|\,A_\air
+    \\
+    \frac{\dd\vec{F}_\water}{\dd\vboat} &= -\frac{\rho_\water}{|\Delta\vwater|}\,A_\water\,\Delta\vwater\,\Delta\vwater^\top-\rho_\water\,|\Delta\vwater|\,A_\water
+    ~,
+\end{align}
+which are order-2 tensors.
 

### 2021-06-19 paper/main.tex
outline of comments, WRONG FORCE LAW

@@
 % ------------
+% - FIX THE FORCE EXPRESSION and discuss it in terms of inner products, tensors, and Einstein summation notation.
+% - Put in running head.
+% - Overleaf seems to have wrong papersize?
 % — Put in the dumbell figure
@@
 \makeatother
-\addtolength{\textheight}{2.20in}
-\addtolength{\topmargin}{-1.00in}
+\addtolength{\textheight}{2.00in}
+\addtolength{\topmargin}{-1.20in}
 \linespread{1.08}
@@
 In detail, the ram-pressure force $\vec{F}_\air$ from the air on the sailboat will be
+HOGG THIS EXPRESSION IS ALL WRONG. THE FORCE IS GIVEN BY DOUBLE-CONTRACTION ON AN ORDER-2 TENSOR. SEE NOTES 2021-06-19. FIX THIS.
 \begin{align}
@@
 
-HOGG: Define direction cosines with the wind using vectors.
+HOGG: Define direction cosines with the wind, and with boat parts, using vectors.
 
 HOGG: PUT FIGURES HERE SHOWING THE BEHAVIOR OF THE BOAT IN THE WIND.
+
+HOGG: Put the random-dot hourglass figure here.
 
@@
 HOGG: Right now, the concept of forwards and backwards is a bit soft; our boat is quadrupolar! But we will get more serious about deliberately sailing in the correct direction in the next Section.
+
+HOGG: Comment: It is easy to sail downwind fttw.
+
+HOGG: Comment: It is easy to sail upwind. But impossible to sail upwind fttw.
+
+HOGG: Comment: You can't sail directly into the wind, or within XXX radians of directly into the wind. This XXX depends on dimensionless sail and boat ratios.
 
@@
 
-HOGG:...
+HOGG: IN THIS SECTION EXPLICITLY DEFINE ``GOOD SAILING''.
+
+HOGG: FIGURE showing good sailing for a set of boat orientations.
+
+HOGG: Figure showing direction cosines of sail relative to keel/boat orientation.
+
+HOGG: Put the equivalent of the hourglass figure.
+
+HOGG: Comment: The settings of the sail look ``tight'' relative to standard sailing practice.
+But this boat has flat sails, so it's hard to compare the exact settings with sailor experiences (which are all with curved sails).
 
 \section{Competitive sailing}\label{sec:racing}
+
+HOGG IN THIS SECTION EXPLICITLY DEFINE ``COMPETITIVE SAILING''.
 

### 2021-06-19 paper/main.tex
note to self
start fixing force model

@@
 \newcommand{\destination}{\text{dest}}
+\newcommand{\good}{\text{(good)}}
+\newcommand{\best}{\text{(best)}}
 \newcommand{\sail}{\text{s}}
@@
 \newcommand{\vboat}{\vec{v}_\boat}
+\newcommand{\vdest}{\vec{v}_\destination}
 
@@
 \makeatother
-\addtolength{\textheight}{2.00in}
-\addtolength{\topmargin}{-1.20in}
+\setlength{\textwidth}{5.00in}
+\setlength{\textheight}{9.00in}
+\setlength{\topmargin}{-0.40in}
+\pagestyle{myheadings}
+\markright{\textsf{Hogg \& Kleban / Ram-pressure model for sailing}}
 \linespread{1.08}
@@
 \begin{align}
-    \uvec_{\perp,\sail} &= \begin{bmatrix}\cos\theta_\sail \\ \sin\theta_\sail\end{bmatrix} ~ ; ~ \uvec_{\parallel,\sail} = \begin{bmatrix}-\sin\theta_\sail \\ \cos\theta_\sail\end{bmatrix} ~,
+    \uvec_{\perp\sail} &= \begin{bmatrix}\cos\theta_\sail \\ \sin\theta_\sail\end{bmatrix} ~ ; ~ \uvec_{\parallel\sail} = \begin{bmatrix}-\sin\theta_\sail \\ \cos\theta_\sail\end{bmatrix} ~,
 \end{align}
@@
 
-Our sailboat is going to be an anisotropic object such that wind that is perpendicular to the sail will produce more ram pressure than wind at any other angle (and far more than wind that ``goes along'' the sail direction).
-In detail, the ram-pressure force $\vec{F}_\air$ from the air on the sailboat will be
+Our sailboat is going to be an anisotropic object such that wind that hits the sail perpendicular to the large surface of the sail (in the rest frame of the boat) will produce more ram pressure than wind at any other angle (and far more than wind that ``goes along'' the sail direction).
+We can construct the complete force law on the sailboat by thinking about ram-pressure forces on surfaces.
+If the fluid (air or water in this case) is thought of as being composed of particles that bounce elastically off of a flat surface of area $A$, then the ram-pressure force on that surface is
+\begin{align}
+    \vec{F} &= \eta\,\rho\,A\,(\uvec_\perp^\top\vec{v})^2\,\uvec_\perp ~,
+\end{align}
+where $\eta$ is a dimensionless constant (naively 2, but HOGG), $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the velocity of the fluid in the rest frame of the surface, and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
+If you like thinking about tensors, this can be thought of as a tensor expression, in which $\rho$, $A$, and the unit vectors are combined into a three-index (order-three) tensor $\mathbb{A}$, which is then contracted onto the velocity twice:
+\begin{align}
+    \mathbb{A} &\equiv \eta\,A\,\uvec_\perp\,\uvec_\perp^\top\otimes\uvec_\perp^\top ~.
+\end{align}
+In Einstein summation notation, the three-index tensor is contracted as follows
+\begin{align}
+    [\vec{F}]_i &= \rho\,[\mathbb{A}]_{ijk}\,[\vec{v}]_j\,[\vec{v}]_k ~,
+    \\
+    [\mathbb{A}]_{ijk} &= \eta\,A\,[\uvec_\perp]_i\,[\uvec_\perp]_j\,[\uvec_\perp]_k
+    ~,
+\end{align}
+where the $[\vec{F}]_i$ notation indicates the $i$th component or coordinate of the vector, the tensor $\mathbb{A}$ has three indices, and the repeated indices are summed over.
+
+HOGG: PARAGRAPH AND EQUATIONS ALL WRONG: In detail, the ram-pressure force $\vec{F}_\air$ from the air on the sailboat will be
 HOGG THIS EXPRESSION IS ALL WRONG. THE FORCE IS GIVEN BY DOUBLE-CONTRACTION ON AN ORDER-2 TENSOR. SEE NOTES 2021-06-19. FIX THIS.
@@
 If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel.
-KLEBAN CHECK THIS: This is a quartic root-finding problem so it has no simple closed form; we solve it numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
+This is a quadratic root-finding problem so it has HOGG SOME SIMPLE CLOSED FORM, WHICH IS...
+we CAN ALSO MAYBE solve it numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
 
@@
 
+HOGG: Comment: It is easy to sail faster than the wind speed.
+
 HOGG: Comment: It is easy to sail upwind. But impossible to sail upwind fttw.
@@
 HOGG: IN THIS SECTION EXPLICITLY DEFINE ``GOOD SAILING''.
+\begin{align}\label{eq:good}
+    \uvec_{\perp\sail}^\good &\leftarrow \argmax_{\uvec_{\perp\sail}} \left[\uvec_{\parallel\keel}^\top\,(\vboat-\vwater)\right] ~,
+\end{align}
+where .... HOGG IMPLICITLY $\vboat-\vwater$ is a function of all the properties of the world and the boat boat, along with the orientation of its keel and sail.
 
@@
 
+HOGG: Final comment: But this is not the BEST a sailor can do...
+
 \section{Competitive sailing}\label{sec:racing}
@@
 The sailor should then maximize the scalar product $\uvec_\destination^\top\,\vboat$
-\begin{align}
-    \uvec_{\perp,\sail}^\ast,\uvec_{\perp\keel}^\ast &\leftarrow \argmax_{\uvec_{\perp,\sail},\uvec_{\perp\keel}} (\uvec_\destination^\top\,\vboat) ~,
+\begin{align}\label{eq:best}
+    \uvec_{\perp\sail}^\best,\uvec_{\perp\keel}^\best &\leftarrow \argmax_{\uvec_{\perp\sail},\uvec_{\perp\keel}} \left[\uvec_\destination^\top\,(\vboat-\vdest)\right] ~,
 \end{align}
-where $\uvec_{\perp,\sail}^\ast,\uvec_{\perp\keel}^\ast$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and implicitly $\vboat$ is being thought of as a function of these orientations.
+where $\uvec_{\perp\sail}^\best,\uvec_{\perp\keel}^\best$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and (because we are Galilean-relativistic) this velocity is relative to the velocity $\vdest$ of the destination (which might not be stationary with respect to the water).
+In this expresion \eqref{eq:best}, implicitly $\vboat$ is being thought of as a function of these orientations (and more).
 Note that the sailor \emph{does not necessarily want} to travel directly towards the destination:

### 2021-06-20 paper/main.tex
new and improved boat model

@@
 \newcommand{\uvec}{\vec{\hat{e}}}
-\newcommand{\tensor}[1]{\mathbf{#1}}
+\newcommand{\tensor}[1]{\mathbb{#1}}
+\renewcommand{\flat}{\text{flat}}
+\newcommand{\iso}{\text{iso}}
 \newcommand{\air}{\text{air}}
@@
 
-We are going to work in 2D, where velocities $\vec{v}$ are column vectors (or $\vec{v}\in\mathbb{R}^{2\times 1}$).
+We are going to work in 2D, with velocities $\vec{v}$ and forces $\vec{F}$ that are column vectors (or $2\times 1$ matrices).
 To stay coordinate-free, we are going to talk about angles as orientations of unit vectors like $\uvec_{\perp\sail}$ and $\uvec_{\parallel\sail}$ (which will be defined below).
@@
 If the fluid (air or water in this case) is thought of as being composed of particles that bounce elastically off of a flat surface of area $A$, then the ram-pressure force on that surface is
+\begin{align}\label{eq:flat}
+    \vec{F}_\flat &= \eta\,\rho\,A\,(\uvec_\perp^\top\vec{v})^2\,\uvec_\perp ~,
+\end{align}
+where $\eta$ is a dimensionless constant, $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the velocity of the fluid in the rest frame of the surface, and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
+Na\"ively in the particle-collision model the constant $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{Viscosities only matter for very small objects, much smaller than sailboats.
+The way to see this is HOGG (CITE).}.
+We're going to treat $\eta$ as an uninteresting constant that doesn't depend on velocity, such that it can be combined with the area $A$ in everything that follows.
+If you like thinking about tensors, the flat-surface force \eqref{eq:flat}  can be thought of as a tensor expression, in which $\rho$, $A$, and the unit vectors are combined into a three-index (order-three) tensor $\tensor{A}$, which is then contracted onto the velocity twice:
 \begin{align}
-    \vec{F} &= \eta\,\rho\,A\,(\uvec_\perp^\top\vec{v})^2\,\uvec_\perp ~,
-\end{align}
-where $\eta$ is a dimensionless constant (naively 2, but HOGG), $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the velocity of the fluid in the rest frame of the surface, and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
-If you like thinking about tensors, this can be thought of as a tensor expression, in which $\rho$, $A$, and the unit vectors are combined into a three-index (order-three) tensor $\mathbb{A}$, which is then contracted onto the velocity twice:
-\begin{align}
-    \mathbb{A} &\equiv \eta\,A\,\uvec_\perp\,\uvec_\perp^\top\otimes\uvec_\perp^\top ~.
+    \tensor{A} &\equiv \eta\,A\,\uvec_\perp\,\uvec_\perp^\top\otimes\uvec_\perp^\top ~.
 \end{align}
@@
 \begin{align}
-    [\vec{F}]_i &= \rho\,[\mathbb{A}]_{ijk}\,[\vec{v}]_j\,[\vec{v}]_k ~,
+    [\vec{F}_\flat]_i &= \rho\,[\tensor{A}]_{ijk}\,[\vec{v}]_j\,[\vec{v}]_k ~,
     \\
-    [\mathbb{A}]_{ijk} &= \eta\,A\,[\uvec_\perp]_i\,[\uvec_\perp]_j\,[\uvec_\perp]_k
+    [\tensor{A}]_{ijk} &= \eta\,A\,[\uvec_\perp]_i\,[\uvec_\perp]_j\,[\uvec_\perp]_k
     ~,
 \end{align}
-where the $[\vec{F}]_i$ notation indicates the $i$th component or coordinate of the vector, the tensor $\mathbb{A}$ has three indices, and the repeated indices are summed over.
+where the $[\vec{v}]_i$ notation indicates the $i$th component or coordinate of the vector, the tensor $\tensor{A}$ has three indices, and the repeated indices are summed over.
+If, instead of a planar object, you have an isotropic object (a sphere in fully 3D situations, a floating cylinder in our 2D water world), the ram-pressure force becomes
+\begin{align}\label{eq:iso}
+    \vec{F}_\iso &= \eta\,\rho\,A\,|\vec{v}|\,\vec{v}
+\end{align}
+where now it's a different constant $\eta$, now $A$ is the cross-sectional or projected area of the surface, and $\vec{v}$ is still the fluid velocity in the rest frame of the object.
+Again, $\eta$ depends on velocity, but again we are going to ignore that.
 
-HOGG: PARAGRAPH AND EQUATIONS ALL WRONG: In detail, the ram-pressure force $\vec{F}_\air$ from the air on the sailboat will be
-HOGG THIS EXPRESSION IS ALL WRONG. THE FORCE IS GIVEN BY DOUBLE-CONTRACTION ON AN ORDER-2 TENSOR. SEE NOTES 2021-06-19. FIX THIS.
-\begin{align}
-    \Delta\vair &\equiv \vair-\vboat
-    \\
-    \vec{F}_\air &= \rho_\air\,|\Delta\vair|\,\tensor{A}_\air\,\Delta\vair \label{eq:Fair}
-    \\
-    \tensor{A}_\air &= A_{\perp\sail}\,\uvec_{\perp\sail}\,\uvec_{\perp\sail}^\top + A_{\parallel\sail},\uvec_{\parallel\sail}\,\uvec_{\parallel\sail}^\top \label{eq:Aair}
-    \\
-    A_{\perp\sail} &\gg A_{\parallel\sail} \label{eq:sail}
-    ~,
-\end{align}
-where $\rho_\air$ is the mass density of air, $\vair,\vboat$ are the velocities of the air and boat, $\tensor{A}_\air$ is a Hermitian tensor with units of area, $A_{\perp\sail}, A_{\parallel\sail}$ are two constants with units of area defining the geometry of the sail, $\uvec_{\perp\sail}$ is a unit vector pointing normal to the surface of the (assumed planar) sail, and $\uvec_{\parallel\sail}$ is a unit vector perpendicular to that.
-If you are uncomfortable with the expression for the tensor $\tensor{A}_\air$ in \eqref{eq:Aair}, look at the discussion of tensors in, for example, \cite{kusse}.
-The inequality \eqref{eq:sail} says (loosely) that the sail (plus hull) is much bigger face-on than edge-on.
-The force $\vec{F}_\water$ from the water is similar
-\begin{align}
-    \Delta\vwater &\equiv \vwater-\vboat
-    \\
-    \vec{F}_\water &= \rho_\water\,|\Delta\vwater|\,\tensor{A}_\water\,\Delta\vwater \label{eq:Fwater}
-    \\
-    \tensor{A}_\water &= A_{\perp\keel}\,\uvec_{\perp\keel}\,\uvec_{\perp\keel}^\top + A_{\parallel\keel},\uvec_{\parallel\keel}\,\uvec_{\parallel\keel}^\top
-    \\
-    A_{\perp\keel} &\gg A_{\parallel\keel} ~,
-\end{align}
-where we have gone from air to water and sail ``s'' to keel ``k''.
-The ram-pressure forces given in \eqref{eq:Fair} and \eqref{eq:Fwater} can be justified with a simple molecular model in which air and water molecules are elastically bouncing off the sail and keel surfaces; this is the same set of assumptions that leads to, for example, the ideal gas law (HOGG CITE).
+Our model for a sailboat is going to be that it is a composite object, composed of a flat sail, a flat keel, and an isotropic partially submerged hull.
+Thus the ram-pressure force on the sailboat will have four components, a force on the flat sail from the air, a force on the flat keel from the water, a force on the hull from the air, and a force on the hull from the water.
+This model is extremely simplistic!
+In some sense it is the closest possible model to the spherical cow that will sail non-trivially.
+HOGG: COMPLETE THIS PARAGRAPH WITH EQUATIONS.
 
@@
 If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel.
-This is a quadratic root-finding problem so it has HOGG SOME SIMPLE CLOSED FORM, WHICH IS...
-we CAN ALSO MAYBE solve it numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
+This is a root-finding problem that (probably) has no closed form\footnote{It might be interesting to note that the flat-surface force \eqref{eq:flat} is a vector-valued quadratic form, so if one constructed the sailboat model completely from flat surfaces, the force law, in this approximation would be a quadratic form.
+That might lead to a closed-form solution for the equation \eqref{eq:sailing} of steady sailing.}.
+We solve equation~\eqref{eq:sailing} numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
 
@@
 HOGG: Right now, the concept of forwards and backwards is a bit soft; our boat is quadrupolar! But we will get more serious about deliberately sailing in the correct direction in the next Section.
+
+HOGG: Comment: You can't achieve zero velocity wrt either the air or the water.
 
@@
 
-HOGG PUT FIGURES HERE SHOWING TRAJECTORIES FOR SIMPLE JOURNEYS.
+HOGG PUT FIGURES HERE SHOWING TRAJECTORIES FOR SIMPLE JOURNEYS. MAYBE??
+
+\section{Boat properties and boat design}
+
+HOGG show how the hourglass plots depend on the qualities of the boat, such as: Relative area of sail and keel. Relative area of hull and keel. Etc.
 
@@
 Note that in practice, Newton's method works amazingly well.
-For Newton's method we make use of the derivatives
-\begin{align}
-    \frac{\dd\vec{F}_\air}{\dd\vboat} &= -\frac{\rho_\air}{|\Delta\vair|}\,A_\air\,\Delta\vair\,\Delta\vair^\top-\rho_\air\,|\Delta\vair|\,A_\air
-    \\
-    \frac{\dd\vec{F}_\water}{\dd\vboat} &= -\frac{\rho_\water}{|\Delta\vwater|}\,A_\water\,\Delta\vwater\,\Delta\vwater^\top-\rho_\water\,|\Delta\vwater|\,A_\water
-    ~,
-\end{align}
-which are order-2 tensors.
+For Newton's method we make use of the derivatives---HOGG GIVE HERE THE DERIVATIVES of the ram-pressure expressions---which are order-2 tensors.
 

### 2021-06-20 paper/main.tex
finished writing force law

@@
 \newcommand{\air}{\text{air}}
-\newcommand{\water}{\text{wat}}
+\newcommand{\water}{\text{water}}
 \newcommand{\boat}{\text{boat}}
@@
 \newcommand{\best}{\text{(best)}}
-\newcommand{\sail}{\text{s}}
-\newcommand{\keel}{\text{k}}
+\newcommand{\sail}{\text{sail}}
+\newcommand{\keel}{\text{keel}}
+\newcommand{\hull}{\text{hull}}
+\renewcommand{\above}{\text{above}}
+\newcommand{\below}{\text{below}}
 \newcommand{\vair}{\vec{v}_\air}
@@
 \end{align}
-where $\eta$ is a dimensionless constant, $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the velocity of the fluid in the rest frame of the surface, and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
-Na\"ively in the particle-collision model the constant $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{Viscosities only matter for very small objects, much smaller than sailboats.
+where $\eta$ is a dimensionless prefactor, $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the velocity of the fluid in the rest frame of the surface, and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
+Na\"ively in the particle-collision model the prefactor $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{Viscosities only matter for very small objects, much smaller than sailboats.
 The way to see this is HOGG (CITE).}.
-We're going to treat $\eta$ as an uninteresting constant that doesn't depend on velocity, such that it can be combined with the area $A$ in everything that follows.
+We're going to treat $\eta$ as an uninteresting constant; in particular we are going to assume that it doesn't depend on velocity, such that it can be combined with the area $A$ in everything that follows.
 If you like thinking about tensors, the flat-surface force \eqref{eq:flat}  can be thought of as a tensor expression, in which $\rho$, $A$, and the unit vectors are combined into a three-index (order-three) tensor $\tensor{A}$, which is then contracted onto the velocity twice:
@@
 \end{align}
-where now it's a different constant $\eta$, now $A$ is the cross-sectional or projected area of the surface, and $\vec{v}$ is still the fluid velocity in the rest frame of the object.
+where now it's a different prefactor $\eta$, now $A$ is the cross-sectional or projected area of the surface, and $\vec{v}$ is still the fluid velocity in the rest frame of the object.
 Again, $\eta$ depends on velocity, but again we are going to ignore that.
@@
 In some sense it is the closest possible model to the spherical cow that will sail non-trivially.
-HOGG: COMPLETE THIS PARAGRAPH WITH EQUATIONS.
+In detail the forces from the air are given by
+\begin{align}
+  \vec{F}_\air &= \vec{F}_{\sail-\air} + \vec{F}_{\hull-\air}
+  \\
+  \vec{\Delta v}_\air &\equiv \vec{v}_\air - \vec{v}_\boat
+  \\
+  \vec{F}_{\sail-\air} & = \rho_\air\,A_\sail\,(\uvec_{\perp\sail}^\top\,\vec{\Delta v}_\air)^2\,\uvec_{\perp\sail}
+  \\
+  \vec{F}_{\hull-\air} & = \rho_\air\,A_{\hull,\above}\,|\vec{\Delta v}_\air|\,\vec{\Delta v}_\air ~,
+\end{align}
+where we have dropped the $\eta$ prefactors (or really combined them with the area constants), $\vec{v}_\air,\vec{v}_\boat$ are the velocities of the air and boat, $A_\sail$ is an area related to the area of the sail, $\uvec_{\perp\sail}$ is a unit vector perpendicular to the plane of the planar sail, and $A_{\hull,\above}$ is an area related to the area of the boat's hull that is above the water surface.
+Similarly, the forces from the water are given by
+\begin{align}
+  \vec{F}_\water &= \vec{F}_{\keel-\water} + \vec{F}_{\hull-\water}
+  \\
+  \vec{\Delta v}_\water &\equiv \vec{v}_\water - \vec{v}_\boat
+  \\
+  \vec{F}_{\keel-\water} & = \rho_\water\,A_\keel\,(\uvec_{\perp\keel}^\top\,\vec{\Delta v}_\water)^2\,\uvec_{\perp\keel}
+  \\
+  \vec{F}_{\hull-\water} & = \rho_\water\,A_{\hull,\below}\,|\vec{\Delta v}_\water|\,\vec{\Delta v}_\water ~,
+\end{align}
+where the sail is replaced by the keel everywhere, and $A_{\hull,\below}$ is an area related to the area of the hull below the water surface.
+HOGG NEED WE SAY MORE ABOUT THESE? PROBABLY YES THERE ARE COMMENTS TO BE MADE.
 

### 2021-06-20 paper/main.tex
tweaks

@@
 % - Kilian Walsh
+% - Sanjoy Mahajan
+% - David Grier
 
@@
 
-Our sailboat is going to be an anisotropic object such that wind that hits the sail perpendicular to the large surface of the sail (in the rest frame of the boat) will produce more ram pressure than wind at any other angle (and far more than wind that ``goes along'' the sail direction).
+Our sailboat is going to be a simple anisotropic object such that wind that hits its planar sail produces a ram-pressure force perpendicular to the surface of the sail,
+and water that hits its planar keel produces a ram-pressure force perpendicular to the surface of the keel.
 We can construct the complete force law on the sailboat by thinking about ram-pressure forces on surfaces.
@@
 where the $[\vec{v}]_i$ notation indicates the $i$th component or coordinate of the vector, the tensor $\tensor{A}$ has three indices, and the repeated indices are summed over.
-If, instead of a planar object, you have an isotropic object (a sphere in fully 3D situations, a floating cylinder in our 2D water world), the ram-pressure force becomes
+If, instead of a planar object, you have an isotropic object (a sphere in fully 3D situations, an axi-symmetric boat in our 2D water world), the ram-pressure force becomes
 \begin{align}\label{eq:iso}
@@
 \end{align}
-where now it's a different prefactor $\eta$, now $A$ is the cross-sectional or projected area of the surface, and $\vec{v}$ is still the fluid velocity in the rest frame of the object.
-Again, $\eta$ depends on velocity, but again we are going to ignore that.
+where $\eta$ is a (different) dimensionless prefactor, $A$ is now the cross-sectional or projected area of the surface, and $\vec{v}$ is still the fluid velocity in the rest frame of the object.
+Again $\eta$ depends on velocity, but again we are going to ignore that.
 
@@
 
-Mathematically, steady sailing in steady wind and steady current (constant $\vair, \vwater$) is given by
+\section{Steady sailing}
+
+Mathematically, steady sailing (constant $\vboat$) in steady wind and steady current (constant $\vair, \vwater$) is given by
 \begin{align}\label{eq:sailing}
@@
 But turbulence; we can't ignore that!
+Constant $\eta$ is also wrong.
 

### 2021-06-20 paper/main.tex
best -> better and other changes

@@
 
-HOGG: Comment: It is easy to sail upwind. But impossible to sail upwind fttw.
+HOGG: Comment: It is easy to sail upwind. If the boat is AMAZING you can sail upwind faster than the wind! Probably not possible in practice? KLEBAN?
 
@@
 
-\section{Competitive sailing}\label{sec:racing}
+\section{Better sailing}\label{sec:better}
 
-HOGG IN THIS SECTION EXPLICITLY DEFINE ``COMPETITIVE SAILING''.
+HOGG IN THIS SECTION EXPLICITLY DEFINE ``BETTER SAILING''.
 
@@
 The sailor should then maximize the scalar product $\uvec_\destination^\top\,\vboat$
-\begin{align}\label{eq:best}
-    \uvec_{\perp\sail}^\best,\uvec_{\perp\keel}^\best &\leftarrow \argmax_{\uvec_{\perp\sail},\uvec_{\perp\keel}} \left[\uvec_\destination^\top\,(\vboat-\vdest)\right] ~,
+\begin{align}\label{eq:better}
+    \uvec_{\perp\sail}^\better,\uvec_{\perp\keel}^\better &\leftarrow \argmax_{\uvec_{\perp\sail},\uvec_{\perp\keel}} \left[\uvec_\destination^\top\,(\vboat-\vdest)\right] ~,
 \end{align}
-where $\uvec_{\perp\sail}^\best,\uvec_{\perp\keel}^\best$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and (because we are Galilean-relativistic) this velocity is relative to the velocity $\vdest$ of the destination (which might not be stationary with respect to the water).
-In this expresion \eqref{eq:best}, implicitly $\vboat$ is being thought of as a function of these orientations (and more).
+where $\uvec_{\perp\sail}^\better,\uvec_{\perp\keel}^\better$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and (because we are Galilean-relativistic) this velocity is relative to the velocity $\vdest$ of the destination (which might not be stationary with respect to the water).
+In this expresion \eqref{eq:better}, implicitly $\vboat$ is being thought of as a function of these orientations (and more).
 Note that the sailor \emph{does not necessarily want} to travel directly towards the destination:
@@
 
-Look-up tables for proper boat trim.
-
-Path planning.
+HOGG: How we implement GoodSailing(tm) and BetterSailing(tm). It's a mess!
 
@@
 
+HOGG: Even BetterSailing(tm) is not the best you can do, because you have to do complete successful path planning, and coming about costs time.
+So you will sail beyond the limits of BetterSailing(tm) if you do this correctly.
+This is out of scope for this \documentname.
+
 HOGG: All code used to make the figures in this \documentname{} are available HOGG WHERE?

### 2021-06-20 paper/main.tex
figures

@@
 \newcommand{\good}{\text{(good)}}
-\newcommand{\best}{\text{(best)}}
+\newcommand{\better}{\text{(best)}}
 \newcommand{\sail}{\text{sail}}
@@
 \newcommand{\secref}[1]{Section~\ref{#1}}
+\newcommand{\figref}[1]{Figure~\ref{#1}}
 
@@
 where $\eta$ is a dimensionless prefactor, $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the velocity of the fluid in the rest frame of the surface, and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
-Na\"ively in the particle-collision model the prefactor $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{Viscosities only matter for very small objects, much smaller than sailboats.
-The way to see this is HOGG (CITE).}.
+Na\"ively in the particle-collision model the prefactor $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{%
+Viscosities only matter for very small objects, much smaller than sailboats. The way to see this is HOGG (CITE).}.
 We're going to treat $\eta$ as an uninteresting constant; in particular we are going to assume that it doesn't depend on velocity, such that it can be combined with the area $A$ in everything that follows.
@@
   \\
-  \vec{\Delta v}_\air &\equiv \vec{v}_\air - \vec{v}_\boat
+  \vec{\Delta v}_\air &\equiv \vair - \vboat
   \\
@@
   \\
-  \vec{F}_{\hull-\air} & = \rho_\air\,A_{\hull,\above}\,|\vec{\Delta v}_\air|\,\vec{\Delta v}_\air ~,
+  \vec{F}_{\hull-\air} & = \rho_\air\,A_{\above}\,|\vec{\Delta v}_\air|\,\vec{\Delta v}_\air ~,
 \end{align}
-where we have dropped the $\eta$ prefactors (or really combined them with the area constants), $\vec{v}_\air,\vec{v}_\boat$ are the velocities of the air and boat, $A_\sail$ is an area related to the area of the sail, $\uvec_{\perp\sail}$ is a unit vector perpendicular to the plane of the planar sail, and $A_{\hull,\above}$ is an area related to the area of the boat's hull that is above the water surface.
+where we have dropped the $\eta$ prefactors (or really combined them with the area constants), $\vair,\vboat$ are the velocities of the air and boat, $A_\sail$ is an area related to the area of the sail, $\uvec_{\perp\sail}$ is a unit vector perpendicular to the plane of the planar sail, and $A_{\above}$ is an area related to the area of the boat's hull that is above the water surface.
 Similarly, the forces from the water are given by
@@
   \\
-  \vec{\Delta v}_\water &\equiv \vec{v}_\water - \vec{v}_\boat
+  \vec{\Delta v}_\water &\equiv \vwater - \vboat
   \\
@@
   \\
-  \vec{F}_{\hull-\water} & = \rho_\water\,A_{\hull,\below}\,|\vec{\Delta v}_\water|\,\vec{\Delta v}_\water ~,
+  \vec{F}_{\hull-\water} & = \rho_\water\,A_{\below}\,|\vec{\Delta v}_\water|\,\vec{\Delta v}_\water ~,
 \end{align}
-where the sail is replaced by the keel everywhere, and $A_{\hull,\below}$ is an area related to the area of the hull below the water surface.
+where the sail is replaced by the keel everywhere, and $A_{\below}$ is an area related to the area of the hull below the water surface.
 HOGG NEED WE SAY MORE ABOUT THESE? PROBABLY YES THERE ARE COMMENTS TO BE MADE.
 
-\section{Steady sailing}
+\section{Steady sailing}\label{sec:steady}
 
@@
 If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel.
-This is a root-finding problem that (probably) has no closed form\footnote{It might be interesting to note that the flat-surface force \eqref{eq:flat} is a vector-valued quadratic form, so if one constructed the sailboat model completely from flat surfaces, the force law, in this approximation would be a quadratic form.
+This is a root-finding problem that (probably) has no closed form\footnote{%
+It might be interesting to note that the flat-surface force \eqref{eq:flat} is a vector-valued quadratic form, so if one constructed the sailboat model completely from flat surfaces, the force law, in this approximation would be a quadratic form.
 That might lead to a closed-form solution for the equation \eqref{eq:sailing} of steady sailing.}.
-We solve equation~\eqref{eq:sailing} numerically in the code associated with this \documentname\footnote{HOGG GIVE LINK HERE} by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
+We solve equation~\eqref{eq:sailing} numerically in the code associated with this \documentname\footnote{%
+HOGG GIVE LINK HERE}
+by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
 
-HOGG: Give the settings of all our parameters for the figures we are about to show.
+HOGG: Define direction cosines with the wind, and with boat parts, using vectors. ADD THOSE direction cosines to the figure panels.
 
-HOGG: Define direction cosines with the wind, and with boat parts, using vectors.
+In \figref{fig:steady} we show examples of steady sailing, for a wide range of orientations of the keel and sail.
+All these examples are shown in the water rest frame.
+The boat has four free parameters $(A_\sail,A_\keel,A_{\above},A_{\below})$ all with units of area (or really three independent dimensionless ratios of these four areas).
+The values we chose for these four parameters are given in \secref{sec:design} and discussed in detail there.
+\begin{figure}[tp]
+  \caption{Visualizations of the boat velocity $\vboat$ for different settings of the orientations of the keel and sail, all presented in the water rest frame.
+    The views are top-down; the keel is represented by a thick grey line, the sail is represented by a thin black line, and the wind and boat velocities are represented by black arrows.
+    Most of the settings of the keel and sail lead to downwind travel, but a few lead to upwind travel, for example XXX and YYY. HOGG LABEL THE PANELS.\label{fig:steady}}
+\end{figure}
 
-HOGG: PUT FIGURES HERE SHOWING THE BEHAVIOR OF THE BOAT IN THE WIND.
-
+In \figref{fig:hourglass} we show the distribution of boat velocities one obtains if one randomly orients one's sail and keel....HOGG
 HOGG: Put the random-dot hourglass figure here.
@@
 
-\section{Good sailing}\label{sec:sailing}
+\section{Good sailing}\label{sec:good}
 
@@
 
-A type-A sailor---or maybe a competetive sailor---is trying to get from a current position to a destination as quickly as possible.\footnote{We don't particularly endorse this attitude towards sailing.
-The journey \emph{is} the destination.}
+A type-A sailor---or maybe a competetive sailor---is trying to get from a current position to a destination as quickly as possible\footnote{%
+We don't particularly endorse this attitude towards sailing. The journey \emph{is} the destination.}.
 Our view is that this sailor should start by specifying the unit vector $\uvec_\destination$ that points from the current position of the boat towards the destination of the boat.
@@
 
-\section{Boat properties and boat design}
+\section{Boat properties and boat design}\label{sec:design}
+
+HOGG: In the above experiments and diagrams we made use of a boat with the following properties....
+
+HOGG: These properties might seem unreasonable! However, remember that our boat is just a brutal approximation.
+A true boat has sails that are globally curved\footnote{%
+In the local-curvature or Riemann-curvature senses, sails are not (very) curved; after all, they are made from flat pieces of fabric.
+However, they are curved in the global or colloquial sense: Different patches of the sail have differently oriented normal vectors.},
+and forces that are different from these pure ram-pressure forces.
+The curvature of the sails will in general rotate the air force vectors towards the direction in which the boat is pointed.
 

### 2021-06-21 paper/main.tex
include the first figure
Updates from Overleaf

@@
 \usepackage[utf8]{inputenc}
+\usepackage{graphicx}
+\graphicspath{{./}{../ipynb/}}
 
@@
 \newcommand{\good}{\text{(good)}}
-\newcommand{\better}{\text{(best)}}
+\newcommand{\better}{\text{(better)}}
 \newcommand{\sail}{\text{sail}}
@@
 \markright{\textsf{Hogg \& Kleban / Ram-pressure model for sailing}}
+\newcommand{\figurerule}{\rule[1ex]{\textwidth}{0.2pt}}
 \linespread{1.08}
@@
 The values we chose for these four parameters are given in \secref{sec:design} and discussed in detail there.
-\begin{figure}[tp]
+\begin{figure}[t!]
+  \includegraphics[width=\textwidth]{steady.pdf}
   \caption{Visualizations of the boat velocity $\vboat$ for different settings of the orientations of the keel and sail, all presented in the water rest frame.
   The views are top-down; the keel is represented by a thick grey line, the sail is represented by a thin black line, and the wind and boat velocities are represented by black arrows.
-    Most of the settings of the keel and sail lead to downwind travel, but a few lead to upwind travel, for example XXX and YYY. HOGG LABEL THE PANELS.\label{fig:steady}}
+  Most of the settings of the keel and sail lead to downwind travel, but a few in the top-left quarter of this grid lead to upwind travel.\label{fig:steady}}
+  \figurerule
 \end{figure}
@@
 In \figref{fig:hourglass} we show the distribution of boat velocities one obtains if one randomly orients one's sail and keel....HOGG
+\begin{figure}[t!]
+  \includegraphics[width=\textwidth]{hourglass-random.pdf}
+  \caption{HOGG CAPTION HERE.\label{fig:hourglass}}
+  \figurerule
+\end{figure}
+
 HOGG: Put the random-dot hourglass figure here.
@@
 \end{align}
-where $\uvec_{\perp\sail}^\better,\uvec_{\perp\keel}^\better$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and (because we are Galilean-relativistic) this velocity is relative to the velocity $\vdest$ of the destination (which might not be stationary with respect to the water).
-In this expresion \eqref{eq:better}, implicitly $\vboat$ is being thought of as a function of these orientations (and more).
+where $\uvec_{\perp\sail}^\better,\uvec_{\perp\keel}^\better$ are the best settings of the orientations of the keel and sail, given the direction $\uvec_\destination$ to the destination, and (because we are Galilean-relativistic) this velocity is relative to the velocity $\vdest$ of the destination (which might not be stationary with respect to the water; imagine that we are sailing on a fast-flowing river).
+In this expression \eqref{eq:better}, implicitly $\vboat$ is being thought of as a function of the orientations (and more).
 Note that the sailor \emph{does not necessarily want} to travel directly towards the destination:

### 2021-06-21 paper/main.tex

@@
 % - David Grier
+
+% style notes:
+% ------------
+% - Always air before water, always sail before keel before hull.
 
@@
 \newcommand{\vdest}{\vec{v}_\destination}
+\newcommand{\mps}{\mathrm{m\,s^{-1}}}
 
@@
 
-KLEBAN: Write things about ddwfttw, maybe?
+HOGG: Why study sailing? What's the point?
+
+KLEBAN: Write things about ddwfttw, maybe? For context only!
 
 HOGG: Stuff about Bernoulli and flying and aerofoils that turned out to be wrong; a motivation to reconsider sailing too.
+
+HOGG: Who is our audience? People interested in simple mechanics.
+We eschew real fluid dynamics; we over-simplify.
+This is possible because sailing is, in fact, very simple.
+
+HOGG: We make use of good, coordinate-free expressions.
+We are Galilean-relativistic.
 
@@
 
-In \figref{fig:hourglass} we show the distribution of boat velocities one obtains if one randomly orients one's sail and keel....HOGG
+One comment to make is that if you look carefully at the panels of \figref{fig:steady}, you can see that, even in the rest frame of the water, the boat doesn't sail \emph{precisely} in the direction the keel is pointed.
+It is close in most cases, but not precisely aligned.
+This makes sense, because the wind is always---in some sense---trying to blow the sailboat downwind.
+
+It might seem strange, in \figref{fig:steady}, that the keel orientation is marked but not the direction that the boat (the prow of the boat) is pointing.
+Right now, the concept of ``forwards'' and ``backwards'' for the boat is a bit soft.
+The boat is defined, in some sense, by the force laws.
+These, in turn, are, in some sense, quadrupolar.
+So the boat doesn't really have a front and a back!
+We will get more serious about deliberately sailing in a particular direction in \secref{sec:good}.
+
+In \figref{fig:hourglass} we show the distribution of relative boat--water velocities one obtains if one randomly orients one's sail and keel.
+These orientations are literally randomly chosen angles from the uniform distribution on $[0,2\pi)$.
+Many of the resulting boat--water velocities are near zero, because randomly orienting your keel and sail is not a good way to sail!
 \begin{figure}[t!]
-  \includegraphics[width=\textwidth]{hourglass-random.pdf}
-  \caption{HOGG CAPTION HERE.\label{fig:hourglass}}
+  ~\hfill\includegraphics[width=3in]{hourglass-random.pdf}\hfill~
+  \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for a set of randomly oriented keels and sails.
+  For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:steady}.
+  Points to the left of the leftmost vertical grey line are sail--keel settings for which the boat sails upwind.
+Points to the right of the rightmost vertical grey line are sail--keel settings for which the boat sails downwind faster than the wind.
+  There are no settings that go \emph{directly} upwind, and no settings that go \emph{directly} downwind faster than the wind.\label{fig:hourglass}}
   \figurerule
@@
 
-HOGG: Put the random-dot hourglass figure here.
+There are many comments to make about the distribution of velocities shown in \figref{fig:hourglass}.
+Here are a few:
+There are many sail--keel settings that lead to travel at a boat--water velocity magnitude larger than the wind--water velocity magnitude.
+You can't make any settings that give the boat precisely zero velocity with respect to the water.
+The wind always blows it at least a bit.
+There are many sail--keel settings that lead to downwind travel \emph{faster than the wind}.
+Not \emph{directly} downwind faster than the wind, but faster than the wind.
+That is, a sailboat can beat an air balloon at downwind travel even in stationary water!
 
-HOGG: Comments: You can see from the figures that you don't sail DIRECTLY in the direction the keel is pointed.
+There are, of course, many sail--keel settings that lead to upwind travel.
+Not \emph{directly} upwind, but upwind.
+That makes sense; the whole point (in some sense) of building a model of a sailboat is to understand how it sails upwind.
+In \secref{sec:design} we will return to the upwind travel; we will show that in principle you might be able to build a boat that can travel upwind \emph{faster than the wind}.
+Indeed, this boat (specs in \secref{sec:design}) does not sail upwind very fast, and it can't sail at an angle that is even close to directly upwind.
+This closest angle to directly upwind will also be a function of the design of the boat.
 
-HOGG: Right now, the concept of forwards and backwards is a bit soft; our boat is quadrupolar! But we will get more serious about deliberately sailing in the correct direction in the next Section.
+One might ask: Intuitively, how does upwind travel work?
+You can think of the boat like a slippery wedge, with the wind pushing on one side of the wedge (the sail) and the water pushing on the other side of the wedge (the keel).
+The wedge slips forward.
+Maybe that helps?
 
-HOGG: Comment: You can't achieve zero velocity wrt either the air or the water.
-
-HOGG: Comment: It is easy to sail downwind fttw.
-
-HOGG: Comment: It is easy to sail faster than the wind speed.
-
-HOGG: Comment: It is easy to sail upwind. If the boat is AMAZING you can sail upwind faster than the wind! Probably not possible in practice? KLEBAN?
-
-HOGG: Comment: You can't sail directly into the wind, or within XXX radians of directly into the wind. This XXX depends on dimensionless sail and boat ratios.
+Finally, although upwind travel is slow going, with lots of tacking (HOGG DO WE NEED TO DEFINE THIS?) required, if you are situated \emph{on the boat}, upwind travel \emph{feels} very fast.
+Why?
+Because your experience of speed is, at least in part, the feeling of the wind in your face.
+If you think about \figref{fig:hourglass} in terms of the boat-air velocity instead of the boat-water velocity, the upwind travel has much larger boat-air velocities than the downwind travel.
+So even though you can sail fast downwind, it sometimes \emph{feels} faster upwind.
 
@@
 HOGG: FIGURE showing good sailing for a set of boat orientations.
+\begin{figure}[t!]
+  \includegraphics[width=\textwidth]{good.pdf}
+  \caption{Visualizations of GoodSailing(tm) for different settings of the orientation of the keel.
+  Conventions are the same as in \figref{fig:steady}.
+  .\label{fig:good}}
+  \figurerule
+\end{figure}
+
+HOGG: Announce and discuss \figref{fig:hourgood}.
+\begin{figure}[t!]
+  ~\hfill\includegraphics[width=3in]{hourglass-good.pdf}\hfill~
+  \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for GoodSailing(tm) for a set of randomly oriented keels.
+  For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:good}.
+  Compare this to \figref{fig:hourglass}.\label{fig:hourgood}}
+  \figurerule
+\end{figure}
 
 HOGG: Figure showing direction cosines of sail relative to keel/boat orientation.
-
-HOGG: Put the equivalent of the hourglass figure.
 
@@
 HOGG PUT FIGURES HERE SHOWING THE OPTIMAL SETTINGS OF SAIL AND KEEL ORIENTATIONS FOR DIFFERENT DESTINATION DIRECTIONS.
+
+HOGG discuss the hourglass figure.
+\begin{figure}[t!]
+  ~\hfill\includegraphics[width=3in]{hourglass-better.pdf}\hfill~
+  \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for BetterSailing(tm) for a set of randomly oriented destination vectors.
+  For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:better}.
+  Compare this to \figref{fig:hourglass} and also to \figref{fig:hourgood}.\label{fig:hourbetter}}
+  \figurerule
+\end{figure}
 

### 2021-06-21 paper/main.tex
lots of figures, figure captions, and text

@@
 % - Always air before water, always sail before keel before hull.
+% - Need to choose typesetting for GoodSailing and BetterSailing and then be consistent.
 
@@
 % ------------
-% - FIX THE FORCE EXPRESSION and discuss it in terms of inner products, tensors, and Einstein summation notation.
-% - Put in running head.
+% - FIND THE PRIOR LITERATURE!!
 % - Overleaf seems to have wrong papersize?
-% — Put in the dumbell figure
 % - Reconsider the post-equation commas? (HWR SAY).
@@
 % — Make some comments on the situations in which the ram-pressure force approximation is more appropriate---gale force or lull?
+% - Is the latex footnote-footnote separation fucked?
 
-\documentclass{article}
+\documentclass[letterpaper]{article}
 \usepackage[utf8]{inputenc}
@@
 \graphicspath{{./}{../ipynb/}}
+\usepackage{biblatex}
+\addbibresource{sailing.bib}
 
@@
 % text macros
-\usepackage{biblatex}
-\addbibresource{sailing.bib}
 \newcommand{\documentname}{\textsl{Article}}
@@
 
-HOGG: We make use of good, coordinate-free expressions.
+HOGG: We make use of good, coordinate-free expressions, a la \cite{kusse}.
 We are Galilean-relativistic.
@@
 
-Finally, although upwind travel is slow going, with lots of tacking (HOGG DO WE NEED TO DEFINE THIS?) required, if you are situated \emph{on the boat}, upwind travel \emph{feels} very fast.
+Finally, although upwind travel is slow going, with lots of tacking (HOGG DO WE NEED TO DEFINE THIS?) required, if you are situated \emph{on the boat}, upwind travel \emph{feels very fast}.
 Why?
 Because your experience of speed is, at least in part, the feeling of the wind in your face.
-If you think about \figref{fig:hourglass} in terms of the boat-air velocity instead of the boat-water velocity, the upwind travel has much larger boat-air velocities than the downwind travel.
+If you think about \figref{fig:hourglass} in terms of the boat-air velocity instead of the boat-water velocity---that is, if you Galilean-boost to the rest frame of the air---the upwind travel has much larger boat-air velocities than the downwind travel.
 So even though you can sail fast downwind, it sometimes \emph{feels} faster upwind.
@@
   Conventions are the same as in \figref{fig:steady}.
+  GoodSailing(tm) is unsuccessful in the last two cases, where the pointing is too close to directly upwind.
   .\label{fig:good}}
@@
   For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:good}.
-  Compare this to \figref{fig:hourglass}.\label{fig:hourgood}}
+  Compare this to \figref{fig:hourglass}.
+  As in \figref{fig:good}, when the keel orientation is too close to directly upwind, the boat goes downwind instead.\label{fig:hourgood}}
   \figurerule
@@
 The sailor wants to make as much progress as possible in that direction, but is willing to tack back-and-forth to get there, if it gets the boat there faster.
-So the metric really is the projection of the boat velocity onto the displacement vector pointing from the current position of the boat to the destination.
-
-While this is simple to state, the numerical optimization hurts...HOGG
+So the optimization objective really is the projection of the boat velocity onto the displacement vector pointing from the current position of the boat to the destination.
+While this is all simple to state, the global optimization is a bit painful to implement in code.
+We give details in \secref{sec:implementation}.
 
 HOGG PUT FIGURES HERE SHOWING THE OPTIMAL SETTINGS OF SAIL AND KEEL ORIENTATIONS FOR DIFFERENT DESTINATION DIRECTIONS.
+\begin{figure}[t!]
+  \includegraphics[width=\textwidth]{better.pdf}
+  \caption{Visualizations of BetterSailing(tm) for different directions to a distant destination.
+  Conventions are the same as in \figref{fig:steady} and \figref{fig:good}, except that now the destination direction is indicated with a dashed path line.
+  BetterSailing(tm) often involves sailing at a substantial angle to the direct line towards the destination.\label{fig:better}}
+  \figurerule
+\end{figure}
 
@@
 In the local-curvature or Riemann-curvature senses, sails are not (very) curved; after all, they are made from flat pieces of fabric.
-However, they are curved in the global or colloquial sense: Different patches of the sail have differently oriented normal vectors.},
+However, they are curved in the global or colloquial sense: Different patches of the sail have differently oriented normal vectors.
+It is in this colloquial sense that we are using the words ``curved'' and ``flat''.},
 and forces that are different from these pure ram-pressure forces.
@@
 
-HOGG: How we implement GoodSailing(tm) and BetterSailing(tm). It's a mess!
+HOGG: How we implement GoodSailing(tm) and BetterSailing(tm) optimizations. It's a mess!
 

### 2021-06-24 paper/main.tex
Updates from Overleaf

@@
 
+HOGG: What really is ram pressure? It is a model that the force comes mainly from momentum change or momentum transport in the fluid.
+
 HOGG: Who is our audience? People interested in simple mechanics.
@@
 In detail the forces from the air are given by
-\begin{align}
+\begin{align}\label{eq:startmodel}
   \vec{F}_\air &= \vec{F}_{\sail-\air} + \vec{F}_{\hull-\air}
   \\
-  \vec{\Delta v}_\air &\equiv \vair - \vboat
+  \vec{\Delta v}_\air &\equiv \vair - \vboat \label{eq:deltav}
   \\
-  \vec{F}_{\sail-\air} & = \rho_\air\,A_\sail\,(\uvec_{\perp\sail}^\top\,\vec{\Delta v}_\air)^2\,\uvec_{\perp\sail}
+  \vec{F}_{\sail-\air} & = \rho_\air\,A_\sail\,(\uvec_{\perp\sail}^\top\,\vec{\Delta v}_\air)^2\,\uvec_{\perp\sail} \label{eq:Fsailair}
   \\
@@
   \\
-  \vec{F}_{\hull-\water} & = \rho_\water\,A_{\below}\,|\vec{\Delta v}_\water|\,\vec{\Delta v}_\water ~,
+  \vec{F}_{\hull-\water} & = \rho_\water\,A_{\below}\,|\vec{\Delta v}_\water|\,\vec{\Delta v}_\water ~,\label{eq:endmodel}
 \end{align}
@@
 HOGG NEED WE SAY MORE ABOUT THESE? PROBABLY YES THERE ARE COMMENTS TO BE MADE.
+
+The above model could be called ``isotropic boat, thin planar sail and keel''.
+There is an alternative model to be made, using the same elements, which could be called ``parallelepiped sail, parallelepiped keel''.
+This alternative model uses only the flat-plane force law \eqref{eq:flat}, but the sail and keel each are treated as parallelepipeds with large areas $A_\sail$ and $A_\keel$ and then much smaller perpendicular areas $A_{\sail\parallel}$ and $A_{\keel\parallel}$.
+So, for example, the force from the air on the boat in this model is
+\begin{align}
+  \vec{F}_\air &= \vec{F}_{\sail-\air} + \vec{F}_{\parallel-\air}
+  \\
+  \vec{F}_{\parallel-\air} & = \rho_\air\,A_{\sail\parallel}\,(\uvec_{\parallel\sail}^\top\,\vec{\Delta v}_\air)^2\,\uvec_{\parallel\sail}
+   ~,
+\end{align}
+where $F_{\sail-\air}$ and $\vec{\Delta v}_\air$ are as defined in \eqref{eq:Fsailair} and \eqref{eq:deltav}, and $\uvec_{\parallel\sail}$ is the unit vector perpendicular to $\uvec_{\perp\sail}$.
+The water force is similarly updated.
+We have implemented both of these models, the isotropic boat with thin planar sail and keel, and the parallelepiped sail and parallelepiped keel.
+Experimentation shows that these two models give almost exactly identical results, so choose whichever model you prefer.
+In what follows, we use the \emph{isotropic boat with thin planar sail and keel} model of equations \eqref{eq:startmodel} through \eqref{eq:endmodel}.
 
@@
   GoodSailing(tm) is unsuccessful in the last two cases, where the pointing is too close to directly upwind.
-  .\label{fig:good}}
+  In these cases the the maximum of the GoodSailing(tm) objective \eqref{eq:good} is negative.\label{fig:good}}
   \figurerule
@@
   Compare this to \figref{fig:hourglass}.
-  As in \figref{fig:good}, when the keel orientation is too close to directly upwind, the boat goes downwind instead.\label{fig:hourgood}}
+  As is visible in \figref{fig:good}, when the keel orientation is too close to directly upwind, the boat goes downwind instead; there are no \emph{directly} upwind options.\label{fig:hourgood}}
   \figurerule
@@
 
-A type-A sailor---or maybe a competetive sailor---is trying to get from a current position to a destination as quickly as possible\footnote{%
+A type-A sailor---or maybe a competitive sailor---is trying to get from a current position to a destination as quickly as possible\footnote{%
 We don't particularly endorse this attitude towards sailing. The journey \emph{is} the destination.}.
@@
   Conventions are the same as in \figref{fig:steady} and \figref{fig:good}, except that now the destination direction is indicated with a dashed path line.
-  BetterSailing(tm) often involves sailing at a substantial angle to the direct line towards the destination.\label{fig:better}}
+  These plots assume that the destination is stationary with respect to the water ($\vdest-\vwater=\vec{0}$).
+  BetterSailing(tm) often involves sailing at a substantial angle to the direct line towards the destination.
+  In these cases, the idea is that the best route to the destination involves frequent tacking back and forth across the path to the destination.\label{fig:better}}
   \figurerule
@@
 
+HOGG: iso-planar vs parallelepipeds...
+
 HOGG: Take some time to bash the Bernoulli bullshit.
@@
 \paragraph{Acknowledgements:}
-It is a pleasure to thank Hans-Walter Rix (MPIA) for valuable discussions.
+It is a pleasure to thank Hans-Walter Rix (MPIA) and Soledad Villar (JHU) for valuable discussions.
 

### 2021-07-02 paper/main.tex
symmetry book

@@
 % - David Grier
+% - All those in the acks.
 
@@
 % - Need to choose typesetting for GoodSailing and BetterSailing and then be consistent.
+% - Maybe the "flat" force should be the "planar" force?
 
@@
 % - FIND THE PRIOR LITERATURE!!
-% - Overleaf seems to have wrong papersize?
+% - Should Terry Tao be cited way earlier?
+% - Make sure I cite {symmetry} appropriately in various places. Probably also point out that it is a fun longer read.
+% - Make sure we aren't claiming that the model is NOVEL. It is just SIMPLE.
+% - Should we re-write the planar force to not require that the unit vector be pointed down-wind? YES WE SHUOLD.
+% - Overleaf seems to have wrong papersize? See papersize/geometry hacks.
 % - Reconsider the post-equation commas? (HWR SAY).
@@
 \usepackage[utf8]{inputenc}
+\usepackage[letterpaper]{geometry}
 \usepackage{graphicx}
@@
 \newcommand{\mps}{\mathrm{m\,s^{-1}}}
+\newcommand{\mmps}{\mathrm{m^2\,s^{-1}}}
 
@@
     Another consequence is that it is possible for an extremely anisotropic (extremely well designed) boat to sail downwind (relative to the water) \emph{faster than the wind}, on a tacking course.
-    Interestingly, nothing about the mechanical model or discussion depend on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing.
+    Interestingly, nothing about the mechanical model or discussion depend on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; but the important thing is that the model \emph{does} depend on what's assumed about the angular re-direction of the wind momentum flux.
 \end{abstract}
@@
 
-HOGG: What really is ram pressure? It is a model that the force comes mainly from momentum change or momentum transport in the fluid.
+HOGG: What do we really mean by ram pressure?
+We'll explain in \secref{sec:ram} but briefly:
+It is a model that the force comes mainly from momentum change or momentum transport in the fluid. Here and below, compare to content (especially Section 8.4 maybe?) of \cite{symmetry}.
 
@@
 We are Galilean-relativistic.
+
+\section{Why ram-pressure forces? Why planar sails?}\label{sec:ram}
+
+Sailboats and sails---like airplanes and wings---are beautiful curved structures that are designed to redirect air with as little turbulence as possible.
+So why would we make a planar-sail approximation and model that planar sail with a ram-pressure force?
+Ram-pressure forces are generally associated with bow shocks and turbulent wakes.
+Many things (cars, launch vehicles) are designed to \emph{avoid} the ram-pressure effect known as air resistance or drag.
+
+Fundamentally, a sailboat sail works by redirecting the momentum of the air, in the rest frame of the boat.
+More specifically, there is a momentum flux (momentum per unit time) that intersects the sail (and some neighborhood around the sail).
+The air that intersects the sail and neighborhood gets redirected in direction.
+There is therefore not just a momentum flux into the sail, there is a rate of momentum change being input into the air from the sail.
+That momentum transport---the (vector) change in momentum in the air per unit time---is the (vector) force on the air from the sail.
+The sailboat sails using (in part) the equal-and-opposite reaction to this force.
+
+This momentum-transport explanation of a sail (or a wing) has an expression that is, essentially, a particle flux (number per time) times a change in momentum per particle.
+These are also the physical components of a ram-pressure force.
+Indeed, the most basic derivation of the ram-pressure force is in the microscopic model of an ideal gas, in which particles bounce elastically off the surface, changing their momentum.
+HOGG PUT A DIAGRAM HERE AND EQUATIONS.
+In this ideal-gas model, the mean force on the surface can be thought of as the combination of many tiny impulses per unit time from these elastic collisions.
+What's important about the model, however, is the momentum transport:
+The incoming momentum flux upstream of the surface is different than the outgoing momentum flux downstream of the surface.
+The force on the sail can be thought of as being generated by the vector change in that momentum flux.
+
+In this paper we are not arguing that ram pressure on flat sails is a complete explanation of a sailboat!
+We are arguing that the sailboat can be modeled as having two important surfaces---the sail and the keel---one of which deviates the flow of air in the rest-frame of the boat, and the other deviates the flow of water.
+The angles of the sail and keel with respect to the air--boat and water-boat velocity vectors set the angular deviations of the two fluids, and therefore both the magnitudes and the directions of the sail and keel forces.
+The directions of the forces can be manipulated to create amazing boat dynamics and kinematics.
+The point of this \documentname{} is that much of the amazing dynamics can be observed in an amazingly simple model for the sailboat.
+So while we use the phrases ``flat sail'' and ``ram-pressure'' in what follows to describe our model, we really mean ``simple momentum transport''.
 
@@
 We can construct the complete force law on the sailboat by thinking about ram-pressure forces on surfaces.
-If the fluid (air or water in this case) is thought of as being composed of particles that bounce elastically off of a flat surface of area $A$, then the ram-pressure force on that surface is
+As we note above, if the fluid (air or water in this case) is thought of as being composed of particles that bounce elastically off of a flat surface of area $A$, then the ram-pressure force on that surface is
 \begin{align}\label{eq:flat}
@@
 Na\"ively in the particle-collision model the prefactor $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{%
-Viscosities only matter for very small objects, much smaller than sailboats. The way to see this is HOGG (CITE).}.
+Viscosities only matter for very small objects, much smaller than sailboats.
+The way to see this is by computing the Reynolds number, which, loosely speaking, is the ratio of inertial forces to viscous forces.
+It is a typical lengthscale (like the height of the sail) times a typical velocity (like the relative velocity of the boat and air) divided by the kinematic viscosity (something like $10^{-5}\,\mmps$ for air and $10^{-6}\,\mmps$ for water).
+Typical Reynolds numbers for both the boat--air contact and the boat--water contact are therefore on the order of $10^6$ to $10^7$.
+That is, viscosity plays no first-order role in sailing.
+In detail it shapes the turbulence in the wakes coming off the sail and keel edges, but that is a tiny detail from perspective of this brutal model.}.
 We're going to treat $\eta$ as an uninteresting constant; in particular we are going to assume that it doesn't depend on velocity, such that it can be combined with the area $A$ in everything that follows.
+
 If you like thinking about tensors, the flat-surface force \eqref{eq:flat}  can be thought of as a tensor expression, in which $\rho$, $A$, and the unit vectors are combined into a three-index (order-three) tensor $\tensor{A}$, which is then contracted onto the velocity twice:
@@
 \end{align}
-where $F_{\sail-\air}$ and $\vec{\Delta v}_\air$ are as defined in \eqref{eq:Fsailair} and \eqref{eq:deltav}, and $\uvec_{\parallel\sail}$ is the unit vector perpendicular to $\uvec_{\perp\sail}$.
+where $F_{\sail-\air}$ and $\vec{\Delta v}_\air$ are as defined in \eqref{eq:Fsailair} and \eqref{eq:deltav}, and $\uvec_{\parallel\sail}$ is the unit vector perpendicular to $\uvec_{\perp\sail}$ (but also pointing downwind).
 The water force is similarly updated.
@@
 
+HOGG take the limit of infinite area ratios and show that you get absurd results. Also understand and cite the Terry Tao blog post \cite{tao}.
+
 \section{Numerical implementation notes}\label{sec:implementation}
@@
 \paragraph{Acknowledgements:}
-It is a pleasure to thank Hans-Walter Rix (MPIA) and Soledad Villar (JHU) for valuable discussions.
+It is a pleasure to thank Will Kinney (Buffalo), Hans-Walter Rix (MPIA), and Soledad Villar (JHU) for valuable discussions.
 

### 2021-07-04 paper/main.tex
bibliography and annotations

@@
 \usepackage[utf8]{inputenc}
+\usepackage{hyperref}
 \usepackage[letterpaper]{geometry}
@@
 \graphicspath{{./}{../ipynb/}}
-\usepackage{biblatex}
-\addbibresource{sailing.bib}
+% \usepackage{biblatex}
+% \addbibresource{sailing.bib}
 
@@
 This is possible because sailing is, in fact, very simple.
+We are different from the prior literature (including \cite{symmetry, explained}) because we are giving an executable model that can be played with.
+This is a sailing sandbox.
 
@@
 
-\section{Ram-pressure sailboat}\label{sec:boat}
+\section{Ram-pressure sailboat}\label{sec:model}
 
@@
   \\
-  \vec{F}_{\hull-\air} & = \rho_\air\,A_{\above}\,|\vec{\Delta v}_\air|\,\vec{\Delta v}_\air ~,
+  \vec{F}_{\hull-\air} & = \rho_\air\,A_{\above}\,|\vec{\Delta v}_\air|\,\vec{\Delta v}_\air \label{eq:Fhullair} ~,
 \end{align}
@@
 
-In \secref{sec:boat} we showed how the boat moves, given any arbitrary setting of the orientation of the keel and the sail.
+In \secref{sec:steady} we showed how the boat moves (in steady state), given any arbitrary setting of the orientation of the keel and the sail.
 However, one doesn't sail by randomly setting the keel and sail!
@@
 
-\raggedright
-\printbibliography
+% \raggedright
+\footnotesize
+\bibliographystyle{annotate}
+\bibliography{sailing.bib}
+% \printbibliography
 \end{document}

### 2021-07-05 paper/main.tex
lift, and references

@@
 \title{\bfseries%
-A simple, anisotropic ram-pressure model for sailing}
+A simple momentum-transport model for sailing}
 \author{DWH, MK, others?}
@@
 
-HOGG: Stuff about Bernoulli and flying and aerofoils that turned out to be wrong; a motivation to reconsider sailing too.
+HOGG: Stuff about Bernoulli and flying and aerofoils that turned out to be wrong (see, for one of many, many examples, \cite{lift}).
+This is something of a motivation to reconsider sailing too.
+Note that some books (especially \cite{sails}) get the lift from the sail and keel correct: It has to involve changing the direction of flow!
 
@@
 This is possible because sailing is, in fact, very simple.
-We are different from the prior literature (including \cite{symmetry, explained}) because we are giving an executable model that can be played with.
+We are different from the prior literature (including \cite{symmetry, explained, sails}) because we are giving an executable model that can be played with.
 This is a sailing sandbox.
+The only prior work we know that does precisely this is \cite{tao}.
 
@@
 
-\section{Why ram-pressure forces? Why planar sails?}\label{sec:ram}
+\section{Why ram-pressure forces?}\label{sec:ram}
 

### 2021-07-05 paper/main.tex
words, drop some bs

@@
 \pagestyle{myheadings}
-\markright{\textsf{Hogg \& Kleban / Ram-pressure model for sailing}}
 \newcommand{\figurerule}{\rule[1ex]{\textwidth}{0.2pt}}
@@
 A simple momentum-transport model for sailing}
+\markright{\textsf{Hogg \& Kleban / Momentum-transport model for sailing}}
 \author{DWH, MK, others?}
@@
     Another consequence is that it is possible for an extremely anisotropic (extremely well designed) boat to sail downwind (relative to the water) \emph{faster than the wind}, on a tacking course.
-    Interestingly, nothing about the mechanical model or discussion depend on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; but the important thing is that the model \emph{does} depend on what's assumed about the angular re-direction of the wind momentum flux.
+    Interestingly, nothing conceptual about the mechanical model or discussion depend on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; but the important thing is that the model \emph{does} depend on what's assumed about the angular re-direction of the wind momentum flux.
 \end{abstract}
@@
 It is a typical lengthscale (like the height of the sail) times a typical velocity (like the relative velocity of the boat and air) divided by the kinematic viscosity (something like $10^{-5}\,\mmps$ for air and $10^{-6}\,\mmps$ for water).
-Typical Reynolds numbers for both the boat--air contact and the boat--water contact are therefore on the order of $10^6$ to $10^7$.
+Typical Reynolds numbers for both the boat--air contact and the boat--water contact are therefore on the order of $10^6$ to $10^7$ for real boats in typical conditions.
 That is, viscosity plays no first-order role in sailing.
-In detail it shapes the turbulence in the wakes coming off the sail and keel edges, but that is a tiny detail from perspective of this brutal model.}.
-We're going to treat $\eta$ as an uninteresting constant; in particular we are going to assume that it doesn't depend on velocity, such that it can be combined with the area $A$ in everything that follows.
+In detail it shapes the turbulence in the wakes coming off the sail and keel edges, but that is a tiny detail from perspective of our brutal model.}.
+We're going to treat $\eta$ as an uninteresting constant; in particular we are going to assume that it doesn't depend on velocity, such that it can be combined with the area $A$ (to make an effective area) in everything that follows.
+If you don't like that we had to orient the unit vector $\uvec_\perp$ ``down-flow'', then you can manage the signs by re-writing \eqref{eq:flat} as
+\begin{align}
+    \vec{F}_\flat &= \eta\,\rho\,A\,|\uvec_\perp^\top\vec{v}|\,(\uvec_\perp^\top\vec{v})\,\uvec_\perp ~.
+\end{align}
 
-If you like thinking about tensors, the flat-surface force \eqref{eq:flat}  can be thought of as a tensor expression, in which $\rho$, $A$, and the unit vectors are combined into a three-index (order-three) tensor $\tensor{A}$, which is then contracted onto the velocity twice:
-\begin{align}
-    \tensor{A} &\equiv \eta\,A\,\uvec_\perp\,\uvec_\perp^\top\otimes\uvec_\perp^\top ~.
-\end{align}
-In Einstein summation notation, the three-index tensor is contracted as follows
-\begin{align}
-    [\vec{F}_\flat]_i &= \rho\,[\tensor{A}]_{ijk}\,[\vec{v}]_j\,[\vec{v}]_k ~,
-    \\
-    [\tensor{A}]_{ijk} &= \eta\,A\,[\uvec_\perp]_i\,[\uvec_\perp]_j\,[\uvec_\perp]_k
-    ~,
-\end{align}
-where the $[\vec{v}]_i$ notation indicates the $i$th component or coordinate of the vector, the tensor $\tensor{A}$ has three indices, and the repeated indices are summed over.
 If, instead of a planar object, you have an isotropic object (a sphere in fully 3D situations, an axi-symmetric boat in our 2D water world), the ram-pressure force becomes
@@
 \end{align}
-where we have dropped the $\eta$ prefactors (or really combined them with the area constants), $\vair,\vboat$ are the velocities of the air and boat, $A_\sail$ is an area related to the area of the sail, $\uvec_{\perp\sail}$ is a unit vector perpendicular to the plane of the planar sail, and $A_{\above}$ is an area related to the area of the boat's hull that is above the water surface.
+where we have dropped the $\eta$ prefactors (or really combined them with the area constants, to make the area constants \emph{effective areas}), $\vair,\vboat$ are the velocities of the air and boat, $A_\sail$ is an effective area related to the area of the sail, $\uvec_{\perp\sail}$ is a unit vector perpendicular to the plane of the planar sail, and $A_{\above}$ is an effective area related to the area of the boat's hull that is above the water surface.
 Similarly, the forces from the water are given by
@@
 \end{align}
-where the sail is replaced by the keel everywhere, and $A_{\below}$ is an area related to the area of the hull below the water surface.
+where the sail is replaced by the keel everywhere, and $A_{\below}$ is an effective area related to the area of the hull below the water surface.
 HOGG NEED WE SAY MORE ABOUT THESE? PROBABLY YES THERE ARE COMMENTS TO BE MADE.
@@
 
+HOGG: Because we have such a simple model, there might be a version for a star ship that sails by photon pressure.
+Most thinking about solar sails and reflected-photon propulsion is thinking in terms of sailing ``down wind'' of the nearby star, but with multiple sources of illumination, in principle one might be able to make a photon-pressure vehicle that can sail or tack in arbitrary directions, given illumination from two or more stars.
+HOGG: Someone must have thought of all this before!
+
 HOGG: Even BetterSailing(tm) is not the best you can do, because you have to do complete successful path planning, and coming about costs time.

### 2021-07-05 paper/main.tex
intro text

@@
 
-HOGG: Why study sailing? What's the point?
+We (the authors) are not sailors!
+However, we believe that sailing is a beautiful technology.
+The boat is powered by the wind, and yet it can travel (possibly with a tacking course) in any direction.
+Of course it's not Galilean-relativistic to say that the boat is powered by the wind:
+There is no absolute velocity, and there is a valid rest frame in which the wind velocity is zero.
+The Galilean-relativistic thing to say is that the boat is powered by the relative velocity between the air and the water (the symmetry between the air and the water in the mechanisms of sailing is the fundamental theme of \cite{symmetry}).
+Here we build the simplest possible physical model for a sailboat and sailing that captures all of the leading-order features of the technology.
 
-KLEBAN: Write things about ddwfttw, maybe? For context only!
+KLEBAN: Write things about ddwfttw, maybe? For context only! But it is worth saying that if we can understand sailing with a simple model, we might resolve some disputes about ddwfttw.
 
-HOGG: Stuff about Bernoulli and flying and aerofoils that turned out to be wrong (see, for one of many, many examples, \cite{lift}).
-This is something of a motivation to reconsider sailing too.
-Note that some books (especially \cite{sails}) get the lift from the sail and keel correct: It has to involve changing the direction of flow!
+When we were little, we were taught a story about how airfoils (wings on airplanes, for example) generate lift, in which the air path over the top of the lenticular, curved wing is longer than the path under the bottom, so velocities are different (top to bottom), and thus there is a net force from a difference in Bernoulli pressures.
+This model is now known to be wrong (see, for example, \cite{lift}):
+There is no reason for the air flow to ``join up'' at the end of the airfoil.
+And, besides, airplanes can fly upside-down!
+The best leading-order description is that the lift on the wing comes from \emph{redirection of the flow of air}.
+There is a volumetric rate of momentum change downwards in the air flowing onto the wing, which leads to an equal-and-opposite force upwards on the wing.
+Similarly, the force on a sailboat sail is from the redirection of the flow of the air encountering the sail.
+This point is made admirably in some books on the physics of sailing (especially \cite{sails}).
+And the force on a sailboat keel is from the redirection of the flow of water encountering the keel.
+These forces are sometimes called ``lift'' forces in analogy to the lift force on an airplane wing.
+This \documentname{} is inspired by these observations.
+Sailboats sail by managing momentum changes in the air and water.
+They work on \emph{momentum transport}.
 
@@
 This is a sailing sandbox.
-The only prior work we know that does precisely this is \cite{tao}.
+The only prior work we know that does precisely this is \cite{tao}, but that model is very simplistic (for example, it doesn't lead to steady-state sailing).
 
@@
 \begin{align}
-    \vec{F}_\flat &= \eta\,\rho\,A\,|\uvec_\perp^\top\vec{v}|\,(\uvec_\perp^\top\vec{v})\,\uvec_\perp ~.
+    \vec{F}_\flat &= \eta\,\rho\,A\,|\uvec_\perp^\top\vec{v}|\,(\uvec_\perp^\top\vec{v})\,\uvec_\perp ~;
 \end{align}
+in this formulation, $\uvec_\perp$ can point in either of the two directions that are normal to the surface.
 
@@
 This model is extremely simplistic!
-In some sense it is the closest possible model to the spherical cow that will sail non-trivially.
+In some sense it is the closest possible object to the spherical cow that will sail non-trivially.
 In detail the forces from the air are given by

### 2021-07-05 paper/main.tex
zeroth draft of the introduction

@@
 % - Kilian Walsh
+% - Hyung Don Ryoo
 % - Sanjoy Mahajan
@@
 % - Make sure we aren't claiming that the model is NOVEL. It is just SIMPLE.
-% - Should we re-write the planar force to not require that the unit vector be pointed down-wind? YES WE SHUOLD.
-% - Overleaf seems to have wrong papersize? See papersize/geometry hacks.
 % - Reconsider the post-equation commas? (HWR SAY).
-% — Find references for, and destroy, Bernoulli.
 % — Make some comments on the situations in which the ram-pressure force approximation is more appropriate---gale force or lull?
 % - Is the latex footnote-footnote separation fucked?
+% - Do we still mention "tensors" anywhere? make sure the abstract is consistent with the body.
 
@@
 \usepackage[utf8]{inputenc}
+\usepackage[letterpaper]{geometry} % because Overleaf is weird.
 \usepackage{hyperref}
-\usepackage[letterpaper]{geometry}
 \usepackage{graphicx}
@@
 There is no absolute velocity, and there is a valid rest frame in which the wind velocity is zero.
-The Galilean-relativistic thing to say is that the boat is powered by the relative velocity between the air and the water (the symmetry between the air and the water in the mechanisms of sailing is the fundamental theme of \cite{symmetry}).
-Here we build the simplest possible physical model for a sailboat and sailing that captures all of the leading-order features of the technology.
+The Galilean-relativistic thing to say is that the boat is powered by the relative velocity between the air and the water.
+Indeed, the symmetry between the air and the water in the mechanisms of sailing is the fundamental theme of one of the better books on the physics of sailing \cite{symmetry}.
+Here we build the simplest possible physical model for a sailboat---and sailing---that captures all of the leading-order features of the technology.
 
@@
 These forces are sometimes called ``lift'' forces in analogy to the lift force on an airplane wing.
-This \documentname{} is inspired by these observations.
+This \documentname{} is partially motivated by these observations.
 Sailboats sail by managing momentum changes in the air and water.
@@
 
-HOGG: What do we really mean by ram pressure?
-We'll explain in \secref{sec:ram} but briefly:
-It is a model that the force comes mainly from momentum change or momentum transport in the fluid. Here and below, compare to content (especially Section 8.4 maybe?) of \cite{symmetry}.
+There are multiple books about the physics of sailing (including \cite{symmetry, explained, sails}).
+These books vary in their physical correctness and their comprehensiveness.
+They are mainly aimed at sailors who want to understand better their boats and how to design them and sail them.
+This paper is aimed at a different audience.
+This paper is aimed at physicists and students of physics interested in a simple, implementable, physically justifiable model of sailing.
+This model is a sandbox for experimenting with wind-powered travel and wind-powered machines, which are, in turn, key concepts in sustainable technology.
+Our model of sailing is also something that can be implemented by students new to physics, in extremely simple code (an example of which is released in conjunction with this \documentname\footnote{All code used in this \documentname{} is available under a HOGG WHAT LICENSE?
+at \url{https://github.com/davidwhogg/Sailing}.}).
 
-HOGG: Who is our audience? People interested in simple mechanics.
-We eschew real fluid dynamics; we over-simplify.
-This is possible because sailing is, in fact, very simple.
-We are different from the prior literature (including \cite{symmetry, explained, sails}) because we are giving an executable model that can be played with.
-This is a sailing sandbox.
-The only prior work we know that does precisely this is \cite{tao}, but that model is very simplistic (for example, it doesn't lead to steady-state sailing).
+Importantly in what follows, we deliver only zeroth-order physical explanations.
+In detail, the flow of air over a sail, and water over a keel, involves boundary layers, turbulence, and all the subtleties and complexities of fluid dynamics.
+We drop all of that in favor of a pure-momentum model that erases all of these critical details, and boils all of this down to some effective areas and direction vectors.
+This over-simplification might be frustrating to sailing experts, or fluid dynamicists.
+But our goal is to find the simplest possible, dimensionally correct, Galilean-symmetric rules for a boat that can sail non-trivially.
+You might say that this is an astrophysics-quality model of a sailboat!
+The only prior work we know in this space is \cite{tao}, but that model is (in our view) overly simplistic.
+For example, it doesn't lead to steady-state sailing because the drag forces vanish.
 
-HOGG: We make use of good, coordinate-free expressions, a la \cite{kusse}.
-We are Galilean-relativistic.
+Above, we've mentioned Galilean relativity:
+It is important that any model for a sailboat precisely obey Galilean relativity; it is an extremely important symmetry for understanding sailing \cite{symmetry}.
+Indeed---according to lore---Galilean relativity was discovered by consideration of doing physics experiments in the hold of a sailing ship \cite{galileo}.
+The contemporary view of Galilean relativity is that all velocities that appear in physical law or the prediction of observables must be relative velocities.
+In addition to obeying this symmetry, our model is consistent with dimensional analysis and is equivariant to rotations and reflections.
+We also make use of good, coordinate-free expressions, in the style of (for example) \cite{kusse}.
+
+Our model for forces on a sailboat presented below in \secref{sec:model} is a very generic momentum-transport model.
+In astrophysics contexts, forces from momentum transport from winds or relative motions of bodies and fluids are called ``ram-pressure'' forces.
+We use that term a bit in what follows, but since it means different things to different people, it's fine to substitute in the word ``lift'' for ``ram-pressure'' everywhere.
+We'll discuss these forces conceptually in \secref{sec:ram}, give Galilean-relativistic expressions in \secref{sec:model}, and use them to explain the steady-state motion of a sailboat in \secref{sec:steady}.
+We follow this modeling with some optimization, useful to an idealized sailor on our idealized boat, in \secref{sec:good} and \secref{sec:better}.
+When we optimize the trim of our boat, we learn about what things the boat can do, in terms of sailing speed, and sailing upwind and downwind.
+It's impressive!
+We will discuss the dependence of our results on our boat properties in \secref{sec:design} and the limitations and extensions of our model in \secref{sec:discussion}.
 
@@
 A true boat has sails that are globally curved\footnote{%
-In the local-curvature or Riemann-curvature senses, sails are not (very) curved; after all, they are made from flat pieces of fabric.
+In the local-curvature or Riemann-curvature senses, sails are often not (very) curved; after all, they are often made from flat pieces of fabric.
 However, they are curved in the global or colloquial sense: Different patches of the sail have differently oriented normal vectors.
-It is in this colloquial sense that we are using the words ``curved'' and ``flat''.},
-and forces that are different from these pure ram-pressure forces.
-The curvature of the sails will in general rotate the air force vectors towards the direction in which the boat is pointed.
+It is in this colloquial sense that we are using the words ``curved'' and ``flat''.
+All that said, the most sophisticated sails in the world are also curved in the local sense \cite{sails}.},
+and forces that are different in detail from these pure ram-pressure forces.
+HOGG: REPHRASE: The curvature of the sails will in general rotate the air force vectors towards the direction in which the boat is pointed.
 

### 2021-07-06 paper/main.tex
Kimball and literature

@@
 The boat is powered by the wind, and yet it can travel (possibly with a tacking course) in any direction.
-Of course it's not Galilean-relativistic to say that the boat is powered by the wind:
+Of course it's not consistent with Galilean relativity to say that the boat is powered by the wind:
 There is no absolute velocity, and there is a valid rest frame in which the wind velocity is zero.
@@
 Indeed, the symmetry between the air and the water in the mechanisms of sailing is the fundamental theme of one of the better books on the physics of sailing \cite{symmetry}.
-Here we build the simplest possible physical model for a sailboat---and sailing---that captures all of the leading-order features of the technology.
+Here we build the simplest possible physical model for a sailboat---and sailing---that is consistent with symmetries and captures all of the leading-order features of the technology.
 
@@
 
-There are multiple books about the physics of sailing (including \cite{symmetry, explained, sails}).
+There are multiple books about the physics of sailing (including \cite{symmetry, explained, sails, pos}).
 These books vary in their physical correctness and their comprehensiveness.
@@
 The only prior work we know in this space is \cite{tao}, but that model is (in our view) overly simplistic.
-For example, it doesn't lead to steady-state sailing because the drag forces vanish.
+For example, it doesn't lead to steady-state sailing because the drag forces (in that model) vanish (HOGG CHECK).
+In principle also \cite{pos} delivers similar equations to what we derive below, based on similar considerations and similar approximations, but their goals are not really for their reader to build an implementable model.
 
@@
 If you don't like that we had to orient the unit vector $\uvec_\perp$ ``down-flow'', then you can manage the signs by re-writing \eqref{eq:flat} as
-\begin{align}
+\begin{align}\label{eq:flat2}
     \vec{F}_\flat &= \eta\,\rho\,A\,|\uvec_\perp^\top\vec{v}|\,(\uvec_\perp^\top\vec{v})\,\uvec_\perp ~;
@@
 in this formulation, $\uvec_\perp$ can point in either of the two directions that are normal to the surface.
+The force law \eqref{eq:flat} and \eqref{eq:flat2} is very similar to what's derived or discussed in \cite{symmetry} (Section~8.4) and in \cite{pos} (Chapters~2, 3).
 
-If, instead of a planar object, you have an isotropic object (a sphere in fully 3D situations, an axi-symmetric boat in our 2D water world), the ram-pressure force becomes
+If, instead of a planar object, you have an isotropic object (a sphere in fully 3D situations, an axisymmetric boat in our 2D water world), the ram-pressure force becomes
 \begin{align}\label{eq:iso}
@@
 
-HOGG show how the hourglass plots depend on the qualities of the boat, such as: Relative area of sail and keel. Relative area of hull and keel. Etc.
+HOGG: Introduce drag ratios or lift ratios for the air and the water.
+And then show how the hourglass plots depend on these ratios.
+such as: Relative area of sail and keel.
+Relative area of hull and keel.
+Etc.
 
@@
 
+HOGG: This model implies, indirectly, that keels should be curved! And they are, in high-end sailing (HOGG cite?)
+
 HOGG: Because we have such a simple model, there might be a version for a star ship that sails by photon pressure.

### 2021-07-06 paper/main.tex
Tao re-read; GoodSailing(tm)

@@
 % ------------
-% - FIND THE PRIOR LITERATURE!!
-% - Should Terry Tao be cited way earlier?
+% - Change citations to Tao to make it clear that the model is wrong!!
 % - Make sure I cite {symmetry} appropriately in various places. Probably also point out that it is a fun longer read.
@@
 \usepackage[letterpaper]{geometry} % because Overleaf is weird.
-\usepackage{hyperref}
+\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
 \usepackage{graphicx}
@@
 \setlength{\textwidth}{5.00in}
+\setlength{\oddsidemargin}{3.25in}
+\addtolength{\oddsidemargin}{-0.5\textwidth}
 \setlength{\textheight}{9.00in}
@@
 It's impressive!
-We will discuss the dependence of our results on our boat properties in \secref{sec:design} and the limitations and extensions of our model in \secref{sec:discussion}.
+We will discuss the dependence of our results on our boat properties in \secref{sec:design}.
+We comment on the implemntation of the model in \secref{sec:implementation} and we remark on limitations and extensions in \secref{sec:discussion}.
 
@@
 However, one doesn't sail by randomly setting the keel and sail!
-One sails by pointing the boat in some direction and then setting the sail (or sails) appropriately.
+One sails by pointing the boat in some direction and setting the sail (or sails) appropriately given that boat pointing, or vice versa.
 In our simple, ram-pressure boat, how should we set the planar sail?
 
-HOGG: IN THIS SECTION EXPLICITLY DEFINE ``GOOD SAILING''.
+We could define \emph{GoodSailing(tm)} to be setting the sail orientation optimally, given the orientation of the boat.
+This might look something like the following:
 \begin{align}\label{eq:good}
@@
 \end{align}
-where .... HOGG IMPLICITLY $\vboat-\vwater$ is a function of all the properties of the world and the boat boat, along with the orientation of its keel and sail.
-
-HOGG: FIGURE showing good sailing for a set of boat orientations.
+where $\uvec_{\perp\sail}^\good$ is the best orientation of the (planar) sail conditioned on the orientation of the keel,
+$\uvec_{\parallel\keel}$ is the unit vector pointing along the keel (parallel to the keel, in the forwards direction),
+the inner product gives the projection of the boat--water velocity vector onto the boat heading direction,
+and (for optimization purposes) the velocity difference $\vboat-\vwater$ is implicitly a function of the sail and keel orientations, according to the model of \secref{sec:steady}.
+This optimization is one-dimensional, but it is a global optimization, so it requires some attention; details of our implmentation of the optimization are given in \secref{sec:implementation}.
+Some examples of GoodSailing(tm) settings of sail and keel are shown in \figref{fig:good}, along with the resulting boat velocity vectors.
+A summary of all possible boat--water velocity vectors possible, for GoodSailing(tm) at many different keel orientations, is shown in \figref{fig:hourgood}.
 \begin{figure}[t!]
@@
 \end{figure}
-
-HOGG: Announce and discuss \figref{fig:hourgood}.
 \begin{figure}[t!]
@@
   Compare this to \figref{fig:hourglass}.
-  As is visible in \figref{fig:good}, when the keel orientation is too close to directly upwind, the boat goes downwind instead; there are no \emph{directly} upwind options.\label{fig:hourgood}}
+  As is visible in \figref{fig:good}, when the keel orientation is too close to directly upwind, the boat goes downwind instead; there are no \emph{directly} upwind options.
+This figure is directly comparable to [HOGG WHAT FIGURES IN THE BOOKS ABOUT SAILING?].\label{fig:hourgood}}
   \figurerule
@@
 
-HOGG: Figure showing direction cosines of sail relative to keel/boat orientation.
+HOGG: Figure showing direction cosines of sail relative to keel/boat orientation?
 
-HOGG: Comment: The settings of the sail look ``tight'' relative to standard sailing practice.
+HOGG: Comments on GoodSailing(tm): As we saw in \secref{sec:steady}, you can sail dwfttw, and you can sail upwind, and you can sail fttw in general.
+But you can't sail dufttw or ddfttw.
+The settings of the sail look ``tight'' relative to standard sailing practice.
 But this boat has flat sails, so it's hard to compare the exact settings with sailor experiences (which are all with curved sails).
 
-HOGG: Final comment: But this is not the BEST a sailor can do...
+HOGG: Final comment: But this is not the BEST a sailor can do, because (a) boat doesn't move in the $\uvec_{\parallel\keel}$ direction, (b) because the destination might also be moving wrt the water, and (c) because you might want to take a tacking path...
 
@@
 % \raggedright
-\footnotesize
-\bibliographystyle{annotate}
+\bibliographystyle{hogg-plain-annote}
 \bibliography{sailing.bib}

### 2021-07-08 paper/main.tex
small changes

@@
 \author{DWH, MK, others?}
-\date{June 2021}
+\date{2021}
 \begin{document}
@@
 A summary of all possible boat--water velocity vectors possible, for GoodSailing(tm) at many different keel orientations, is shown in \figref{fig:hourgood}.
+This latter figure is the HOGG WHAT DIAGRAM that is standard in all books on the subject of sailing (HOGG GIVE REFERENCES).
 \begin{figure}[t!]
@@
 
-HOGG: Comments on GoodSailing(tm): As we saw in \secref{sec:steady}, you can sail dwfttw, and you can sail upwind, and you can sail fttw in general.
-But you can't sail dufttw or ddfttw.
-The settings of the sail look ``tight'' relative to standard sailing practice.
-But this boat has flat sails, so it's hard to compare the exact settings with sailor experiences (which are all with curved sails).
+There are many comments to be made here about GoodSailing(tm), but mainly the point is that GoodSailing(tm) confirms what we saw in \secref{sec:steady}.
+You can sail upwind, you can sail faster than the wind, and you can even sail downwind faster than the wind.
+You can't sail \emph{directly} upwind, and you can't sail \emph{directly} downwind faster than the wind.
+You can't give the boat precisely zero velocity with respect to either the water or the air, and the boat doesn't sail exactly in the direction it is pointed (it sails downwind of that, relative to the water).
+In detail, to an experienced sailor, the trim of the sail in \figref{fig:good} might look a little ``tight'' relative to standard best practice.
+But of course our sail is a planar approximation to a real sail; the curvature of the sail makes these angles wrong in detail (the angles shown in \figref{fig:good} are not quite right for either the boom or the bulk of any curved sail).
+The details of \figref{fig:hourgood} will depend on the physical properties or design of the boat.
+We return to this point in \secref{sec:design}.
 
-HOGG: Final comment: But this is not the BEST a sailor can do, because (a) boat doesn't move in the $\uvec_{\parallel\keel}$ direction, (b) because the destination might also be moving wrt the water, and (c) because you might want to take a tacking path...
+GoodSailing(tm) is good!
+But it is not the \emph{best} a sailor can do, for various reasons:
+One is that the boat does not move relative to the water exactly in the $\uvec_{\parallel\keel}$ direction, so the optimization for the projection onto $\uvec_{\parallel\keel}$ might not be quite right.
+Another is that the destination might also be moving with respect to the water (for example, if the destination is on the banks of a flowing river), so the relative velocity in \eqref{eq:good} is the wrong relative velocity.
+Yet another is that you might have a destination in mind that is directly upwind or directly downwind and you will need to not just optimize your velocity but instead take a non-trivial tacking path.
+We address a few of these next.
 
@@
 
-Now if you want to plan a complete path or journey for the sailboat in (unrealistically) steady wind, you... HOGG
-
-HOGG PUT FIGURES HERE SHOWING TRAJECTORIES FOR SIMPLE JOURNEYS. MAYBE??
+BetterSailing(tm) is better than GoodSailing(tm).
+However, BetterSailing(tm) is still not the best we can do.
+HOGG: Now if you want to plan a complete path or journey for the sailboat in (unrealistically) steady wind, you have to know yet more things... out of scope? ... HOGG
 

### 2021-07-09 paper/main.tex
delete
better and design sections

@@
 % ------------
+% - Put (m/s) onto hourglass figure captions.
+% - Make figure comparing different boats.
 % - Change citations to Tao to make it clear that the model is wrong!!
@@
 \graphicspath{{./}{../ipynb/}}
-% \usepackage{biblatex}
-% \addbibresource{sailing.bib}
 
@@
 \newcommand{\vdest}{\vec{v}_\destination}
+\newcommand{\mm}{\mathrm{m^2}}
 \newcommand{\mps}{\mathrm{m\,s^{-1}}}
@@
 Our view is that this sailor should start by specifying the unit vector $\uvec_\destination$ that points from the current position of the boat towards the destination of the boat.
-The sailor should then maximize the scalar product $\uvec_\destination^\top\,\vboat$
+The sailor should then maximize the scalar product of the boat's velocity (relative to the destination) onto the unit vector pointing to the destination.
+That is, we could define BetterSailing(tm) as
 \begin{align}\label{eq:better}
@@
 
-HOGG PUT FIGURES HERE SHOWING THE OPTIMAL SETTINGS OF SAIL AND KEEL ORIENTATIONS FOR DIFFERENT DESTINATION DIRECTIONS.
+We show examples of BetterSailing(tm) in \figref{fig:better}, where we show the direction to the destination and the settings of the keel and sail.
+In these examples, we have set the velocity of the destination equal to the velocity of the water, but this will not be the case in general.
+In \figref{fig:hourbetter} we show the velocities obtained in BetterSailing(tm), which are a subset of the velocities obtained in GoodSailing(tm).
+In particular, the velocities that are not optimal for some particular destination are not used.
 \begin{figure}[t!]
@@
 \end{figure}
-
-HOGG discuss the hourglass figure.
 \begin{figure}[t!]
@@
 
-BetterSailing(tm) is better than GoodSailing(tm).
-However, BetterSailing(tm) is still not the best we can do.
-HOGG: Now if you want to plan a complete path or journey for the sailboat in (unrealistically) steady wind, you have to know yet more things... out of scope? ... HOGG
+BetterSailing(tm) is better than GoodSailing(tm), because BetterSailing(tm) takes better account of the sailor's objectives.
+However, BetterSailing(tm) is \emph{still} not the best we can do.
+For example, a sailor must in general take non-trivial tacking paths from origin to destination (especially when the destination is upwind, but even when the destination is downwind).
+The corners of that path---``coming about'' in sailing parlance---cost time, so their frequency should be balanced against slightly non-optimal sailing directions.
+Thus an even better sailor will put the boat into configurations that don't appear in \figref{fig:hourbetter}.
+The full exposition of this requires a model for the costs of coming about.
+That's outside the scope of this \documentname.
 
@@
 
-HOGG: In the above experiments and diagrams we made use of a boat with the following properties....
+In all of the experiments and figures given above, we made use of a model boat with the following effective areas:
+\begin{align}
+  A_\sail &= XXX\,\mm \\
+  A_\keel &= YYY\,\mm \\
+  A_\above &= ZZZ\,\mm \\
+  A_\below &= QQQ\,\mm ~.
+\end{align}
+This boat is designed such that $\rho_\air\,A_\sail = \rho_\water\,A_\keel$, which seems sensible, but is perhaps not industry standard.
+This boat also has very good drag ratios or lift ratios:
+The ratio of effective area of sail to hull-above is 256, as is the ratio of effective area of keel to hull-below.
+All these ratios were set not by analyzing any real boats, but by making the velocity curve shown in \figref{fig:hourgood} look reasonable, or accord with our intuitions and expectations.
+HOGG: NEED A BETTER NAME THAN VELOCITY CURVE; WHAT IS THIS DIAGRAM CALLED?
+Recall, for one, that our boat is a brutal approximation!
+For two, these areas are \emph{effective areas} within this brutal approximation, not measured areas.
 
-HOGG: These properties might seem unreasonable! However, remember that our boat is just a brutal approximation.
+As we change the ratios of the effective areas, the shapes of the velocity curves change.
+In \figref{fig:design} we show the effect on the boat velocity curves of changing the ratios $A_\sail/A_\keel$, $A_\sail/A_\above$, and $A_\keel/A_\below$....HOGG
+\begin{figure}[t!]
+  ~\hfill HOGG \hfill~
+  \caption{Same as \figref{fig:hourgood} but for varying values of the three ratios $A_\sail/A_\keel$, $A_\sail/A_\above$, and $A_\keel/A_\below$.\label{fig:design}}
+  \figurerule
+\end{figure}
+
+HOGG: Comment on what we seen in \figref{fig:design}.
+HOGG take the limit of infinite area ratios and show that you get absurd results.
+
+HOGG: 
 A true boat has sails that are globally curved\footnote{%
@@
 Etc.
-
-HOGG take the limit of infinite area ratios and show that you get absurd results. Also understand and cite the Terry Tao blog post \cite{tao}.
 

### 2021-07-09 paper/main.tex
dwsign

@@
 Maybe that helps?
+HOGG SAYS: I THINK WE NEED A DIAGRAM HERE, SHOWING DWFTTW and UW.
 
@@
 For two, these areas are \emph{effective areas} within this brutal approximation, not measured areas.
+Probably the least realistic of these effective-area ratios is $A_\keel/A_\below$; probably real boats rarely have this keel lift ratio in the hundreds.
+In this sense, maybe our model is more appropriate for an ice yacht than a sailboat?
 

### 2021-07-11 paper/main.tex
numbers

@@
 % ------------
-% - Put (m/s) onto hourglass figure captions.
 % - Make figure comparing different boats.
@@
 % - Do we still mention "tensors" anywhere? make sure the abstract is consistent with the body.
+% - What claims in the literature exist about maximum boat speeds?
+Have I seen a 2x wind claim?
+We can beat that.
+Have I seen a "never faster downwind than the wind" or than 0.5 times the wind claim?
+We can beat that too.
 
@@
   ~\hfill\includegraphics[width=3in]{hourglass-good.pdf}\hfill~
-  \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for GoodSailing(tm) for a set of randomly oriented keels.
+  \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for GoodSailing(tm) for a set keel orientations.
   For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:good}.
@@
   ~\hfill\includegraphics[width=3in]{hourglass-better.pdf}\hfill~
-  \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for BetterSailing(tm) for a set of randomly oriented destination vectors.
+  \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for BetterSailing(tm) for a set of destination vectors of different orientations.
   For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:better}.
@@
 \begin{align}
-  A_\sail &= XXX\,\mm \\
-  A_\keel &= YYY\,\mm \\
-  A_\above &= ZZZ\,\mm \\
-  A_\below &= QQQ\,\mm ~.
+  A_\sail &= 1400\,\mm \\
+  A_\keel &= 2\,\mm \\
+  A_\above &= 5.6\,\mm \\
+  A_\below &= 0.008\,\mm ~.
 \end{align}
@@
 This boat also has very good drag ratios or lift ratios:
-The ratio of effective area of sail to hull-above is 256, as is the ratio of effective area of keel to hull-below.
+The ratio of effective area of sail to hull-above is 250, as is the ratio of effective area of keel to hull-below.
 All these ratios were set not by analyzing any real boats, but by making the velocity curve shown in \figref{fig:hourgood} look reasonable, or accord with our intuitions and expectations.

### 2021-07-15 paper/main.tex
small tweaks and notes

@@
 We will discuss the dependence of our results on our boat properties in \secref{sec:design}.
-We comment on the implemntation of the model in \secref{sec:implementation} and we remark on limitations and extensions in \secref{sec:discussion}.
+We comment on the implementation of the model in \secref{sec:implementation} and we remark on limitations and extensions in \secref{sec:discussion}.
 
@@
 
-Fundamentally, a sailboat sail works by redirecting the momentum of the air, in the rest frame of the boat.
+Fundamentally, a sailboat sail works by redirecting the momentum of the air.
+%, in the rest frame of the boat.
+%It does that in every frame!
 More specifically, there is a momentum flux (momentum per unit time) that intersects the sail (and some neighborhood around the sail).
-The air that intersects the sail and neighborhood gets redirected in direction.
-There is therefore not just a momentum flux into the sail, there is a rate of momentum change being input into the air from the sail.
+The air that intersects the sail and neighborhood gets redirected, and so there is  a rate of momentum change on the air due to the sail.
+%in direction.
+%There is therefore not just a momentum flux into the sail, there is a rate of momentum change being input into the air from the sail.
 That momentum transport---the (vector) change in momentum in the air per unit time---is the (vector) force on the air from the sail.
-The sailboat sails using (in part) the equal-and-opposite reaction to this force.
+Half of the explanation for how a sailboat sails uses the equal-and-opposite reaction to this force.
 
-This momentum-transport explanation of a sail (or a wing) has an expression that is, essentially, a particle flux (number per time) times a change in momentum per particle.
+This momentum-transport explanation of a sail (or a wing) 
+%has an expression that is, essentially,
+can be expressed as a particle flux (number per time) times a change in momentum per particle.
 These are also the physical components of a ram-pressure force.
@@
 The incoming momentum flux upstream of the surface is different than the outgoing momentum flux downstream of the surface.
-The force on the sail can be thought of as being generated by the vector change in that momentum flux.
+%The force on the sail can be thought of as being generated by the vector change in that momentum flux.
+%A bit repetitive, possibly trim more
 
@@
 \end{align}
-where $\eta$ is a dimensionless prefactor, $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the velocity of the fluid in the rest frame of the surface, and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
+where $\eta$ is a dimensionless prefactor, $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the %velocity of the fluid in the rest frame of the surface,
+relative velocity between the fluid and the surface,
+and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
+MK - I'M NOT A FAN OF THIS NOTATION FOR INNER PRODUCT, ESPECIALLY WITH UPPER AND LOWER Ts.
+
 Na\"ively in the particle-collision model the prefactor $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{%
@@
 \begin{figure}[t!]
-  ~\hfill HOGG \hfill~
-  \caption{Same as \figref{fig:hourgood} but for varying values of the three ratios $A_\sail/A_\keel$, $A_\sail/A_\above$, and $A_\keel/A_\below$.\label{fig:design}}
+  \includegraphics[width=0.5\textwidth]{design_A.pdf}%
+  \includegraphics[width=0.5\textwidth]{design_C.pdf}\\%
+  ~\hfill\includegraphics[width=0.5\textwidth]{design_E.pdf}\hfill~
+  \caption{Each panel is analagous to \figref{fig:hourgood} but in each case we have varied one of the dimensionless ratios....HOGG....for varying values of the three ratios $A_\sail/A_\keel$, $A_\sail/A_\above$, and $A_\keel/A_\below$.\label{fig:design}}
   \figurerule
@@
 HOGG: Comment on what we seen in \figref{fig:design}.
-HOGG take the limit of infinite area ratios and show that you get absurd results.
+HOGG: Note that if you set $\rho_\air\,A_\sail=\rho_\water\,A_\keel$ you get the fastest possible cross-wind sailing speed.
+Is that a result, and is it known?
+
+HOGG maybe take the limit of infinite lift ratios and show that you get absurd results.
 
@@
 
-HOGG: Introduce drag ratios or lift ratios for the air and the water.
-And then show how the hourglass plots depend on these ratios.
-such as: Relative area of sail and keel.
-Relative area of hull and keel.
-Etc.
+HOGG: make sure we consistently use the term "lift ratio" for the air and the water, and that this is a real term.
 

### 2021-08-02 paper/main.tex

@@
 % - Kilian Walsh
-% - Hyung Don Ryoo
 % - Sanjoy Mahajan
@@
 % ------------
-% - Make figure comparing different boats.
-% - Change citations to Tao to make it clear that the model is wrong!!
+% - Switch to calling the sail and keel orientations "effective orientations" not just orientations. And say why.
+% - make sure I clearly introduce the concept of lift ratio and stick with it.
+% - Change citations to Tao to make it clear that the Tao model is (probably) wrong!!
 % - Make sure I cite {symmetry} appropriately in various places. Probably also point out that it is a fun longer read.
@@
 We can beat that too.
+% - Respond to comments by Hyung Don Ryoo, in email to Hogg, 2021-07-15.
 
@@
     The solutions corresponding to sailing in steady wind correspond to boat velocities that deliver vanishing total force, given the angles of the sail and keel.
-    We use this model to derive simple rules for sailing that are not far off from standard sailing practice, for sailing in different directions (including upwind).
+    We use this model to derive simple properties of sailing that are not far off from standard folklore and measurements.
     One consequence of this model is that a sailboat does not move (relative to the water) precisely in the direction that its keel points (the direction of the prow); it moves slightly downwind of that.
-    Another consequence is that it is possible for an extremely anisotropic (extremely well designed) boat to sail downwind (relative to the water) \emph{faster than the wind}, on a tacking course.
-    Interestingly, nothing conceptual about the mechanical model or discussion depend on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; but the important thing is that the model \emph{does} depend on what's assumed about the angular re-direction of the wind momentum flux.
+    Another consequence is that it is possible for an extremely well designed boat (very good lift ratios) to sail downwind (relative to the water) \emph{faster than the wind}, on a tacking course.
+    Another is that cross-wind sailing is fastest when the effective area of the sail is larger than the effective area of the keel by the ratio of the density of water to the density of air.
+    Nothing conceptual about this model depends on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; the important thing is that the model \emph{does} depend on angular re-direction of fluid momentum flux.
 \end{abstract}

### 2021-08-29 paper/main.tex
todo items
Ryoo

@@
 % ------------
+% - Add to the abstract the point that the sailboat is defined by 3 dimensionless numbers: The sail-to-keel ratio, and two lift ratios.
+% - confirm that "lift ratio" is the correct terminology.
 % - Switch to calling the sail and keel orientations "effective orientations" not just orientations. And say why.
@@
 \paragraph{Acknowledgements:}
-It is a pleasure to thank Will Kinney (Buffalo), Hans-Walter Rix (MPIA), and Soledad Villar (JHU) for valuable discussions.
+It is a pleasure to thank Will Kinney (Buffalo), Hans-Walter Rix (MPIA), Hyung-Don Ryoo (NYU), and Soledad Villar (JHU) for valuable discussions and comments.
 

### 2021-11-14 paper/main.tex
Updates from Overleaf

--- 2021-08-02 paper/main.tex
@@
     Another is that cross-wind sailing is fastest when the effective area of the sail is larger than the effective area of the keel by the ratio of the density of water to the density of air.
-    Nothing conceptual about this model depends on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; the important thing is that the model \emph{does} depend on angular re-direction of fluid momentum flux.
+    Nothing conceptual about this model depends on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; the important thing is that the model \emph{does} depend on angular re-direction of fluid momentum flux; sail curvature can help with this.
 \end{abstract}

### 2021-11-20 paper/main.tex
small things

@@
 The boat is powered by the wind, and yet it can travel (possibly with a tacking course) in any direction.
-Of course it's not consistent with Galilean relativity to say that the boat is powered by the wind:
+Of course it's not consistent with Galilean relativity\footnote{We refer to Galilean relativity repeatedly in this \documentname.
+Of course Galilean relativity is just the low-velocity approximation to special relativity \cite{sr}.
+But we are not going to consider air--water velocity differences that are an appreciable fraction of the speed of light (except maybe in the discussion in \secref{sec:discussion}).} to say that the boat is powered by the wind:
 There is no absolute velocity, and there is a valid rest frame in which the wind velocity is zero.
@@
 They are mainly aimed at sailors who want to understand better their boats and how to design them and sail them.
-This paper is aimed at a different audience.
 This paper is aimed at physicists and students of physics interested in a simple, implementable, physically justifiable model of sailing.
@@
 This over-simplification might be frustrating to sailing experts, or fluid dynamicists.
-But our goal is to find the simplest possible, dimensionally correct, Galilean-symmetric rules for a boat that can sail non-trivially.
+But our goal is to find the simplest possible, dimensionally correct, Galilean-relativistic rules for a boat that can sail non-trivially.
 You might say that this is an astrophysics-quality model of a sailboat!
@@
 In addition to obeying this symmetry, our model is consistent with dimensional analysis and is equivariant to rotations and reflections.
-We also make use of good, coordinate-free expressions, in the style of (for example) \cite{kusse}.
+We also make use of good, coordinate-free expressions.
 
@@
 That momentum transport---the (vector) change in momentum in the air per unit time---is the (vector) force on the air from the sail.
-Half of the explanation for how a sailboat sails uses the equal-and-opposite reaction to this force.
+Half of the explanation for how a sailboat sails uses the equal-and-opposite reaction to this force (the other half of the explanation will come from the keel).
 
@@
 In this paper we are not arguing that ram pressure on flat sails is a complete explanation of a sailboat!
-We are arguing that the sailboat can be modeled as having two important surfaces---the sail and the keel---one of which deviates the flow of air in the rest-frame of the boat, and the other deviates the flow of water.
+We are arguing that the sailboat can be modeled as having two important surfaces---the sail and the keel---one of which deviates the flow of air, and the other deviates the flow of water.
 The angles of the sail and keel with respect to the air--boat and water-boat velocity vectors set the angular deviations of the two fluids, and therefore both the magnitudes and the directions of the sail and keel forces.
 The directions of the forces can be manipulated to create amazing boat dynamics and kinematics.
-The point of this \documentname{} is that much of the amazing dynamics can be observed in an amazingly simple model for the sailboat.
+The point of this \documentname{} is that much of the amazing dynamics can be understood in an amazingly simple model for the sailboat.
 So while we use the phrases ``flat sail'' and ``ram-pressure'' in what follows to describe our model, we really mean ``simple momentum transport''.
@@
 
-We are going to work in 2D, with velocities $\vec{v}$ and forces $\vec{F}$ that are column vectors (or $2\times 1$ matrices).
+We are going to work in 2D (you can think of the two dimensions as the surface of a map of a local patch of the Earth, if you like), with velocities $\vec{v}$ and forces $\vec{F}$ that are column vectors (or $2\times 1$ matrices).
 To stay coordinate-free, we are going to talk about angles as orientations of unit vectors like $\uvec_{\perp\sail}$ and $\uvec_{\parallel\sail}$ (which will be defined below).
@@
 where the sail is replaced by the keel everywhere, and $A_{\below}$ is an effective area related to the area of the hull below the water surface.
-HOGG NEED WE SAY MORE ABOUT THESE? PROBABLY YES THERE ARE COMMENTS TO BE MADE.
+HOGG NEED WE SAY MORE ABOUT THESE? PROBABLY YES THERE ARE COMMENTS TO BE MADE. LIKE FOR INSTANCE LIFT RATIOS.
 
@@
 
-HOGG: Take some time to bash the Bernoulli bullshit.
+HOGG: Take some time to bash the Bernoulli bullshit. Note also that the different-path-length part of that argument is explicitly not true for a sail!
 
@@
 Most thinking about solar sails and reflected-photon propulsion is thinking in terms of sailing ``down wind'' of the nearby star, but with multiple sources of illumination, in principle one might be able to make a photon-pressure vehicle that can sail or tack in arbitrary directions, given illumination from two or more stars.
+This might involve a relativistic generalization of the above.
 HOGG: Someone must have thought of all this before!

### 2021-11-23 paper/main.tex
figure caption, notes to self

@@
 We can beat that.
-Have I seen a "never faster downwind than the wind" or than 0.5 times the wind claim?
+Have I seen a "never faster downwind than the wind" (as in Kimball) or than 0.5 times the wind claim?
 We can beat that too.
 % - Respond to comments by Hyung Don Ryoo, in email to Hogg, 2021-07-15.
+% - Make sure the spirit of the bibliographic comments get pulled into the main text.
+% - Somewhere comment that the wind is a function of height, which means that another approximation has to do with what particular mean wind speed we are considering.
 
@@
 You might say that this is an astrophysics-quality model of a sailboat!
-The only prior work we know in this space is \cite{tao}, but that model is (in our view) overly simplistic.
-For example, it doesn't lead to steady-state sailing because the drag forces (in that model) vanish (HOGG CHECK).
+The only prior work we know in this space is \cite{tao}, but that model is (in our view) overly simplistic, and maybe dimensionally incorrect (it appears to make forces proportional to velocities, instead of squared velocities).
 In principle also \cite{pos} delivers similar equations to what we derive below, based on similar considerations and similar approximations, but their goals are not really for their reader to build an implementable model.
@@
 \begin{align}\label{eq:flat}
-    \vec{F}_\flat &= \eta\,\rho\,A\,(\uvec_\perp^\top\vec{v})^2\,\uvec_\perp ~,
+    \vec{F}_\flat &= \eta\,\rho\,A\,(\uvec_\perp\cdot\vec{v})^2\,\uvec_\perp ~,
 \end{align}
@@
 relative velocity between the fluid and the surface,
-and $\uvec_\perp^\top\vec{v}$ is the scalar inner product (dot product) of the two vectors.
-MK - I'M NOT A FAN OF THIS NOTATION FOR INNER PRODUCT, ESPECIALLY WITH UPPER AND LOWER Ts.
+and $\uvec_\perp\cdot\vec{v}$ is the scalar inner product (dot product) of the two vectors.
 
@@
 \begin{align}\label{eq:flat2}
-    \vec{F}_\flat &= \eta\,\rho\,A\,|\uvec_\perp^\top\vec{v}|\,(\uvec_\perp^\top\vec{v})\,\uvec_\perp ~;
+    \vec{F}_\flat &= \eta\,\rho\,A\,|\uvec_\perp\cdot\vec{v}|\,(\uvec_\perp\cdot\vec{v})\,\uvec_\perp ~;
 \end{align}
@@
   \\
-  \vec{F}_{\sail-\air} & = \rho_\air\,A_\sail\,(\uvec_{\perp\sail}^\top\,\vec{\Delta v}_\air)^2\,\uvec_{\perp\sail} \label{eq:Fsailair}
+  \vec{F}_{\sail-\air} & = \rho_\air\,A_\sail\,(\uvec_{\perp\sail}\cdot\vec{\Delta v}_\air)^2\,\uvec_{\perp\sail} \label{eq:Fsailair}
   \\
@@
   \\
-  \vec{F}_{\keel-\water} & = \rho_\water\,A_\keel\,(\uvec_{\perp\keel}^\top\,\vec{\Delta v}_\water)^2\,\uvec_{\perp\keel}
+  \vec{F}_{\keel-\water} & = \rho_\water\,A_\keel\,(\uvec_{\perp\keel}\cdot\vec{\Delta v}_\water)^2\,\uvec_{\perp\keel}
   \\
@@
 where the sail is replaced by the keel everywhere, and $A_{\below}$ is an effective area related to the area of the hull below the water surface.
-HOGG NEED WE SAY MORE ABOUT THESE? PROBABLY YES THERE ARE COMMENTS TO BE MADE. LIKE FOR INSTANCE LIFT RATIOS.
+
+In some of what follows, we will make use of dimensionless ratios of effective areas.
+For example, the ratio of sail to keel $A_\sail/A_\keel$ will affect the ratio of upwind to downwind sailing speeds (and, surprisingly, affect cross-wind sailing).
+For another, the ratios of sail and keel to the relevant parts of the hull $A_\sail/A_\above$ and $A_\keel/A_\below$ act like ``lift ratios'' in the study of flying:
+They control the ratio of useful sailing forces to drag.
+We will refer to these---na\"ively---as lift ratios in what follows.
 
@@
   \\
-  \vec{F}_{\parallel-\air} & = \rho_\air\,A_{\sail\parallel}\,(\uvec_{\parallel\sail}^\top\,\vec{\Delta v}_\air)^2\,\uvec_{\parallel\sail}
+  \vec{F}_{\parallel-\air} & = \rho_\air\,A_{\sail\parallel}\,(\uvec_{\parallel\sail}\cdot\vec{\Delta v}_\air)^2\,\uvec_{\parallel\sail}
    ~,
@@
 \begin{align}\label{eq:good}
-    \uvec_{\perp\sail}^\good &\leftarrow \argmax_{\uvec_{\perp\sail}} \left[\uvec_{\parallel\keel}^\top\,(\vboat-\vwater)\right] ~,
+    \uvec_{\perp\sail}^\good &\leftarrow \argmax_{\uvec_{\perp\sail}} \left[\uvec_{\parallel\keel}\cdot(\vboat-\vwater)\right] ~,
 \end{align}
@@
 A summary of all possible boat--water velocity vectors possible, for GoodSailing(tm) at many different keel orientations, is shown in \figref{fig:hourgood}.
-This latter figure is the HOGG WHAT DIAGRAM that is standard in all books on the subject of sailing (HOGG GIVE REFERENCES).
+This latter figure is sometimes called a ``speed diagram'' \cite{pos}; it is standard in all books on the subject of sailing (HOGG GIVE REFERENCES and figure numbers).
 \begin{figure}[t!]
@@
   As is visible in \figref{fig:good}, when the keel orientation is too close to directly upwind, the boat goes downwind instead; there are no \emph{directly} upwind options.
-This figure is directly comparable to [HOGG WHAT FIGURES IN THE BOOKS ABOUT SAILING?].\label{fig:hourgood}}
+This figure is sometimes called a ``speed curve'' \cite{pos}.
+It is directly comparable to [HOGG WHAT FIGURES IN THE BOOKS ABOUT SAILING?].\label{fig:hourgood}}
   \figurerule
 \end{figure}
-
-HOGG: Figure showing direction cosines of sail relative to keel/boat orientation?
 
@@
 
-HOGG IN THIS SECTION EXPLICITLY DEFINE ``BETTER SAILING''.
-
 A type-A sailor---or maybe a competitive sailor---is trying to get from a current position to a destination as quickly as possible\footnote{%
@@
 \begin{align}\label{eq:better}
-    \uvec_{\perp\sail}^\better,\uvec_{\perp\keel}^\better &\leftarrow \argmax_{\uvec_{\perp\sail},\uvec_{\perp\keel}} \left[\uvec_\destination^\top\,(\vboat-\vdest)\right] ~,
+    \uvec_{\perp\sail}^\better,\uvec_{\perp\keel}^\better &\leftarrow \argmax_{\uvec_{\perp\sail},\uvec_{\perp\keel}} \left[\uvec_\destination\cdot(\vboat-\vdest)\right] ~,
 \end{align}
@@
   For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:better}.
-  Compare this to \figref{fig:hourglass} and also to \figref{fig:hourgood}.\label{fig:hourbetter}}
+  Compare this speed curve to \figref{fig:hourglass} and also to \figref{fig:hourgood}.\label{fig:hourbetter}}
   \figurerule
@@
 \begin{align}
-  A_\sail &= 1400\,\mm \\
+  A_\sail &= 1400\,\mm \label{eq:boat1}\\
   A_\keel &= 2\,\mm \\
   A_\above &= 5.6\,\mm \\
-  A_\below &= 0.008\,\mm ~.
+  A_\below &= 0.008\,\mm \label{eq:boat4} ~.
 \end{align}
@@
 This boat also has very good drag ratios or lift ratios:
-The ratio of effective area of sail to hull-above is 250, as is the ratio of effective area of keel to hull-below.
-All these ratios were set not by analyzing any real boats, but by making the velocity curve shown in \figref{fig:hourgood} look reasonable, or accord with our intuitions and expectations.
-HOGG: NEED A BETTER NAME THAN VELOCITY CURVE; WHAT IS THIS DIAGRAM CALLED?
+The sail lift ratio (the ratio of effective area of sail to hull-above) is 250, as is the keel lift ratio (the ratio of effective area of keel to hull-below).
+All these ratios were set not by analyzing any real boats, but by making the speed curve shown in \figref{fig:hourgood} look reasonable, or accord with our intuitions and expectations.
 Recall, for one, that our boat is a brutal approximation!
@@
   ~\hfill\includegraphics[width=0.5\textwidth]{design_E.pdf}\hfill~
-  \caption{Each panel is analagous to \figref{fig:hourgood} but in each case we have varied one of the dimensionless ratios....HOGG....for varying values of the three ratios $A_\sail/A_\keel$, $A_\sail/A_\above$, and $A_\keel/A_\below$.\label{fig:design}}
+  \caption{Each panel is analagous to \figref{fig:hourgood} but in each case we have varied one of the boat's dimensionless ratios.
+In each panel, the grey points show the same speed curve as that plotted in \figref{fig:hourgood}, which was computed for the boat described by the parameters given in \eqref{eq:boat1} through \eqref{eq:boat4}, and for a wind speed (relative to the water) of $8\,\mps$.
+  In the top-left panel, the small and large black points show variations as we vary the sail lift ratio $A_\sail/A_\above$ from HOGG (small dots) to HOGG (large dots).
+  In the top-right panel, the small and large black dots show variations as we vary the keel lift ratio $A_\keel/A_\below$ from HOGG (small dots) to HOGG (large dots).
+  In the bottom-left panel, the small and large black dots show the variations as we vary the sail-to-keel ratio $A_\sail/A_\keel$ from HOGG (small dots) to HOGG (large dots).\label{fig:design}}
   \figurerule
@@
 
-HOGG: make sure we consistently use the term "lift ratio" for the air and the water, and that this is a real term.
+HOGG: make sure we consistently use the term ``lift ratio'' for the air and the water, and that this is a real term.
 
@@
 
+HOGG: Speaking of misconceptions: These results probably do indeed show that you can travel directly downwind faster than the wind. That maybe resolves some folk conflicts or arguments?
+
 HOGG: This model implies, indirectly, that keels should be curved! And they are, in high-end sailing (HOGG cite?)

### 2021-12-11 paper/main.tex
discussion and notes to self

@@
 
-Finally, although upwind travel is slow going, with lots of tacking (HOGG DO WE NEED TO DEFINE THIS?) required, if you are situated \emph{on the boat}, upwind travel \emph{feels very fast}.
+Finally, although upwind travel is slow going, with lots of tacking (HOGG DO WE NEED TO DEFINE THIS? YES WE DO) required, if you are situated \emph{on the boat}, upwind travel \emph{feels very fast}.
 Why?
@@
 The sail lift ratio (the ratio of effective area of sail to hull-above) is 250, as is the keel lift ratio (the ratio of effective area of keel to hull-below).
-All these ratios were set not by analyzing any real boats, but by making the speed curve shown in \figref{fig:hourgood} look reasonable, or accord with our intuitions and expectations.
-Recall, for one, that our boat is a brutal approximation!
+All these ratios were set not by analyzing any real boats, but by making the speed curve shown in \figref{fig:hourgood} look reasonable, or accord with expectations set by the books on the physics of sailing HOGG CITE THESE.
+
+In thinking about these areas, recall, for one, that our boat is a brutal approximation!
 For two, these areas are \emph{effective areas} within this brutal approximation, not measured areas.
@@
 In this sense, maybe our model is more appropriate for an ice yacht than a sailboat?
+HOGG BUT SEE THE AMERICA'S CUP BOATS!
+
+HOGG NOW ADDRESS THE EXCELLENT LIFT RATIOS.
 
 As we change the ratios of the effective areas, the shapes of the velocity curves change.
-In \figref{fig:design} we show the effect on the boat velocity curves of changing the ratios $A_\sail/A_\keel$, $A_\sail/A_\above$, and $A_\keel/A_\below$....HOGG
+In \figref{fig:design} we show the effect on the boat velocity curves of changing the ratios $A_\sail/A_\keel$, $A_\sail/A_\above$ (the sail lift ratio), and $A_\keel/A_\below$ (the keel lift ratio)....HOGG
 \begin{figure}[t!]
@@
 
-HOGG: Comment on what we seen in \figref{fig:design}.
+HOGG: Comment on what we see in \figref{fig:design}.
 HOGG: Note that if you set $\rho_\air\,A_\sail=\rho_\water\,A_\keel$ you get the fastest possible cross-wind sailing speed.
@@
 
-HOGG: What did we find? What's important about that?
+This \documentname{} presented what we believe to be the simplest model of sailing that is both physically realizable and also reproduces most of the salient features of the behavior of sailboats.
+The model is simple to state, makes use of nothing more than undergraduate mechanics, conforms to Galilean relativity and coordinate freedom, and is extremely easy to implement numerically.
+It makes explicit that the boat is powered by the relative velocity of the air and water.
+It demonstrates that there is nothing physically strange about being able to travel, by sailboat, upwind, or downwind faster than the wind, in both cases with some tacking.
+Along the way, we found that there are best sail-to-keel area ratios for cross-wind sailing, and that keels ought to be curved like sails; both of these ``predictions'' seem to have evidence in the best-in-class sailing technologies of today.
 
-HOGG: Galilean symmetry, coordinate freedom?
+The model makes brutal assumptions.
+In some sense, this is the ``spherical cow'' model of a sailboat.
+It literally assumed (incorrectly) that the hull of the boat is spatially isotropic (cylindrically symmetric).
+It assumed that viscosities are negligible, and that the effects of turbulence are all perfectly captured by a trivial ram-pressure expression (force proportional to an area times velocity squared).
+The model assumed that the drag coefficients $\eta$ do not depend on the angles of the sail or keel, nor on the relative speeds of air, water, and boat.
+All these assumptions are wrong in detail.
+They could be fixed and accuracy could be improved, but at the expense of adding many new parameters and constants, which distract from the principal story:
+Sailing is a very simple phenomenon from a fundamental-physics perspective.
 
-HOGG: What were our assumptions?
-What are the myriad ways in which those assumptions must be wrong?
-For example, boat has an asymmetry in the wind too, so really the boat must also matter.
-Viscosity; why can we ignore it?
-But turbulence; we can't ignore that!
-Constant $\eta$ is also wrong.
-
-HOGG: iso-planar vs parallelepipeds...
-
-HOGG: Take some time to bash the Bernoulli bullshit. Note also that the different-path-length part of that argument is explicitly not true for a sail!
+HOGG: This is a momentum-transport model! NOW Take some time to bash the Bernoulli bullshit. Note also that the different-path-length part of that argument is explicitly not true for a sail!
 

### 2021-12-11 paper/main.tex
relativistic and notes

@@
 % - Somewhere comment that the wind is a function of height, which means that another approximation has to do with what particular mean wind speed we are considering.
+% - Are we clear that we also find that you can sail faster than the wind, in general, like cross-wind? Emphasize that when we first figure it out.
 
@@
 
-Na\"ively in the particle-collision model the prefactor $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity\footnote{%
+Na\"ively in the particle-collision model the prefactor $\eta$ is 2, but in the world of real fluids it depends on the shape of the surface, the velocity, and (if things are tiny) the viscosity.\footnote{%
 Viscosities only matter for very small objects, much smaller than sailboats.
@@
 That is, viscosity plays no first-order role in sailing.
-In detail it shapes the turbulence in the wakes coming off the sail and keel edges, but that is a tiny detail from perspective of our brutal model.}.
+In detail it shapes the turbulence in the wakes coming off the sail and keel edges, but that is a tiny detail from perspective of our brutal model.}
 We're going to treat $\eta$ as an uninteresting constant; in particular we are going to assume that it doesn't depend on velocity, such that it can be combined with the area $A$ (to make an effective area) in everything that follows.
@@
 If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel.
-This is a root-finding problem that (probably) has no closed form\footnote{%
+This is a root-finding problem that (probably) has no closed form.\footnote{%
 It might be interesting to note that the flat-surface force \eqref{eq:flat} is a vector-valued quadratic form, so if one constructed the sailboat model completely from flat surfaces, the force law, in this approximation would be a quadratic form.
-That might lead to a closed-form solution for the equation \eqref{eq:sailing} of steady sailing.}.
+That might lead to a closed-form solution for the equation \eqref{eq:sailing} of steady sailing.}
 We solve equation~\eqref{eq:sailing} numerically in the code associated with this \documentname\footnote{%
@@
 
-A type-A sailor---or maybe a competitive sailor---is trying to get from a current position to a destination as quickly as possible\footnote{%
-We don't particularly endorse this attitude towards sailing. The journey \emph{is} the destination.}.
+A type-A sailor---or maybe a competitive sailor---is trying to get from a current position to a destination as quickly as possible.\footnote{%
+We don't particularly endorse this attitude towards sailing. The journey \emph{is} the destination.}
 Our view is that this sailor should start by specifying the unit vector $\uvec_\destination$ that points from the current position of the boat towards the destination of the boat.
@@
 
-In all of the experiments and figures given above, we made use of a model boat with the following effective areas:
+In all of the experiments and figures given above, we made use of a model boat with the following effective areas: HOGG THESE ARE ABSURD. FIX AND RERUN.
 \begin{align}
@@
 HOGG: 
-A true boat has sails that are globally curved\footnote{%
+A true boat has sails that are globally curved,\footnote{%
 In the local-curvature or Riemann-curvature senses, sails are often not (very) curved; after all, they are often made from flat pieces of fabric.
@@
 It is in this colloquial sense that we are using the words ``curved'' and ``flat''.
-All that said, the most sophisticated sails in the world are also curved in the local sense \cite{sails}.},
+All that said, the most sophisticated sails in the world are also curved in the local sense \cite{sails}.}
 and forces that are different in detail from these pure ram-pressure forces.
@@
 It makes explicit that the boat is powered by the relative velocity of the air and water.
-It demonstrates that there is nothing physically strange about being able to travel, by sailboat, upwind, or downwind faster than the wind, in both cases with some tacking.
-Along the way, we found that there are best sail-to-keel area ratios for cross-wind sailing, and that keels ought to be curved like sails; both of these ``predictions'' seem to have evidence in the best-in-class sailing technologies of today.
+It demonstrates that it is easy for a sailboat to travel faster than the wind (or, to be specific, to travel relative to the water faster than the magnitude of the relative wind-water velocity).
+It demonstrates that there is nothing physically strange about being able to travel, by sailboat, upwind, or downwind faster than the wind, in both cases by taking a tacking course.
+Along the way, we found that the best sail-to-keel effective-area ratio for cross-wind sailing is the ratio of the density of water to the density of air; and we found that keels ought to be curved like sails.
+Both of these ``predictions'' seem to have evidence in the best-in-class sailing technologies of today.
 
-The model makes brutal assumptions.
+This model makes brutal assumptions and approximations.
 In some sense, this is the ``spherical cow'' model of a sailboat.
@@
 The model assumed that the drag coefficients $\eta$ do not depend on the angles of the sail or keel, nor on the relative speeds of air, water, and boat.
-All these assumptions are wrong in detail.
+The model ignored that the wind is gusty, and that wind speeds (and water speeds) vary with height (and depth).
+All these approximations are wrong in detail.
 They could be fixed and accuracy could be improved, but at the expense of adding many new parameters and constants, which distract from the principal story:
@@
 
-HOGG: Because we have such a simple model, there might be a version for a star ship that sails by photon pressure.
-Most thinking about solar sails and reflected-photon propulsion is thinking in terms of sailing ``down wind'' of the nearby star, but with multiple sources of illumination, in principle one might be able to make a photon-pressure vehicle that can sail or tack in arbitrary directions, given illumination from two or more stars.
+We expressed pride that our model is explicitly Galilean-relativistic.
+That begs the question: What about special relativity?
+Of course our model is consistent with special relativity; Galilean relativity is the low-velocity limit, when all relative velocities have magnitudes way less than the speed of light.
+Now what about \emph{relativistic sailing?}
+Because we have such a simple model, we conjecture that there might be a version for a star ship that sails by photon pressure (or on stellar winds or supernovae ejecta or other quasi-relativistic flows).
+Most contemporary thinking about solar sails and reflected-photon propulsion is thinking in terms of sailing purely ``down wind'' of a single, nearby star, which dominates the photon pressure on any active surface.
+But the Galaxy is filled with stars!
+With multiple sources of illumination, in principle one might be able to make a photon-pressure vehicle that can sail or tack in arbitrary directions, given illumination from multiple stars.
 This might involve a relativistic generalization of the above.
-HOGG: Someone must have thought of all this before!
+HOGG: Someone must have thought of all this before! CITE THEM.
 

### 2021-12-16 paper/main.tex
content

@@
 These are also the physical components of a ram-pressure force.
-Indeed, the most basic derivation of the ram-pressure force is in the microscopic model of an ideal gas, in which particles bounce elastically off the surface, changing their momentum.
+Indeed, the most basic derivation of the ram-pressure force is in the microscopic model of an ideal gas, in which particles bounce elastically off the surface, changing their momenta.
 HOGG PUT A DIAGRAM HERE AND EQUATIONS.
@@
 It demonstrates that there is nothing physically strange about being able to travel, by sailboat, upwind, or downwind faster than the wind, in both cases by taking a tacking course.
-Along the way, we found that the best sail-to-keel effective-area ratio for cross-wind sailing is the ratio of the density of water to the density of air; and we found that keels ought to be curved like sails.
-Both of these ``predictions'' seem to have evidence in the best-in-class sailing technologies of today.
+Along the way, we found that the best sail-to-keel effective-area ratio for cross-wind sailing is the ratio of the density of water to the density of air; and we found that keels ought to be curved like sails (HOGG DO WE EXPLICITLY DISCUSS THIS ABOVE?
+IF NOT, WE SHOULD, WHERE WE DISCUSS THE SYMMETRY BETWEEN SAIL AND KEEL, OR WHERE WE DISCUSS THE CURVATURE OF THE SAILS.)
+For both of these ``predictions'' there seems to be evidence in the designs of the best-in-class sailing vessels today.
 
@@
 The model assumed that the drag coefficients $\eta$ do not depend on the angles of the sail or keel, nor on the relative speeds of air, water, and boat.
-The model ignored that the wind is gusty, and that wind speeds (and water speeds) vary with height (and depth).
+The model ignored that the wind is gusty, and that wind speeds (and water speeds), in the real world, vary with height (and depth).
 All these approximations are wrong in detail.
@@
 
-HOGG: Even BetterSailing(tm) is not the best you can do, because you have to do complete successful path planning, and coming about costs time.
-So you will sail beyond the limits of BetterSailing(tm) if you do this correctly.
-This is out of scope for this \documentname.
+We defined GoodSailing(tm), which is an optimal setting of the orientation of the sail, given the orientation of the boat.
+And we defined BetterSailing(tm), which is an optimal setting of both the orientation of the boat and the orientation of the sail, given the direction to the destination.
+But even this is not the best you can do, because optimal sailing (from a time-to-destination perspective) involves complete path planning, constrained to start at the origin and end at the destination.
+For destinations that are close to directly downwind or directly upwind of the origin, this complete path planning will involve tacking the boat.
+Each tack involves a ``coming about'' or a rotation of the boat to put the wind on the other beam (HOGG CORRECT LANGUAGE HERE?).
+Each coming about costs time.
+So optimal path planning must involve that cost, factored against the requirement (if you delay coming about) that you will sail beyond the limits of what BetterSailing(tm) recommends, given your destination.
+That is, the global optimal path will not be locally optimal at all points.
+A full discussion of this path planning is out of scope for this \documentname, but clearly interesting from an engineering perspective.
 

### 2021-12-23 paper/main.tex
Updates from Overleaf

@@
 % - Are we clear that we also find that you can sail faster than the wind, in general, like cross-wind? Emphasize that when we first figure it out.
+% - Need to pull in most recent figures; parameters have changed.
 
@@
     We present a simplified model for the forces on a sailboat.
-    Our model is that there is a ram-pressure force on the sail, with a magnitude and direction that depend on the alignment of the relative air--boat velocity with the sail orientation according to an anisotropic drag tensor.
-    Similarly, there is a ram-pressure force on the keel, depending on the relative water--boat velocity.
+    Our model is that there is a ram-pressure force on the sail, with a magnitude and direction that depend on the relative air--boat velocity and the sail orientation.
+    Similarly, there is a ram-pressure force on the keel, depending on the relative water--boat velocity and the keel orientation.
     The resulting expression for the total force on the boat is coordinate-free and consistent with Galilean relativity.
     The solutions corresponding to sailing in steady wind correspond to boat velocities that deliver vanishing total force, given the angles of the sail and keel.
-    We use this model to derive simple properties of sailing that are not far off from standard folklore and measurements.
+    We use this simple model to derive simple properties of sailing that are not far off from standard folklore and measurements.
     One consequence of this model is that a sailboat does not move (relative to the water) precisely in the direction that its keel points (the direction of the prow); it moves slightly downwind of that.
@@
 
-HOGG: Define direction cosines with the wind, and with boat parts, using vectors. ADD THOSE direction cosines to the figure panels.
-
 In \figref{fig:steady} we show examples of steady sailing, for a wide range of orientations of the keel and sail.
@@
 
-In all of the experiments and figures given above, we made use of a model boat with the following effective areas: HOGG THESE ARE ABSURD. FIX AND RERUN.
+In all of the experiments and figures given above, we made use of a model boat with the following effective areas: HOGG SAY: THESE ARE NOW CONSISTENT WITH THE NOTEBOOK, BUT NOT WITH THE FIGURES.
+NEED TO PULL IN NEW UPDATED FIGURES.
 \begin{align}
-  A_\sail &= 1400\,\mm \label{eq:boat1}\\
-  A_\keel &= 2\,\mm \\
-  A_\above &= 5.6\,\mm \\
-  A_\below &= 0.008\,\mm \label{eq:boat4} ~.
+  A_\sail / A_\keel &= 700 \quad\mbox{(sail-to-keel ratio)}\label{eq:boat1}\\
+  A_\sail / A_\above &= 200 \quad\mbox{(sail lift ratio)}\\
+  A_\keel / A_\below &= 200 \quad\mbox{(keel lift ratio)}\label{eq:boat4}
 \end{align}
@@
 In each panel, the grey points show the same speed curve as that plotted in \figref{fig:hourgood}, which was computed for the boat described by the parameters given in \eqref{eq:boat1} through \eqref{eq:boat4}, and for a wind speed (relative to the water) of $8\,\mps$.
-  In the top-left panel, the small and large black points show variations as we vary the sail lift ratio $A_\sail/A_\above$ from HOGG (small dots) to HOGG (large dots).
-  In the top-right panel, the small and large black dots show variations as we vary the keel lift ratio $A_\keel/A_\below$ from HOGG (small dots) to HOGG (large dots).
-  In the bottom-left panel, the small and large black dots show the variations as we vary the sail-to-keel ratio $A_\sail/A_\keel$ from HOGG (small dots) to HOGG (large dots).\label{fig:design}}
+  In the top-left panel, the small and large black points show variations as we vary the sail lift ratio $A_\sail/A_\above$ from 40 (small dots) to 200 (grey dots) to 1000 (large dots).
+  In the top-right panel, the small and large black dots show variations as we vary the keel lift ratio $A_\keel/A_\below$ from 40 (small dots) to 200 (grey dots) to 1000 (large dots).
+  In the bottom-left panel, the small and large black dots show the variations as we vary the sail-to-keel ratio $A_\sail/A_\keel$ from 140 (small dots) to 700 (grey dots) to 3500 (large dots).\label{fig:design}}
   \figurerule

### 2021-12-23 paper/main.tex
remove some HOGGs from text; update boat properties

@@
 % - Are we clear that we also find that you can sail faster than the wind, in general, like cross-wind? Emphasize that when we first figure it out.
-% - Need to pull in most recent figures; parameters have changed.
+% - Note that even a small change in the lift ratio on the sail and you can't go upwind!
 
@@
 
-In all of the experiments and figures given above, we made use of a model boat with the following effective areas: HOGG SAY: THESE ARE NOW CONSISTENT WITH THE NOTEBOOK, BUT NOT WITH THE FIGURES.
-NEED TO PULL IN NEW UPDATED FIGURES.
+In all of the experiments and figures given above, we made use of a model boat with the following effective-area ratios:
 \begin{align}

### 2021-12-23 paper/main.tex

@@
 % - Note that even a small change in the lift ratio on the sail and you can't go upwind!
+% - Move comments from the annotated bibliography up into the text somewhere and drop the annotations.
 
@@
 \newcommand{\figref}[1]{Figure~\ref{#1}}
+\newcommand{\code}[1]{\texttt{#1}}
 
@@
 This model is a sandbox for experimenting with wind-powered travel and wind-powered machines, which are, in turn, key concepts in sustainable technology.
-Our model of sailing is also something that can be implemented by students new to physics, in extremely simple code (an example of which is released in conjunction with this \documentname\footnote{All code used in this \documentname{} is available under a HOGG WHAT LICENSE?
-at \url{https://github.com/davidwhogg/Sailing}.}).
+Our model of sailing is also something that can be implemented by students new to physics, in extremely simple code (an example of which is released in conjunction with this \documentname\footnote{All code used in this \documentname{} is available under the MIT License at \url{https://github.com/davidwhogg/Sailing}.}).
 
@@
   \includegraphics[width=0.5\textwidth]{design_C.pdf}\\%
-  ~\hfill\includegraphics[width=0.5\textwidth]{design_E.pdf}\hfill~
+  \rule{0.25\textwidth}{0pt}\includegraphics[width=0.5\textwidth]{design_E.pdf}
   \caption{Each panel is analagous to \figref{fig:hourgood} but in each case we have varied one of the boat's dimensionless ratios.
@@
   In the top-right panel, the small and large black dots show variations as we vary the keel lift ratio $A_\keel/A_\below$ from 40 (small dots) to 200 (grey dots) to 1000 (large dots).
-  In the bottom-left panel, the small and large black dots show the variations as we vary the sail-to-keel ratio $A_\sail/A_\keel$ from 140 (small dots) to 700 (grey dots) to 3500 (large dots).\label{fig:design}}
+  In the bottom panel, the small and large black dots show the variations as we vary the sail-to-keel ratio $A_\sail/A_\keel$ from 140 (small dots) to 700 (grey dots) to 3500 (large dots).\label{fig:design}}
   \figurerule
@@
 
-HOGG: Newton's method (CITE), invert, solve, etc.
-Note that in practice, Newton's method works amazingly well.
-For Newton's method we make use of the derivatives---HOGG GIVE HERE THE DERIVATIVES of the ram-pressure expressions---which are order-2 tensors.
+All the code used in this \documentname{} is available open-source under the \textsl{MIT License} at \url{https://github.com/davidwhogg/Sailing}.
+Here we review a few of the implementation details.
+
+Steady sailing \eqref{eq:sailing}, being defined by the solution to a nonlinear equation, can be found by Newton's method.
+We find it by iterating to convergence the operation
+\begin{equation}
+    \vec{v} \leftarrow \vec{v} - \left[\frac{\dd \vec{F}}{\dd \vec{v}}\right]_{\vec{v}}^{-1}\cdot \vec{F}(\vec{v}) ~,
+\end{equation}
+where the derivative is a $3\times 3$ matrix, the inverse is a matrix inverse, and $\vec{F}(\vec{v})$ is given above in \eqref{eq:sailing}, which, in turn, depends on equations in \secref{sec:model}.
+In detail, the matrix inverse is not ever explicitly computed, but instead the code executes a \code{numpy.linalg.solve} (HOGG CITE NUMPY) operation, which is more numerically stable.
+Newton's method depends on an initialization.
+In practice, initializing at the mean of the water and air velocities works in every case.
+Newton's method may not be necessary here: The equations are quadratic in the velocity, so there might be a closed-form solution. KLEBAN: THOUGHTS??
+Newton's method requires the $3\times 3$ derivative tensor $\dd\vec{F}/\dd\vec{v}$.
+The code explicitly delivers this derivative, but it would be possible to use \code{jax} (HOGG CITE JAX) or any other autodifferentiation package to get the tensor automatically.
 

### 2021-12-26 paper/main.tex

@@
 
+HOGG: Note that being able to sail upwind either requires excellent sail lift ratio, or else small sail-to-keel ratio.
+
+HOGG: Various books (any books?
+find citations) say that you can't sail downwind faster than the wind.
+We don't find this, but it is certainly true for many settings of the dimensionless ratios.
+So maybe this says something about the realities of constructed, real-world boats?
+
 HOGG maybe take the limit of infinite lift ratios and show that you get absurd results.
@@
 
+HOGG: Since sails and keels have parallel roles in this, if it behooves sails to be curved, it probably also behooves keels to be curved.
+HOGG: Evidence that contemporary racing boats have keels with an active trailing edge? Also the rudder is involved.
+
 HOGG: make sure we consistently use the term ``lift ratio'' for the air and the water, and that this is a real term.
@@
 All the code used in this \documentname{} is available open-source under the \textsl{MIT License} at \url{https://github.com/davidwhogg/Sailing}.
+The code depends on \code{numpy} (HOGG CITE), \code{scipy} (HOGG CITE), and \code{matplotlib} (HOGG CITE).
 Here we review a few of the implementation details.
@@
 where the derivative is a $3\times 3$ matrix, the inverse is a matrix inverse, and $\vec{F}(\vec{v})$ is given above in \eqref{eq:sailing}, which, in turn, depends on equations in \secref{sec:model}.
-In detail, the matrix inverse is not ever explicitly computed, but instead the code executes a \code{numpy.linalg.solve} (HOGG CITE NUMPY) operation, which is more numerically stable.
+In detail, the matrix inverse is not ever explicitly computed, but instead the code executes a \code{numpy.linalg.solve} (HOGG CITE NUMPY) operation, which is more numerically stable for this operation.
 Newton's method depends on an initialization.
@@
 
-HOGG: How we implement GoodSailing(tm) and BetterSailing(tm) optimizations. It's a mess!
+GoodSailing(tm) is a global optimization defined in \eqref{eq:good}.
+We solve this global optimization by starting with a grid of 16 equally spaced (in angle) sail orientations $\uvec_{\perp\sail}$ in the range 0 to $\pi$ radians (in whatever coordinate system; it doesn't matter).
+We evaluate the objective function of \eqref{eq:good} on this grid.
+The best value for the sail orientation $\uvec_{\perp\sail}$ from that grid search is used as the initialization for a local optimization.
+In detail we use the \code{scipy} (HOGG CITE) implementation of the Nelder--Mead algorithm (HOGG CITE) but this could probably be improved by using analytic derivatives and a gradient descent method.
+
+Similarly, BetterSailing(tm) is a global optimization defined in \eqref{eq:better}.
+We solve this global optimization by starting with a uniformly spaced grid of 16 keel orientations $\uvec_{\perp\keel}$ within $\pm1/8$ of a radian of the direction to the destination.
+For each of these keel orientations, we find the GoodSailing(tm) optimization and compute the objective function of \eqref{eq:better} for the orienations $\uvec^\good_{\perp\sail},\uvec_{\perp\keel}$.
+The pair of best values of these orientations from this grid is used as the initialization for a local optimization.
+Again we use Nelder--Mead, but probably this could be improved with a gradient descent, such as conjugate gradient (CITE).
+There are a few precisely degenerate orientations for the destination direction (for example, when the destination is precisely upwind); the code currently breaks these degeneracies in an uncontrolled way.
 

### 2021-12-27 paper/main.tex
design discussion

@@
 
-HOGG: Comment on what we see in \figref{fig:design}.
-HOGG: Note that if you set $\rho_\air\,A_\sail=\rho_\water\,A_\keel$ you get the fastest possible cross-wind sailing speed.
-Is that a result, and is it known?
+There are a number of boat-design considerations visible in \figref{fig:design}.
+Sailing upwind is possible only if either the sail lift ratio is large, or the sail-to-keel ratio is small, or both.
+Since upwind sailing is critical for the success of a boat, and since obtaining very high lift ratios can be difficult, we predict that most commercial sailboats have sail-to-keel ratios smaller than the 700 of our fiducial boat.
+Sailing downwind faster than the wind is possible only if the keel lift ratio is large, or the sail-to-keel ratio is large.
+Sailing cross-wind is at its best when the sail-to-keel ratio is equal to the ratio of the density of water to the density of air.
+For context, these ratios (lift ratios and sail-to-keel ratio) are ratios of effective areas, not real areas.
 
-HOGG: Note that being able to sail upwind either requires excellent sail lift ratio, or else small sail-to-keel ratio.
+These design considerations relate to rigging (HOGG: CORRECT WORD) choices that sailors make for different journeys.
+For example, in downwind travel, often much larger or additional sails are employed (HOGG CITE).
 
-HOGG: Various books (any books?
-find citations) say that you can't sail downwind faster than the wind.
-We don't find this, but it is certainly true for many settings of the dimensionless ratios.
-So maybe this says something about the realities of constructed, real-world boats?
+Various books (HOGG find citations) say that you can't sail downwind faster than the wind.
+We don't find this, but it is certainly true that sailing downwind faster than the wind is impossible for many settings of the dimensionless ratios.
+Perhaps in real boats, with limited lift ratios, the importance of being able to sail upwind forces a small sail-t0-keel ratio and the sacrifice of failing downwind faster than the wind, which is rarely used in the normal practice of sailing.
 
@@
 
-HOGG: Since sails and keels have parallel roles in this, if it behooves sails to be curved, it probably also behooves keels to be curved.
-HOGG: Evidence that contemporary racing boats have keels with an active trailing edge? Also the rudder is involved.
+One amusing comment to make about the curvature of sails is that---since sails and keels have parallel roles in sailing---if it behooves sails to be curved, it probably also behooves keels to be curved.
+HOGG: Evidence that contemporary racing boats have keels with an active trailing edge? CITE! Also the rudder is involved CITE.
 
-HOGG: make sure we consistently use the term ``lift ratio'' for the air and the water, and that this is a real term.
+HOGG: make sure we use the term ``lift ratio'' for the air and the water consistently with respect to the literature, and that this is a real term.
+
+KLEBAN: Should we quantitatively report on the conditions for dwfttw? And the conditions for upwind? Right now this is all qualitative.
 

### 2021-12-29 paper/main.tex
Removed some HOGGs; added citations.

@@
 For two, these areas are \emph{effective areas} within this brutal approximation, not measured areas.
-Probably the least realistic of these effective-area ratios is $A_\keel/A_\below$; probably real boats rarely have this keel lift ratio in the hundreds.
+Probably the least realistic of these effective-area ratios is the keel lift ratio $A_\keel/A_\below$; probably real boats rarely have this lift ratio in the hundreds.
 In this sense, maybe our model is more appropriate for an ice yacht than a sailboat?
@@
 
-HOGG NOW ADDRESS THE EXCELLENT LIFT RATIOS.
-
 As we change the ratios of the effective areas, the shapes of the velocity curves change.
-In \figref{fig:design} we show the effect on the boat velocity curves of changing the ratios $A_\sail/A_\keel$, $A_\sail/A_\above$ (the sail lift ratio), and $A_\keel/A_\below$ (the keel lift ratio)....HOGG
+In \figref{fig:design} we show the effect on the boat velocity curves of changing the sail-to-keel ratio $A_\sail/A_\keel$, the sail lift ratio $A_\sail/A_\above$, and the keel lift ratio $A_\keel/A_\below$ (the keel lift ratio).
+There are a number of boat-design considerations visible in this \figurename:
+Sailing upwind is possible only if either the sail lift ratio is large, or the sail-to-keel ratio is small, or both.
+Since upwind sailing is critical for the success of a boat, and since obtaining very high lift ratios can be difficult, we predict that most commercial sailboats have sail-to-keel ratios smaller than the 700 of our fiducial boat.
+Sailing downwind faster than the wind is possible only if the keel lift ratio is large, or the sail-to-keel ratio is large.
+Sailing cross-wind is at its best when the sail-to-keel ratio is equal to the ratio of the density of water to the density of air.
+For context, recall that these ratios (lift ratios and sail-to-keel ratio) are ratios of effective areas, not real areas.
 \begin{figure}[t!]
@@
 
-There are a number of boat-design considerations visible in \figref{fig:design}.
-Sailing upwind is possible only if either the sail lift ratio is large, or the sail-to-keel ratio is small, or both.
-Since upwind sailing is critical for the success of a boat, and since obtaining very high lift ratios can be difficult, we predict that most commercial sailboats have sail-to-keel ratios smaller than the 700 of our fiducial boat.
-Sailing downwind faster than the wind is possible only if the keel lift ratio is large, or the sail-to-keel ratio is large.
-Sailing cross-wind is at its best when the sail-to-keel ratio is equal to the ratio of the density of water to the density of air.
-For context, these ratios (lift ratios and sail-to-keel ratio) are ratios of effective areas, not real areas.
-
 These design considerations relate to rigging (HOGG: CORRECT WORD) choices that sailors make for different journeys.
@@
 We don't find this, but it is certainly true that sailing downwind faster than the wind is impossible for many settings of the dimensionless ratios.
-Perhaps in real boats, with limited lift ratios, the importance of being able to sail upwind forces a small sail-t0-keel ratio and the sacrifice of failing downwind faster than the wind, which is rarely used in the normal practice of sailing.
+Perhaps in real boats, with limited lift ratios, the importance of being able to sail upwind forces a small sail-to-keel ratio and the sacrifice of sailing downwind faster than the wind, which is rarely used in the normal practice of sailing.
 
-HOGG maybe take the limit of infinite lift ratios and show that you get absurd results.
-
-HOGG: 
 A true boat has sails that are globally curved,\footnote{%
@@
 and forces that are different in detail from these pure ram-pressure forces.
-HOGG: REPHRASE: The curvature of the sails will in general rotate the air force vectors towards the direction in which the boat is pointed.
+The curvature of the sails will in general lead to differences between the sail orientation we compute for GoodSailing(tm) and the orientation of, say, the boom of any real sailboat.
+That is, in just the same sense that the areas are effective areas, the orientations are effective orientations.
 
@@
 All the code used in this \documentname{} is available open-source under the \textsl{MIT License} at \url{https://github.com/davidwhogg/Sailing}.
-The code depends on \code{numpy} (HOGG CITE), \code{scipy} (HOGG CITE), and \code{matplotlib} (HOGG CITE).
+The code depends on \code{numpy} \cite{numpy}, \code{scipy} \cite{scipy}, and \code{matplotlib} \cite{matplotlib}.
 Here we review a few of the implementation details.
@@
 where the derivative is a $3\times 3$ matrix, the inverse is a matrix inverse, and $\vec{F}(\vec{v})$ is given above in \eqref{eq:sailing}, which, in turn, depends on equations in \secref{sec:model}.
-In detail, the matrix inverse is not ever explicitly computed, but instead the code executes a \code{numpy.linalg.solve} (HOGG CITE NUMPY) operation, which is more numerically stable for this operation.
+In detail, the matrix inverse is not ever explicitly computed, but instead the code executes a \code{numpy.linalg.solve} \cite{numpy} operation, which is more numerically stable for this operation.
 Newton's method depends on an initialization.
@@
 The best value for the sail orientation $\uvec_{\perp\sail}$ from that grid search is used as the initialization for a local optimization.
-In detail we use the \code{scipy} (HOGG CITE) implementation of the Nelder--Mead algorithm (HOGG CITE) but this could probably be improved by using analytic derivatives and a gradient descent method.
+In detail we use the \code{scipy} \cite{scipy} implementation of the Nelder--Mead algorithm \cite{nelder-mead} but this could probably be improved by using analytic derivatives and a gradient descent method.
 
@@
 The pair of best values of these orientations from this grid is used as the initialization for a local optimization.
-Again we use Nelder--Mead, but probably this could be improved with a gradient descent, such as conjugate gradient (CITE).
+Again we use Nelder--Mead, but probably this could be improved with a gradient descent, such as conjugate gradient \cite{conjugate-gradient}.
 There are a few precisely degenerate orientations for the destination direction (for example, when the destination is precisely upwind); the code currently breaks these degeneracies in an uncontrolled way.
@@
 It demonstrates that there is nothing physically strange about being able to travel, by sailboat, upwind, or downwind faster than the wind, in both cases by taking a tacking course.
-Along the way, we found that the best sail-to-keel effective-area ratio for cross-wind sailing is the ratio of the density of water to the density of air; and we found that keels ought to be curved like sails (HOGG DO WE EXPLICITLY DISCUSS THIS ABOVE?
-IF NOT, WE SHOULD, WHERE WE DISCUSS THE SYMMETRY BETWEEN SAIL AND KEEL, OR WHERE WE DISCUSS THE CURVATURE OF THE SAILS.)
+Along the way, we found that the best sail-to-keel effective-area ratio for cross-wind sailing is the ratio of the density of water to the density of air; and we found that keels ought to be curved like sails.
 For both of these ``predictions'' there seems to be evidence in the designs of the best-in-class sailing vessels today.
@@
 HOGG: Speaking of misconceptions: These results probably do indeed show that you can travel directly downwind faster than the wind. That maybe resolves some folk conflicts or arguments?
-
-HOGG: This model implies, indirectly, that keels should be curved! And they are, in high-end sailing (HOGG cite?)
+In practice it might be hard to build a boat with good enough dimensionless ratios (HOGG refer back to design results).
+This might be why some sources say it is impossible (HOGG CITE).
 

### 2021-12-30 paper/main.tex
notes to self

@@
 % - confirm that "lift ratio" is the correct terminology.
-% - Switch to calling the sail and keel orientations "effective orientations" not just orientations. And say why.
-% - make sure I clearly introduce the concept of lift ratio and stick with it.
 % - Change citations to Tao to make it clear that the Tao model is (probably) wrong!!
@@
 % - Do we still mention "tensors" anywhere? make sure the abstract is consistent with the body.
+% - Respond to comments by Hyung Don Ryoo, in email to Hogg, 2021-07-15.
+% - Make sure the spirit of the bibliographic comments get pulled into the main text (and removed from the bibiliography).
+% - Fix the et al typography in the bibliography?
+% - Are we clear that we also find that you can sail faster than the wind, in general, like cross-wind? Emphasize that when we first figure it out.
 % - What claims in the literature exist about maximum boat speeds?
@@
 Have I seen a "never faster downwind than the wind" (as in Kimball) or than 0.5 times the wind claim?
-We can beat that too.
-% - Respond to comments by Hyung Don Ryoo, in email to Hogg, 2021-07-15.
-% - Make sure the spirit of the bibliographic comments get pulled into the main text.
-% - Somewhere comment that the wind is a function of height, which means that another approximation has to do with what particular mean wind speed we are considering.
-% - Are we clear that we also find that you can sail faster than the wind, in general, like cross-wind? Emphasize that when we first figure it out.
-% - Note that even a small change in the lift ratio on the sail and you can't go upwind!
-% - Move comments from the annotated bibliography up into the text somewhere and drop the annotations.
+We can perhaps beat that too.
+% - Should we be more careful to use Galilean-relativistic language, like instead of "upwind" say "upwind relative to the water"? Like always?
 
@@
     The solutions corresponding to sailing in steady wind correspond to boat velocities that deliver vanishing total force, given the angles of the sail and keel.
-    We use this simple model to derive simple properties of sailing that are not far off from standard folklore and measurements.
+    We use this simple model to derive a phenomenology of sailing that is not far off from standard folklore and measurements.
     One consequence of this model is that a sailboat does not move (relative to the water) precisely in the direction that its keel points (the direction of the prow); it moves slightly downwind of that.
@@
     Another is that cross-wind sailing is fastest when the effective area of the sail is larger than the effective area of the keel by the ratio of the density of water to the density of air.
-    Nothing conceptual about this model depends on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; the important thing is that the model \emph{does} depend on angular re-direction of fluid momentum flux; sail curvature can help with this.
+    Nothing conceptual about this model depends on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; the important thing is that the model \emph{does} depend on angular re-direction of fluid momentum flux; sail curvature can help with this, as could keel curvature.
 \end{abstract}
@@
 That might lead to a closed-form solution for the equation \eqref{eq:sailing} of steady sailing.}
-We solve equation~\eqref{eq:sailing} numerically in the code associated with this \documentname\footnote{%
-HOGG GIVE LINK HERE}
-by means of the vector generalization of Newton's method (we return to this in \secref{sec:implementation}).
+We solve equation~\eqref{eq:sailing} numerically in our code (\secref{sec:implementation})
+by means of the vector generalization of Newton's method.
 
@@
 and forces that are different in detail from these pure ram-pressure forces.
-The curvature of the sails will in general lead to differences between the sail orientation we compute for GoodSailing(tm) and the orientation of, say, the boom of any real sailboat.
-That is, in just the same sense that the areas are effective areas, the orientations are effective orientations.
+The curvature of the sails is probably necessary to deliver high lift ratios; redirection of momentum with low drag usually requires carefully tuned surfaces (as seen in wings and propellers as well).
+The curvature of the sails will in general lead to differences between the sail orientation we compute in our simple model for GoodSailing(tm) and the orientation of, say, the boom of any real sailboat.
+That is, in just the same sense that the areas we use are effective areas, the orientations we use are effective orientations.
 

### 2021-12-30 paper/main.tex
define tacking and coming about

@@
 That makes sense; the whole point (in some sense) of building a model of a sailboat is to understand how it sails upwind.
-In \secref{sec:design} we will return to the upwind travel; we will show that in principle you might be able to build a boat that can travel upwind \emph{faster than the wind}.
+In \secref{sec:design} we will return to the conditions under which upwind travel is possible.
 Indeed, this boat (specs in \secref{sec:design}) does not sail upwind very fast, and it can't sail at an angle that is even close to directly upwind.
-This closest angle to directly upwind will also be a function of the design of the boat.
+In order to travel directly upwind, the boat must ``tack'': It must sail at an angle to the wind with the wind coming over the starboard side, then ``come about'' and sail at an angle to the wind with the wind coming over the port side, and repeat.
+This tacking (with repeated comings about) is how sailboats are sailed to upwind destinations.
 
@@
 
-Finally, although upwind travel is slow going, with lots of tacking (HOGG DO WE NEED TO DEFINE THIS? YES WE DO) required, if you are situated \emph{on the boat}, upwind travel \emph{feels very fast}.
+Finally, although upwind travel is slow going, with lots of tacking required, if you are situated \emph{on the boat}, upwind travel \emph{feels very fast}.
 Why?
 Because your experience of speed is, at least in part, the feeling of the wind in your face.
-If you think about \figref{fig:hourglass} in terms of the boat-air velocity instead of the boat-water velocity---that is, if you Galilean-boost to the rest frame of the air---the upwind travel has much larger boat-air velocities than the downwind travel.
-So even though you can sail fast downwind, it sometimes \emph{feels} faster upwind.
+If you think about \figref{fig:hourglass} in terms of the boat--air velocity instead of the boat--water velocity---that is, if you Galilean-boost to the rest frame of the air---the upwind travel has much larger boat--air velocities than the downwind travel.
+So even though you can sail fast (relative to the water) downwind, it sometimes \emph{feels} faster upwind, because your strongest perception is of the relative boat--air velocity, not the relative boat--water velocity.
 
@@
   \caption{The $x$ and $y$ coordinates of the relative boat--water velocity $\vboat-\vwater$ for GoodSailing(tm) for a set keel orientations.
-  For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:good}.
+  For all of these trials, the relative air--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:good}.
   Compare this to \figref{fig:hourglass}.

### 2021-12-30 paper/main.tex
trying to get rid of HOGGs

@@
 % - Should we be more careful to use Galilean-relativistic language, like instead of "upwind" say "upwind relative to the water"? Like always?
+% - Mention sailing hydrofoil where we discuss the keel lift ratio: The point of the hyrdrofoil is to reduce A_below.
 
@@
 For example, a sailor must in general take non-trivial tacking paths from origin to destination (especially when the destination is upwind, but even when the destination is downwind).
-The corners of that path---``coming about'' in sailing parlance---cost time, so their frequency should be balanced against slightly non-optimal sailing directions.
-Thus an even better sailor will put the boat into configurations that don't appear in \figref{fig:hourbetter}.
+The corners of that path---each of the comings about---cost time, so their frequency should be balanced against slightly non-optimal sailing directions.
+That is, a BestSailing(tm) sailor will put the boat into configurations that don't appear in \figref{fig:hourbetter}.
 The full exposition of this requires a model for the costs of coming about.
@@
 
-These design considerations relate to rigging (HOGG: CORRECT WORD) choices that sailors make for different journeys.
+These design considerations relate to sail choices that sailors make for different journeys.
 For example, in downwind travel, often much larger or additional sails are employed (HOGG CITE).
@@
 And we defined BetterSailing(tm), which is an optimal setting of both the orientation of the boat and the orientation of the sail, given the direction to the destination.
-But even this is not the best you can do, because optimal sailing (from a time-to-destination perspective) involves complete path planning, constrained to start at the origin and end at the destination.
+But (as we mention in \secref{sec:better}) even this is not the best you can do, because optimal sailing (from a time-to-destination perspective) involves complete path planning, constrained to start at the origin and end at the destination.
 For destinations that are close to directly downwind or directly upwind of the origin, this complete path planning will involve tacking the boat.
-Each tack involves a ``coming about'' or a rotation of the boat to put the wind on the other beam (HOGG CORRECT LANGUAGE HERE?).
+Each tack involves a coming about---a rotation of the boat---to put the wind on the other side of the boat and sail.
 Each coming about costs time.
@@
 
-HOGG: All code used to make the figures in this \documentname{} are available HOGG WHERE?
-
 \paragraph{Acknowledgements:}
 It is a pleasure to thank Will Kinney (Buffalo), Hans-Walter Rix (MPIA), Hyung-Don Ryoo (NYU), and Soledad Villar (JHU) for valuable discussions and comments.
+All code used to make the figures in this \documentname{} are available under the open-source \textsl{MIT License} at \url{https://github.com/davidwhogg/Sailing}.
 

### 2021-12-31 paper/main.tex
Getting rid of HOGGs, other adjustments

@@
 % - Add to the abstract the point that the sailboat is defined by 3 dimensionless numbers: The sail-to-keel ratio, and two lift ratios.
-% - confirm that "lift ratio" is the correct terminology.
 % - Change citations to Tao to make it clear that the Tao model is (probably) wrong!!
@@
 \addtolength{\oddsidemargin}{-0.5\textwidth}
-\setlength{\textheight}{9.00in}
-\setlength{\topmargin}{-0.40in}
+\setlength{\textheight}{9.40in}
+\setlength{\topmargin}{-0.50in}
+\setlength{\footskip}{0.25in}
 \pagestyle{myheadings}
@@
 They control the ratio of useful sailing forces to drag.
-We will refer to these---na\"ively---as lift ratios in what follows.
+We will refer to these---na\"ively---as \emph{lift ratios} in what follows, even though that is slightly at odds with the usual use of the term ``lift ratio''.\footnote{The usual definition of the lift ratio is ratio of the lift force to the drag force, at a given speed and pitch (orientation).
+Here we are defining an orientation-free ratio of effective areas.
+At every setting of the sail orientation, the ratio of effective areas $A_\sail/A_\above$ is proportional to the true sail lift ratio.
+For this reason, we refer to the ratio of effective areas as the lift ratio in what follows.}
 
@@
 If we want to operationalize steady sailing in code, steady sailing can be thought of as a root-finding problem, in which we find the vector $\vboat$ that makes \eqref{eq:sailing} true, given the orientations $\uvec_{\perp\sail}, \uvec_{\perp\keel}$ of the sail and keel.
-This is a root-finding problem that (probably) has no closed form.\footnote{%
+This is a root-finding problem that (probably---KLEBAN??) has no closed form.\footnote{%
 It might be interesting to note that the flat-surface force \eqref{eq:flat} is a vector-valued quadratic form, so if one constructed the sailboat model completely from flat surfaces, the force law, in this approximation would be a quadratic form.
@@
   For all of these trials, the relative wind--water velocity is $\vair-\vwater=(8\,\mps)\,\uvec_x$ (that is, $8\,\mps$ in the $x$ direction), as it is in every panel of \figref{fig:steady}.
-  Points to the left of the leftmost vertical grey line are sail--keel settings for which the boat sails upwind.
-Points to the right of the rightmost vertical grey line are sail--keel settings for which the boat sails downwind faster than the wind.
+  Points to the left of the leftmost vertical grey line are sail--keel settings for which the boat sails upwind (relative to the water).
+Points to the right of the rightmost vertical grey line are sail--keel settings for which the boat sails downwind (relative to the water) faster than the wind.
   There are no settings that go \emph{directly} upwind, and no settings that go \emph{directly} downwind faster than the wind.\label{fig:hourglass}}
@@
 This boat also has very good drag ratios or lift ratios:
-The sail lift ratio (the ratio of effective area of sail to hull-above) is 250, as is the keel lift ratio (the ratio of effective area of keel to hull-below).
 All these ratios were set not by analyzing any real boats, but by making the speed curve shown in \figref{fig:hourgood} look reasonable, or accord with expectations set by the books on the physics of sailing HOGG CITE THESE.
@@
 These design considerations relate to sail choices that sailors make for different journeys.
-For example, in downwind travel, often much larger or additional sails are employed (HOGG CITE).
+For example, in downwind travel, often much larger or additional sails are employed \cite{sails} (HOGG CONFIRM APPROPRIATENESS OF THIS CITATION).
 
@@
 HOGG: Evidence that contemporary racing boats have keels with an active trailing edge? CITE! Also the rudder is involved CITE.
-
-HOGG: make sure we use the term ``lift ratio'' for the air and the water consistently with respect to the literature, and that this is a real term.
 
@@
 Newton's method requires the $3\times 3$ derivative tensor $\dd\vec{F}/\dd\vec{v}$.
-The code explicitly delivers this derivative, but it would be possible to use \code{jax} (HOGG CITE JAX) or any other autodifferentiation package to get the tensor automatically.
+The code explicitly delivers this derivative, but it would be possible to use any autodifferentiation framework to get the tensor in closed form automatically.
 
@@
 
-HOGG: This is a momentum-transport model! NOW Take some time to bash the Bernoulli bullshit. Note also that the different-path-length part of that argument is explicitly not true for a sail!
+This is a momentum-transport model.
+The flight of birds and planes is also explained by momentum transport.
+There is a common misconception about airfoils that they generate lift primarily because of Bernoulli effect and differing path lengths across the top and bottom.
+This misconception is now debunked \cite{lift}.
+A version of this debunked argument has also occasionally been applied to the forces on sails \cite{explained}.
+However, it is not applicable, not just because it has been generally debunked, but also because the windward and leeward sides of the sail have the same path lengths, so the differing-path-length part of the argument can't be applied \cite{sails} (HOGG CHECK THAT THIS IS THE CORRECT CITE).
+Flying and sailing are very related; they both involve redirecting the momenta of fluids, as does swimming.
 
-HOGG: Speaking of misconceptions: These results probably do indeed show that you can travel directly downwind faster than the wind. That maybe resolves some folk conflicts or arguments?
-In practice it might be hard to build a boat with good enough dimensionless ratios (HOGG refer back to design results).
-This might be why some sources say it is impossible (HOGG CITE).
+Speaking of misconceptions: These results probably do indeed show that you can travel directly downwind faster than the wind.
+Of course our sailboat can't (it has to tack), but a pair of sailboats working in tandem could tack oppositely downwind and pull a cargo directly downwind faster than the wind.
+That finding, perhaps, resolves some folk conflicts or arguments that it is physically impossible?
+In practice it might be hard to build a boat with good enough dimensionless ratios (see \secref{design}).
+Maybe it would be easier to build ice boats with the required ratios.
+This might be why some sources say it is impossible (KLEBAN CITE).
+
+Speaking of ice boats: These could have great properties!
+These are boats that, instead of having a keel in the water, have thin blades that slide on lake ice.
+With the change from a keel in water to blades on ice, the equivalent of the keel lift ratio could be almost arbitrarily large for an ice boat, and the equivalent of the sail-to-keel ratio could approach zero, because ice is a solid and does not flow (on the relevant time scales).
+Our design investigations (\secref{sec:design}) suggest that ice boats should be capable of extremely fast upwind travel.
 
 We expressed pride that our model is explicitly Galilean-relativistic.
-That begs the question: What about special relativity?
-Of course our model is consistent with special relativity; Galilean relativity is the low-velocity limit, when all relative velocities have magnitudes way less than the speed of light.
+That begs the question: What about \emph{special relativity} \cite{sr}?
+Of course our model is consistent with special relativity; Galilean relativity is the low-velocity limit, when all relative velocities have magnitudes much less than the speed of light.
 Now what about \emph{relativistic sailing?}
@@
 But the Galaxy is filled with stars!
-With multiple sources of illumination, in principle one might be able to make a photon-pressure vehicle that can sail or tack in arbitrary directions, given illumination from multiple stars.
-This might involve a relativistic generalization of the above.
-HOGG: Someone must have thought of all this before! CITE THEM.
+With multiple sources of illumination, in principle one might be able to make a photon-pressure vehicle that can sail or tack in arbitrary directions, given illumination (and maybe also gravitational forces) from multiple stars.
+This might involve a relativistic generalization of the above, because the momenta would be photon momenta, and Doppler shifts would replace velocity differences.
 

### 2023-06-17 paper/main.tex
small changes and tweaks

@@
 % - Mention sailing hydrofoil where we discuss the keel lift ratio: The point of the hyrdrofoil is to reduce A_below.
+% - Answer Gruzinov's question: How does one sail downwind fttw? What does it look like.
+% - Explain why amateur sailboats have huge keels and professional boats have small keels. Upwind sailing is the answer!
+% - Kleban points about non-trivial wind force f(theta) and references (in email, maybe).
+% - Add little flags on masts of figures to show wind wrt boat.
 
@@
 \usepackage[letterpaper]{geometry} % because Overleaf is weird.
-\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\usepackage[hidelinks]{hyperref}
 \usepackage{graphicx}
@@
 Indeed---according to lore---Galilean relativity was discovered by consideration of doing physics experiments in the hold of a sailing ship \cite{galileo}.
-The contemporary view of Galilean relativity is that all velocities that appear in physical law or the prediction of observables must be relative velocities.
+The contemporary view of Galilean relativity is that all velocities that appear in physical law or the prediction of observables must be relative velocities, or in other words, velocity differences.
 In addition to obeying this symmetry, our model is consistent with dimensional analysis and is equivariant to rotations and reflections.
@@
 When we optimize the trim of our boat, we learn about what things the boat can do, in terms of sailing speed, and sailing upwind and downwind.
-It's impressive!
+Spoiler: It is impressive!
 We will discuss the dependence of our results on our boat properties in \secref{sec:design}.
@@
 Ram-pressure forces are generally associated with bow shocks and turbulent wakes.
-Many things (cars, launch vehicles) are designed to \emph{avoid} the ram-pressure effect known as air resistance or drag.
+Many things (cars, propellers, launch vehicles) are designed to \emph{avoid} the ram-pressure effect known as air resistance or drag.
 
@@
 %has an expression that is, essentially,
-can be expressed as a particle flux (number per time) times a change in momentum per particle.
+can be expressed as a particle flux (number per area per time) times a change in momentum per particle.
 These are also the physical components of a ram-pressure force.
@@
 The directions of the forces can be manipulated to create amazing boat dynamics and kinematics.
-The point of this \documentname{} is that much of the amazing dynamics can be understood in an amazingly simple model for the sailboat.
+The point of this \documentname{} is that much of the amazing dynamics can be understood in a dead simple model for the sailboat.
 So while we use the phrases ``flat sail'' and ``ram-pressure'' in what follows to describe our model, we really mean ``simple momentum transport''.
@@
 for some angle $\theta_\sail$ in that coordinate system.
-We won't particularly use angles much in what follows, but if you want to come back to angles, you can take the arctangent of the components of $\uvec_{\perp\sail}$.
+We won't particularly use angles much in what follows, because we want to remain coordinate free, but if you want to come back to angles, you can take arctangents of ratios of components of unit vectors like $\uvec_{\perp\sail}$.
 
@@
 \end{align}
-where $\eta$ is a dimensionless prefactor, $\rho$ is the density of the fluid, $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow), $\vec{v}$ is the %velocity of the fluid in the rest frame of the surface,
-relative velocity between the fluid and the surface,
+where $\eta$ is a dimensionless prefactor,
+  $\rho$ is the density of the fluid,
+  $\uvec_\perp$ is the unit vector perpendicular to the surface (oriented such that it points down-flow),
+  $\vec{v}$ is the relative velocity of the fluid with respect to the surface,
   and $\uvec_\perp\cdot\vec{v}$ is the scalar inner product (dot product) of the two vectors.
@@
 The way to see this is by computing the Reynolds number, which, loosely speaking, is the ratio of inertial forces to viscous forces.
-It is a typical lengthscale (like the height of the sail) times a typical velocity (like the relative velocity of the boat and air) divided by the kinematic viscosity (something like $10^{-5}\,\mmps$ for air and $10^{-6}\,\mmps$ for water).
+It is a typical length scale (like the width of the sail) times a typical velocity (like the relative velocity of the boat and air) divided by the kinematic viscosity (something like $10^{-5}\,\mmps$ for air and $10^{-6}\,\mmps$ for water).
 Typical Reynolds numbers for both the boat--air contact and the boat--water contact are therefore on the order of $10^6$ to $10^7$ for real boats in typical conditions.
 That is, viscosity plays no first-order role in sailing.
-In detail it shapes the turbulence in the wakes coming off the sail and keel edges, but that is a tiny detail from perspective of our brutal model.}
+In detail it shapes the turbulence in the wakes coming off the sail and keel edges, but that is a tiny detail from the perspective of our brutal model.}
 We're going to treat $\eta$ as an uninteresting constant; in particular we are going to assume that it doesn't depend on velocity, such that it can be combined with the area $A$ (to make an effective area) in everything that follows.
@@
 \paragraph{Acknowledgements:}
-It is a pleasure to thank Will Kinney (Buffalo), Hans-Walter Rix (MPIA), Hyung-Don Ryoo (NYU), and Soledad Villar (JHU) for valuable discussions and comments.
+It is a pleasure to thank
+  Will Kinney (Buffalo),
+  Hans-Walter Rix (MPIA),
+  Hyung-Don Ryoo (NYU), and
+  Soledad Villar (JHU)
+for valuable discussions and comments.
 All code used to make the figures in this \documentname{} are available under the open-source \textsl{MIT License} at \url{https://github.com/davidwhogg/Sailing}.

### 2024-08-02 paper/main.tex
Updates from Overleaf

@@
 % - David Grier
+% - Lehman Garrison, HWR, Dani Maoz?
 % - All those in the acks.
@@
 % ------------
+% - Update figure captions to match new figures!
 % - Add to the abstract the point that the sailboat is defined by 3 dimensionless numbers: The sail-to-keel ratio, and two lift ratios.
@@
 % - Kleban points about non-trivial wind force f(theta) and references (in email, maybe).
-% - Add little flags on masts of figures to show wind wrt boat.
+% - Note that better sailing (tm) is almost never in the direction of the target!
 
@@
   The views are top-down; the keel is represented by a thick grey line, the sail is represented by a thin black line, and the wind and boat velocities are represented by black arrows.
-  Most of the settings of the keel and sail lead to downwind travel, but a few in the top-left quarter of this grid lead to upwind travel.\label{fig:steady}}
+  Most of the settings of the keel and sail lead to downwind travel, but a few in the top-left quarter of this grid lead to upwind travel.
+LEHMAN: HIGHLIGHT THE FRAME WITH THE OPTIMAL ANGLE OF ATTACK IN EACH COLUMN?\label{fig:steady}}
   \figurerule
@@
 It is a pleasure to thank
+  Lehman Garrison (Flatiron),
   Will Kinney (Buffalo),

### 2024-08-02 paper/main.tex

@@
   \includegraphics[width=\textwidth]{steady.pdf}
-  \caption{Visualizations of the boat velocity $\vboat$ for different settings of the orientations of the keel and sail, all presented in the water rest frame.
+  \caption{HOGG: THIS CAPTION IS ALL WRONG NOW: ALL PLOTTED VELOCITIES ARE DIFFERENCES.
+Visualizations of the boat velocity $\vboat$ for different settings of the orientations of the keel and sail, all presented in the water rest frame.
   The views are top-down; the keel is represented by a thick grey line, the sail is represented by a thin black line, and the wind and boat velocities are represented by black arrows.

### 2024-08-03 paper/main.tex
Notes to self, based on conversations

@@
 % - David Grier
-% - Lehman Garrison, HWR, Dani Maoz?
+% - Dani Maoz? Joe Shields?
 % - All those in the acks.
@@
 % - Update figure captions to match new figures!
+% - Add to the abstract and the introduction and discussion that this is only a 2-D model of sailing. Not 3-D!
 % - Add to the abstract the point that the sailboat is defined by 3 dimensionless numbers: The sail-to-keel ratio, and two lift ratios.
@@
 % - Mention sailing hydrofoil where we discuss the keel lift ratio: The point of the hyrdrofoil is to reduce A_below.
-% - Answer Gruzinov's question: How does one sail downwind fttw? What does it look like.
 % - Explain why amateur sailboats have huge keels and professional boats have small keels. Upwind sailing is the answer!
@@
 % - Note that better sailing (tm) is almost never in the direction of the target!
+% - Is there an analog of the Betts or Betz limit? Also just generally discussion of power, power dissipated, and so on.
+% - Do we adequately mention sustainability in the introduction?
+% - How to plot the phase space of dimensionless ratios that allow for UW and DWFTTW?
+% - Can we address stability and uniqueness of the solutions to the (quartic, actually) equations?
+% - Do we clearly define the dimensionless ratios that define a boat? There are three.
+% - Kleban wants polars -- plots that are used in the business. Can do!
 
@@
 \usepackage[utf8]{inputenc}
-\usepackage[letterpaper]{geometry} % because Overleaf is weird.
-\PassOptionsToPackage{hyphens}{url}
+\usepackage[letterpaper]{geometry}
+\PassOptionsToPackage{hyphens}{url} % huh?
 \usepackage[hidelinks]{hyperref}
@@
 \title{\bfseries%
-A simple momentum-transport model for sailing}
-\markright{\textsf{Hogg \& Kleban / Momentum-transport model for sailing}}
+How do sailboats work? A simple momentum-transport model}
+\markright{\textsf{Hogg \& Kleban / A momentum-transport model for sailing}}
 \author{DWH, MK, others?}
-\date{2021}
+\date{2021--2024}
 \begin{document}
@@
     Similarly, there is a ram-pressure force on the keel, depending on the relative water--boat velocity and the keel orientation.
-    The resulting expression for the total force on the boat is coordinate-free and consistent with Galilean relativity.
-    The solutions corresponding to sailing in steady wind correspond to boat velocities that deliver vanishing total force, given the angles of the sail and keel.
+    The resulting expression for the total force on the boat is coordinate-free and covariant in the sense of Galilean relativity.
+    The solutions corresponding to sailing in steady wind correspond to boat velocities that deliver vanishing total force, given orientations of the sail and keel.
     We use this simple model to derive a phenomenology of sailing that is not far off from standard folklore and measurements.
@@
 Maybe that helps?
-HOGG SAYS: I THINK WE NEED A DIAGRAM HERE, SHOWING DWFTTW and UW.
+We will say more about upwind sailing in \secref{sec:upwind}.
 
@@
 Because your experience of speed is, at least in part, the feeling of the wind in your face.
-If you think about \figref{fig:hourglass} in terms of the boat--air velocity instead of the boat--water velocity---that is, if you Galilean-boost to the rest frame of the air---the upwind travel has much larger boat--air velocities than the downwind travel.
+If you think about the boat--air velocity instead of the boat--water velocity---that is, if you Galilean-boost to the rest frame of the air---the upwind travel has much larger boat--air velocities than the downwind travel.
 So even though you can sail fast (relative to the water) downwind, it sometimes \emph{feels} faster upwind, because your strongest perception is of the relative boat--air velocity, not the relative boat--water velocity.
@@
 
+\section{Sailing downwind faster than the wind and upwind}\label{sec:upwind}
+
+HOGG WRITE HERE.
+
 \section{Boat properties and boat design}\label{sec:design}
@@
 
-KLEBAN: Should we quantitatively report on the conditions for dwfttw? And the conditions for upwind? Right now this is all qualitative.
+KLEBAN SAY: Should we quantitatively report on the conditions for dwfttw? And the conditions for upwind? Right now this is all qualitative.
+HOGG SAY: YES; THERE are settings for the dimensionless ratios that permit and don't permit it. Not sure how to present this because there are three dimensionless ratios.
 
@@
 for valuable discussions and comments.
-All code used to make the figures in this \documentname{} are available under the open-source \textsl{MIT License} at \url{https://github.com/davidwhogg/Sailing}.
+All code used to solve equations and make figures in this \documentname{} is available under the open-source \textsl{MIT License} at \url{https://github.com/davidwhogg/Sailing}.
 

### 2024-08-07 paper/main.tex
Updates from Overleaf

@@
 % - Kleban wants polars -- plots that are used in the business. Can do!
+% - Discuss "velocity made good" in the text and maybe name-check it in the abstract?
 
@@
 
-We (the authors) are not sailors!
+We (the authors) are not serious sailors.
 However, we believe that sailing is a beautiful technology.
@@
 
+HOGG: Argument about the wedge being squeezed.
+
+HOGG: Diagrams.
+
 \section{Boat properties and boat design}\label{sec:design}
@@
 KLEBAN SAY: Should we quantitatively report on the conditions for dwfttw? And the conditions for upwind? Right now this is all qualitative.
-HOGG SAY: YES; THERE are settings for the dimensionless ratios that permit and don't permit it. Not sure how to present this because there are three dimensionless ratios.
+HOGG SAY: YES; THERE are settings for the dimensionless ratios that permit and don't permit it. Not sure how to present this because there are three dimensionless ratios. Working on it.
 
@@
 That finding, perhaps, resolves some folk conflicts or arguments that it is physically impossible?
-In practice it might be hard to build a boat with good enough dimensionless ratios (see \secref{design}).
+In practice it might be hard to build a boat with good enough dimensionless ratios (see \secref{sec:design}).
 Maybe it would be easier to build ice boats with the required ratios.

### 2024-08-21 paper/main.tex
tiny things
a few notes
caption tweak

@@
 % - Dani Maoz? Joe Shields?
+% - Saskia Hekker
 % - All those in the acks.
@@
 
-We (the authors) are not serious sailors.
+We (the authors) are not serious sailors!
 However, we believe that sailing is a beautiful technology.
@@
 
+HOGG: How are uw and dwfttw the same question in different reference frames?
+
 HOGG: Argument about the wedge being squeezed.
@@
 HOGG: Diagrams.
+
+HOGG: Note about how it feels. Is this repetitive with comments above? If so, remove them and replace with references down to this Section.
 
@@
   \rule{0.25\textwidth}{0pt}\includegraphics[width=0.5\textwidth]{design_E.pdf}
-  \caption{Each panel is analagous to \figref{fig:hourgood} but in each case we have varied one of the boat's dimensionless ratios.
+  \caption{HOGG SOME PARTS OF THIS CAPTION ARE WRONG Each panel is analagous to \figref{fig:hourgood} but in each case we have varied one of the boat's dimensionless ratios.
 In each panel, the grey points show the same speed curve as that plotted in \figref{fig:hourgood}, which was computed for the boat described by the parameters given in \eqref{eq:boat1} through \eqref{eq:boat4}, and for a wind speed (relative to the water) of $8\,\mps$.
@@
   In the bottom panel, the small and large black dots show the variations as we vary the sail-to-keel ratio $A_\sail/A_\keel$ from 140 (small dots) to 700 (grey dots) to 3500 (large dots).\label{fig:design}}
+  \figurerule
+\end{figure}
+
+KLEBAN SAY: Should we quantitatively report on the conditions for dwfttw? And the conditions for upwind? Right now this is all qualitative.
+HOGG SAY: YES; THERE are settings for the dimensionless ratios that permit and don't permit it.
+These are illustrated in \figref{fig:dwfttw}, as a function of lift ratios, for boats with two different values of the sail-to-keel area ratio.
+The brief summary is .... HOGG
+\begin{figure}[t!]
+  \includegraphics[width=0.5\textwidth]{upwind_vmg.pdf}%
+  \includegraphics[width=0.5\textwidth]{downwind_vmg.pdf}\\%
+  \rule{0.25\textwidth}{0pt}\includegraphics[width=0.5\textwidth]{crosswind_vmg.pdf}
+  \caption{HOGG WRITE CAPTION.
+  In the top-left panel, HOGG.
+  In the top-right panel, HOGG.
+  In the bottom panel, HOGG.\label{fig:dwfttw}}
   \figurerule
@@
 HOGG: Evidence that contemporary racing boats have keels with an active trailing edge? CITE! Also the rudder is involved CITE.
-
-KLEBAN SAY: Should we quantitatively report on the conditions for dwfttw? And the conditions for upwind? Right now this is all qualitative.
-HOGG SAY: YES; THERE are settings for the dimensionless ratios that permit and don't permit it. Not sure how to present this because there are three dimensionless ratios. Working on it.
 

### 2024-08-25 paper/main.tex
Updates from Overleaf

--- 2024-08-07 paper/main.tex
@@
 Our model for a sailboat is going to be that it is a composite object, composed of a flat sail, a flat keel, and an isotropic partially submerged hull.
+LEHMAN: ISN'T THE KEEL TECHNICALLY PART OF THE HULL?
+IF SO, THE HULL WOULDN'T REALLY BE ISOTROPIC.
+MAYBE JUST NEED TO DEFINE HULL VS KEEL HERE.
 Thus the ram-pressure force on the sailboat will have four components, a force on the flat sail from the air, a force on the flat keel from the water, a force on the hull from the air, and a force on the hull from the water.
@@
 
+LEHMAN: AN OBSERVATION ABOUT FIG.~2: THE DWFTTW PARAMETER SPACE LOOKS SYMMETRICAL WITH UPWIND SAILING!
+(IS THIS STRICTLY TRUE?) THIS MAKES SOME SENSE: ONCE YOU'RE GOING DWFTTW, YOU HAVE A HEADWIND THAT IS THE SAME AS IF YOU WERE SAILING UPWIND.
+BUT THE VELOCITY RELATIVE TO THE WATER IS DIFFERENT IN THE TWO CASES, SO WHAT RESTORES THE SYMMETRY?
+MAYBE FIG.~7 PROVES THAT THE SYMMETRY IS ACCIDENTAL.
+
+LEHMAN: IS ONE WAY TO THINK ABOUT DWFTTW THAT SOME OF THE "USELESS" FORCE OF THE WIND ON THE SAIL/HULL (I.E.~PERPENDICULAR TO DIRECTION OF TRAVEL) GETS REDIRECTED TO "USEFUL" FORCE BY THE ANGLE OF THE KEEL AGAINST THE WATER?
+IF SO, THIS WOULD SEEM TO REQUIRE THE TOTAL WIND-SAIL FORCE AT A "FAST" ANGLE TO BE GREATER THAN THE WIND-SAIL FORCE DIRECTLY DOWNWIND.
+NOT SURE WHY THIS WOULD BE TRUE, BUT MAYBE MY CONFUSION COULD BE HELPFUL IN DECIDING HOW TO WRITE THIS SECTION...
+
 \section{Boat properties and boat design}\label{sec:design}

### 2024-08-25 paper/main.tex

--- 2024-08-21 paper/main.tex
@@
 
-HOGG: Note about how it feels. Is this repetitive with comments above? If so, remove them and replace with references down to this Section.
+LEHMAN: AN OBSERVATION ABOUT FIG.~2: THE DWFTTW PARAMETER SPACE LOOKS SYMMETRICAL WITH UPWIND SAILING!
+(IS THIS STRICTLY TRUE?) THIS MAKES SOME SENSE: ONCE YOU'RE GOING DWFTTW, YOU HAVE A HEADWIND THAT IS THE SAME AS IF YOU WERE SAILING UPWIND.
+BUT THE VELOCITY RELATIVE TO THE WATER IS DIFFERENT IN THE TWO CASES, SO WHAT RESTORES THE SYMMETRY?
+MAYBE FIG.~7 PROVES THAT THE SYMMETRY IS ACCIDENTAL.
+
+LEHMAN: IS ONE WAY TO THINK ABOUT DWFTTW THAT SOME OF THE "USELESS" FORCE OF THE WIND ON THE SAIL/HULL (I.E.~PERPENDICULAR TO DIRECTION OF TRAVEL) GETS REDIRECTED TO "USEFUL" FORCE BY THE ANGLE OF THE KEEL AGAINST THE WATER?
+IF SO, THIS WOULD SEEM TO REQUIRE THE TOTAL WIND-SAIL FORCE AT A "FAST" ANGLE TO BE GREATER THAN THE WIND-SAIL FORCE DIRECTLY DOWNWIND.
+NOT SURE WHY THIS WOULD BE TRUE, BUT MAYBE MY CONFUSION COULD BE HELPFUL IN DECIDING HOW TO WRITE THIS SECTION...
+
+HOGG: Possible Note about how it feels. Is this repetitive with comments above? If so, remove them and replace with references down to this Section.
 

### 2025-01-18 paper/main.tex
testing git merge
qualitative section start. intro content. notes to selves.

@@
-% This paper is copyright 2021 the authors.
+% This paper is copyright 2021, 2025 the authors.
 
@@
 % - Saskia Hekker
+% - Charlie Peskin
 % - All those in the acks.
@@
 % - Maybe the "flat" force should be the "planar" force?
+% - Do we capitalize north and east and southeast>
 
@@
 % - Discuss "velocity made good" in the text and maybe name-check it in the abstract?
+% -
 
@@
 \title{\bfseries%
-How do sailboats work? A simple momentum-transport model}
+The spherical scow: a simple model for sailboats}
 \markright{\textsf{Hogg \& Kleban / A momentum-transport model for sailing}}
@@
 \begin{abstract}\noindent
-    We present a simplified model for the forces on a sailboat.
-    Our model is that there is a ram-pressure force on the sail, with a magnitude and direction that depend on the relative air--boat velocity and the sail orientation.
-    Similarly, there is a ram-pressure force on the keel, depending on the relative water--boat velocity and the keel orientation.
+    We present a simple model that captures the essential physics governing a sailboat.
+In this model, the sail experiences a ram-pressure force determined by the boat’s velocity relative to the air and the sail’s orientation.
+Similarly, there is a ram-pressure force on the keel depending on the boat’s velocity relative to the water and the keel’s orientation.
+Finally, the hull experiences a force determined by the relative velocities of air and water with respect to the boat.
     The resulting expression for the total force on the boat is coordinate-free and covariant in the sense of Galilean relativity.
-    The solutions corresponding to sailing in steady wind correspond to boat velocities that deliver vanishing total force, given orientations of the sail and keel.
-    We use this simple model to derive a phenomenology of sailing that is not far off from standard folklore and measurements.
+
+    Solutions corresponding to sailing at constant velocity in steady wind correspond to boat velocities that,  for given sail and keel orientations, yield zero net force.
+    We use this simple model to derive a phenomenology of sailing that is not far off from standard folklore and real-world measurements.
+
     One consequence of this model is that a sailboat does not move (relative to the water) precisely in the direction that its keel points (the direction of the prow); it moves slightly downwind of that.
-    Another consequence is that it is possible for an extremely well designed boat (very good lift ratios) to sail downwind (relative to the water) \emph{faster than the wind}, on a tacking course.
-    Another is that cross-wind sailing is fastest when the effective area of the sail is larger than the effective area of the keel by the ratio of the density of water to the density of air.
-    Nothing conceptual about this model depends on the \emph{curvature} of the sail, in contradiction to some folk explanations of sailing; the important thing is that the model \emph{does} depend on angular re-direction of fluid momentum flux; sail curvature can help with this, as could keel curvature.
+    Another consequence is that it is possible for a well-designed boat (with very good lift ratios) to make progress downwind (relative to the water) \emph{faster than the wind}, on a tacking course.
+    Another is that cross-wind sailing is fastest when the effective area of the sail is larger than the effective area of the keel by a factor equal to the ratio of the density of water to the density of air.
+    The model does not include \emph{curvature} of the sail.
+In contradiction to some folk explanations of sailing, our model boat is nevertheless able to sail upwind, and make progress downwind faster than the wind.
+The model \emph{does} depend on angular re-direction of fluid momentum flux (sail curvature would help with this, as would keel curvature).
 \end{abstract}
@@
 The boat is powered by the wind, and yet it can travel (possibly with a tacking course) in any direction.
-Of course it's not consistent with Galilean relativity\footnote{We refer to Galilean relativity repeatedly in this \documentname.
+Of course it is not consistent with Galilean relativity\footnote{We refer to Galilean relativity repeatedly in this \documentname.
 Of course Galilean relativity is just the low-velocity approximation to special relativity \cite{sr}.
-But we are not going to consider air--water velocity differences that are an appreciable fraction of the speed of light (except maybe in the discussion in \secref{sec:discussion}).} to say that the boat is powered by the wind:
-There is no absolute velocity, and there is a valid rest frame in which the wind velocity is zero.
-The Galilean-relativistic thing to say is that the boat is powered by the relative velocity between the air and the water.
+But we are not going to consider air--water velocity differences that are an appreciable fraction of the speed of light (except maybe in the discussion in \secref{sec:discussion}).} to say that the boat is powered by the motion of air:
+There is no absolute velocity, and there is a valid  frame in which the air velocity is zero.
+The Galilean-relativistic description is that the boat is powered by the relative velocity between the air and the water (which is a good definition of ``wind").
 Indeed, the symmetry between the air and the water in the mechanisms of sailing is the fundamental theme of one of the better books on the physics of sailing \cite{symmetry}.
@@
 
-KLEBAN: Write things about ddwfttw, maybe? For context only! But it is worth saying that if we can understand sailing with a simple model, we might resolve some disputes about ddwfttw.
+One of the interesting conceptual questions related to sailing is whether it is possible to progress downwind, in a wind-powered vehicle, faster than the wind.
+To be clear, imagine a balloon with the same density as the air that drifts downwind at exactly the speed of the air.
+Can a sailboat (or other wind-powered vehicle) beat the balloon to a mark dead downwind of the starting point?
+The immediate reaction of many people is that this must be impossible.
+After all, how can a wind-powered vehicle outrun the wind that powers it?
+
+We will see that this is in fact possible according to our simple model (with sufficiently optimal parameters for the sailboat), and  is relatively routine in real (but very high-performance!) boats CITE?.
+Some years ago a pair of engineers even constructed a cart driven by a propeller CITE.
+The propeller is geared to the cart's wheels, which provide the torque that drives the propeller.
+Remarkably, this cart is capable of running straight downwind at several times the speed of the wind.
+Of course, the rotation of the propeller means that not all parts of the vehicle are, in fact, moving dead downwind; indeed, the blades of the propeller act much like the sails of a sailboat tacking on a downwind reach CITE?.
+The \emph{force} driving the cart comes from molecules of air that bounce off these moving, angled surfaces, while the \emph{energy} comes from the fact that the relative motion of air and ground is slowed by the passing cart.
+
+HOGG: DO WE NEED A PARAGRAPH ON SUSTAINABLE ENERGY, new transportation technologies, wind power and so on? That's also motivation.
 
@@
 There is no reason for the air flow to ``join up'' at the end of the airfoil.
-And, besides, airplanes can fly upside-down!
+Besides, airplanes can fly upside-down!
 The best leading-order description is that the lift on the wing comes from \emph{redirection of the flow of air}.
@@
 This point is made admirably in some books on the physics of sailing (especially \cite{sails}).
-And the force on a sailboat keel is from the redirection of the flow of water encountering the keel.
+The force on a sailboat keel is from the redirection of the flow of water encountering the keel.
 These forces are sometimes called ``lift'' forces in analogy to the lift force on an airplane wing.
@@
 
+\section{Qualitative discussion}\label{sec:qualitative}
+
+HOGG: AUDIT FOR REPETITIVENESS BETWEEN THIS AND WHAT'S ABOVE AND BELOW.
+
+Before we introduce our quantitative, but simple, physical model for a sailboat,
+let's look at how sailing works, qualitatively.
+A sailboat works on the \emph{relative velocity} of the air and the water,
+with the sail intersecting the air and the keel intersecting the water.
+That is, sailing in an arbitrary direction depends on there being a difference between the air velocity and the water velocity.
+The relevant air and water velocities are those relative to the boat, because the sail and keel move with the boat.
+
+With good engineering, or when simplified as we simplify below, the pressure forces from the air on the sail and from the water on the keel
+have directions that are close to normal to the sail and keel surfaces.
+(Of course sails are curved but we will consider the mean sail surface normal direction for the purpose of this discussion.)
+Thus if we set the direction of the keel and the sail such that they form a kind of wedge,
+the normal forces from the air and water will squeeze this wedge and the boat will be pushed forward.
+It will be pushed forward in much the same way that a slippery watermelon seed will shoot outwards when squeezed between your fingertips.
+
+The easiest sailing direction to understand is cross-wind.
+Here the wind is coming from (say) the north, and the boat is pointed (roughly) east.
+In the frame of the boat, the air is coming from the northeast, and the water is coming from the southeast.
+The water is coming from the southeast in the frame of the boat because the boat is being pushed downwind by the air.
+The air--sail normal force and water--keel normal force push the boat forward.
+Figure~HOGG shows the geometry of the situation in the water frame, the boat frame, and idealized to the squeezed wedge.
+HOGG: CROSSWIND DIAGRAM IN WATER FRAME, BOAT FRAME, and abstracted down to a wedge.
+
+In some ways the most important sailing direction is upwind (in the frame of the water).
+The ability for boats to sail upwind is critical for making round-trip journeys on our planet with its relatively steady atmospheric currents.
+The boat cannot sail directly upwind but it can make upwind progress by tacking at angles to the wind.
+If the wind is coming from the north, on one one of these tacks the boat might be pointed towards the northeast.
+In the boat frame, the wind is coming from the north-northeast, and the water is coming from the east-northeast because the boat is being pushed downwind by the air.
+Figure~HOGG shows the geometry of this situation.
+
+The strangest (and rarest) sailing direction is downwind faster than the wind (in the frame of the water).
+Again, this must be done by tacking.
+HOGG... KEEP WRITING.
+
+HOGG: HOW FAST can we sail?
+As the boat goes faster, the air and water velocities (relative to the boat) start to converge.
+The wedge gets very narrow.
+This will only work if the forces are very precisely normal.
+That is only true if the lift ratios are very high---drag forces are very low.
+But there is no limit \emph{in principle} to how fast a boat can sail.
+And, indeed sailboats routinely sail substantially faster than the windspeed (relative to the water) CITE.
+
 \section{Why ram-pressure forces?}\label{sec:ram}
@@
 So while we use the phrases ``flat sail'' and ``ram-pressure'' in what follows to describe our model, we really mean ``simple momentum transport''.
+
+HOGG TECHNICAL POINT that ram pressure is in play when Reynolds numbers are huge. They are huge.
 
@@
 Our model for a sailboat is going to be that it is a composite object, composed of a flat sail, a flat keel, and an isotropic partially submerged hull.
-LEHMAN: ISN'T THE KEEL TECHNICALLY PART OF THE HULL?
-IF SO, THE HULL WOULDN'T REALLY BE ISOTROPIC.
-MAYBE JUST NEED TO DEFINE HULL VS KEEL HERE.
+LEHMAN: ISN'T THE KEEL TECHNICALLY PART OF THE HULL? IF SO, THE HULL WOULDN'T REALLY BE ISOTROPIC. MAYBE JUST NEED TO DEFINE HULL VS KEEL HERE.
 Thus the ram-pressure force on the sailboat will have four components, a force on the flat sail from the air, a force on the flat keel from the water, a force on the hull from the air, and a force on the hull from the water.
@@
   Most of the settings of the keel and sail lead to downwind travel, but a few in the top-left quarter of this grid lead to upwind travel.
-LEHMAN: HIGHLIGHT THE FRAME WITH THE OPTIMAL ANGLE OF ATTACK IN EACH COLUMN?\label{fig:steady}}
+  LEHMAN: HIGHLIGHT THE FRAME WITH THE OPTIMAL ANGLE OF ATTACK IN EACH COLUMN?
+  \label{fig:steady}}
   \figurerule

### 2025-01-25 paper/main.tex
various things

@@
 % ------------
+% - Incorporate references and terminology from Leif Ristroph (NYU). See email of 2025-01-24 and 25.
 % - Update figure captions to match new figures!
@@
 \title{\bfseries%
-The spherical scow: a simple model for sailboats}
+The spherical scow:\\
+A simple model for sailing}
 \markright{\textsf{Hogg \& Kleban / A momentum-transport model for sailing}}
@@
   Will Kinney (Buffalo),
+  Leif Ristroph (NYU),
   Hans-Walter Rix (MPIA),
